\documentclass{article}

% NeurIPS 2026 — preprint mode
\usepackage[preprint,nonatbib]{neurips_2026}
\usepackage[numbers,compress,sort]{natbib}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{array}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{rotating}
\usepackage{colortbl}
\usepackage{longtable}

\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,fit,backgrounds,calc}

\hypersetup{colorlinks=true,linkcolor=blue!70!black,citecolor=blue!55!black,urlcolor=blue!55!black}

% Table colour helpers
\definecolor{clyes}{RGB}{220,245,220}
\definecolor{clpart}{RGB}{255,248,210}
\definecolor{clno}{RGB}{252,228,228}
\newcommand{\yes}{\cellcolor{clyes}\checkmark}
\newcommand{\no}{\cellcolor{clno}--}
\newcommand{\pt}{\cellcolor{clpart}$\sim$}

% Convenience macros
\newcommand{\nerf}{NeRF}
\newcommand{\tgs}{3DGS}
\newcommand{\nvs}{NVS}
\newcommand{\etal}{\emph{et~al.}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\cf}{\emph{cf.}}

% ─────────────────────────────────────────────────────────────────────────────
\title{%
  From Novel View Synthesis to Generative World Modeling:\\
  A Comprehensive Survey%
}

\author{%
  ElephantFlow Research \\
  \texttt{\href{https://github.com/elephantflow/ElephantSurvey}{%
    github.com/elephantflow/ElephantSurvey}} \\[2pt]
  \textit{Draft v0.3 — March 2026}
}

% ═════════════════════════════════════════════════════════════════════════════
\begin{document}
\maketitle

% ─────────────────────────────────────────────────────────────────────────────
\begin{abstract}
Novel View Synthesis~(\nvs)---rendering a scene from arbitrary unseen viewpoints
given reference observations---has undergone a paradigm shift from classical
geometry-based reconstruction toward learning-driven generative approaches.
This survey examines \nvs{} not merely as a rendering technique but as the
\textbf{foundational spatial capability} of Generative World Models:
systems that predict visual scene states under arbitrary conditions including
viewpoint changes, temporal dynamics, and agent interactions.

We present a unified conceptual framework organising existing methods along a
\emph{continuous spectrum}---from static multi-view reconstruction to dynamic,
interactive world generation---across three axes:
(i)~scene representation (implicit, explicit, hybrid),
(ii)~generative paradigm (per-scene optimisation, feed-forward inference,
diffusion-based synthesis),
and (iii)~world-modelling capability (geometry, dynamics, controllability).
We survey more than \textbf{180 representative methods} spanning NeRF,
3D Gaussian Splatting, generalizable reconstruction, diffusion-based NVS,
and 4-D world modeling, and propose a roadmap toward physically grounded,
generalisable world models.
\end{abstract}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

\subsection{The Fundamental Problem}

Intelligence requires reasoning about the world beyond what is immediately
visible.
A self-driving vehicle must anticipate the scene around a blind corner;
a robotic arm must predict how a grasped object will appear from the wrist
camera after a planned motion; a surgeon must mentally reconstruct
anatomy from a 2-D scan.
In each case, the core operation is identical: \textbf{synthesising a
plausible visual state at an unobserved vantage point}.

This is the problem of \textbf{Novel View Synthesis~(\nvs)}.
Formally, given a set of reference observations
$\{\mathbf{I}_i, \boldsymbol{\pi}_i\}_{i=1}^{n}$ with known camera poses
$\boldsymbol{\pi}_i$, the goal is to render a photo-realistic image
$\hat{\mathbf{I}}$ at an arbitrary target pose $\boldsymbol{\pi}^{*}$,
faithfully preserving 3-D geometry, surface appearance, and lighting.
The problem is inherently ill-posed when $n$ is small or when the target
pose is far from any reference: an infinite family of 3-D scenes
is consistent with any finite set of 2-D projections.
Resolving this ambiguity requires either dense capture (classical IBR) or
a learned prior over the distribution of plausible scenes (modern neural
and generative approaches).

While \nvs{} was studied in the image-based rendering
era~\citep{levoy1996light,mcmillan1995plenoptic,chen1993view,debevec1996modeling},
the past six years have witnessed an extraordinary paradigm shift:
NeRF~\citep{mildenhall2020nerf} demonstrated that a shallow MLP can represent
complex scenes with view-dependent appearance by differentiably volume-rendering
a continuous radiance field;
3DGS~\citep{kerbl2023gaussian} showed that an explicit set of anisotropic
Gaussians can render at real-time frame rates;
and diffusion-based methods~\citep{liu2023zero1to3,poole2023dreamfusion,yu2024viewcrafter}
have moved the frontier toward \emph{generalisable, generative} synthesis
that requires as few as a single image---or even a text description---as input.

\subsection{From NVS to World Modeling}

We argue that this progression reflects a deeper conceptual shift:
\textbf{\nvs{} is becoming the foundational spatial inference engine of
Generative World Modeling}.

\begin{quote}
  \textbf{Working Definition~(World Model).}\quad
  A \emph{Generative World Model} is a learned system that can predict the
  visual (and optionally physical) state of a scene under arbitrary
  conditions---including viewpoint changes, temporal evolution, and agent
  interactions---in a manner that is \emph{geometrically consistent},
  \emph{physically plausible}, and \emph{generalisable} to scenes and conditions
  outside the training distribution.
\end{quote}

Under this definition, \nvs{} is the special case in which the only varying
condition is camera pose.
A full world model must additionally handle temporal dynamics~(D3),
action conditioning~(D4), and physical plausibility~(D5).
Crucially, \emph{spatial consistency enforced by \nvs{} is a necessary
prerequisite}: a world model without geometric grounding is merely a video
predictor---perceptually plausible in distribution but unreliable for tasks
requiring precise spatial reasoning such as robot manipulation or navigation.

Recent systems are actively closing this gap.
Genie~2~\citep{parkerholder2024genie2} generates interactive 3-D environments
from a single image; Cosmos~\citep{agarwal2025cosmos} trains world foundation
models at internet scale; and ViewCrafter~\citep{yu2024viewcrafter} with
StreetCrafter~\citep{yan2024streetcrafter} demonstrate that camera-controlled
video generation is practical at high resolution.
These developments reflect the field's convergence on a unified model that
subsumes NVS, video generation, and world simulation.

\subsection{Relationship to Prior Surveys}

Table~\ref{tab:surveys} compares this work with the most closely related
prior surveys.
Tewari~\etal~\citep{tewari2022advances} and
Gao~\etal~\citep{gao2022nerf_survey} focus on \nerf{} representations
without addressing world modelling.
Wu~\etal~\citep{wu2024recentGS} and
Fei~\etal~\citep{fei2024gssurvey} cover 3DGS but not generative methods or 4-D.
Chen~\etal~\citep{chen2024survey_drivingNVS} cover \nvs{} for autonomous driving
but lack a world-modeling lens.
To our knowledge, \textbf{no existing survey jointly covers all three paradigms
(\nerf{}, \tgs{}, generative) through the unifying lens of Generative World
Modeling}, nor proposes a continuous-spectrum taxonomy from static
reconstruction to 4-D interactive generation.

\begin{table}[t]
  \centering
  \caption{Comparison with related surveys.
    \textbf{\checkmark}~covered;\; $\sim$~partially;\; --~not covered.}
  \label{tab:surveys}
  \scriptsize
  \setlength{\tabcolsep}{3pt}
  \begin{tabular}{@{}lcccccccc@{}}
    \toprule
    \textbf{Survey} & \textbf{Year}
      & \textbf{NeRF} & \textbf{3DGS} & \textbf{Generative}
      & \textbf{Dynamic} & \textbf{Generalise}
      & \textbf{4D/WM} & \textbf{Unified} \\
    \midrule
    Tewari~\etal~\citep{tewari2022advances}        & 2022 & \yes & \no  & \pt  & \pt & \pt & \no & \no \\
    Gao~\etal~\citep{gao2022nerf_survey}            & 2022 & \yes & \no  & \no  & \pt & \no & \no & \no \\
    Zhu~\etal~\citep{zhu2023survey_ibr}             & 2023 & \pt  & \no  & \no  & \no & \no & \no & \no \\
    Wu~\etal~\citep{wu2024recentGS}                 & 2024 & \no  & \yes & \no  & \pt & \no & \no & \no \\
    Fei~\etal~\citep{fei2024gssurvey}               & 2024 & \no  & \yes & \pt  & \pt & \no & \no & \no \\
    Chen~\etal~\citep{chen2024survey_drivingNVS}    & 2024 & \pt  & \pt  & \no  & \yes& \no & \pt & \no \\
    Zhang~\etal~\citep{zhang2025feedforward_survey} & 2025 & \pt  & \pt  & \no  & \no & \yes& \no & \no \\
    He~\etal~\citep{he2025gs_apps}                  & 2025 & \no  & \yes & \pt  & \pt & \no & \no & \no \\
    \textbf{Ours}                                   & 2026 & \yes & \yes & \yes & \yes& \yes& \yes& \yes\\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{The Unified Spectrum}

We organise all methods along a \emph{continuous spectrum} defined by three axes
(Figure~\ref{fig:spectrum}):

\begin{enumerate}[leftmargin=*,label=\textbf{Axis~\arabic*.}]
  \item \textbf{Input regime}: dense multi-view $\to$ sparse-view $\to$
        single image $\to$ text/unconditional.
  \item \textbf{Temporal scope}: static scene $\to$ deformable/dynamic $\to$
        4-D world generation.
  \item \textbf{Controllability}: viewpoint only $\to$ viewpoint~+~time $\to$
        viewpoint~+~time~+~action.
\end{enumerate}

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
      method/.style={font=\tiny,align=center,inner sep=2pt},
    ]
    \draw[-{Stealth},thick] (0,0)--(9.2,0)
      node[right,font=\small]{Input: dense $\to$ sparse $\to$ text};
    \draw[-{Stealth},thick] (0,0)--(0,5.4)
      node[above,font=\small]{Temporal: static $\to$ dynamic $\to$ 4D};

    \fill[blue!6,rounded corners=3pt]   (0.1,0.1) rectangle (3.0,2.0);
    \fill[purple!6,rounded corners=3pt] (0.1,2.1) rectangle (4.5,4.0);
    \fill[orange!7,rounded corners=3pt] (3.1,0.1) rectangle (8.9,4.0);
    \fill[red!8,rounded corners=3pt]    (4.6,3.1) rectangle (8.9,5.2);

    \node[font=\tiny\bfseries,text=blue!60]   at (1.5,1.8) {Classical/NeRF};
    \node[font=\tiny\bfseries,text=purple!70] at (2.3,3.8) {Dynamic NeRF/4DGS};
    \node[font=\tiny\bfseries,text=orange!70] at (5.8,0.5) {Generalizable/Generative NVS};
    \node[font=\tiny\bfseries,text=red!70]    at (7.0,5.0) {4D World Models};

    \node[method] at (0.9,0.9)  {NeRF};
    \node[method] at (2.0,1.4)  {3DGS};
    \node[method] at (1.5,0.5)  {Instant-NGP};
    \node[method] at (2.8,1.0)  {Mip-NeRF360};
    \node[method] at (0.7,3.0)  {D-NeRF};
    \node[method] at (2.2,3.4)  {4DGS};
    \node[method] at (1.5,2.5)  {Nerfies};
    \node[method] at (3.5,0.9)  {pixelNeRF\\IBRNet};
    \node[method] at (4.8,1.5)  {pixelSplat\\MVSplat};
    \node[method] at (5.8,2.2)  {DUSt3R\\MASt3R};
    \node[method] at (6.5,1.0)  {Zero-1-to-3\\SyncDreamer};
    \node[method] at (7.8,1.6)  {DreamFusion\\GaussianDreamer};
    \node[method] at (8.2,0.6)  {LRM\\TriplaneGaussian};
    \node[method] at (5.2,3.7)  {ViewCrafter};
    \node[method] at (6.8,4.5)  {Genie2\\Cosmos};
    \node[method] at (8.0,3.8)  {GAIA-1\\UniSim};

    \draw[-{Stealth[length=9pt]},dashed,thick,red!60]
      (1.5,1.5)--(7.5,4.4)
      node[midway,above,sloped,font=\scriptsize\bfseries,red!70]
        {$\to$ Generative World Modeling};
  \end{tikzpicture}
  \caption{The unified NVS spectrum. The diagonal arrow marks the evolutionary
    trajectory toward 4-D Generative World Modeling.}
  \label{fig:spectrum}
\end{figure}

\subsection{Contributions and Paper Structure}

\textbf{Contributions:}
(1)~A unified continuous-spectrum taxonomy across three axes (input regime,
temporal scope, controllability);
(2)~Cross-paradigm coverage of $>$180 methods with explicit World Modeling
implications in each chapter;
(3)~A related-survey gap analysis (Table~\ref{tab:surveys}) substantiating
the novelty of the World Modeling lens;
(4)~A five-direction open-problems roadmap.

The remainder is organised as follows.
\S\ref{sec:background} establishes background and a unified formulation.
\S\ref{sec:nerf}--\ref{sec:generalisable} survey the three representation
paradigms (NeRF, 3DGS, feed-forward reconstruction).
\S\ref{sec:generative} covers generative \nvs{} with particular depth on
camera-controlled diffusion.
\S\ref{sec:world} examines 4-D generation, neural driving simulators, and
world foundation models.
\S\ref{sec:datasets}--\ref{sec:eval} review datasets and evaluation protocols.
\S\ref{sec:open} proposes a research roadmap.

% ─────────────────────────────────────────────────────────────────────────────
\section{Background and Unified Formulation}
\label{sec:background}

\subsection{Classical Image-Based Rendering}

Before the neural era, \nvs{} was approached through
\emph{image-based rendering}~(IBR)~\citep{mcmillan1995plenoptic,chen1993view,levoy1996light}.
The foundational insight is that images directly encode light transport:
sufficiently dense sampling of the plenoptic function~\citep{adelson1991plenoptic}
permits view interpolation without reasoning about explicit 3-D geometry.

Light Field Rendering~\citep{levoy1996light} and the Lumigraph~\citep{gortler1996lumigraph}
parameterise the light field on a two-plane slab, enabling photo-realistic view
interpolation within the captured light-field volume.
View interpolation~\citep{chen1993view} warps and blends reference images
using depth estimates, anticipating modern depth-based view synthesis.
Unstructured Lumigraphs and view-dependent texture mapping~\citep{debevec1996modeling}
relaxed capture requirements to unstructured camera sets.
Structure from Motion~(SfM)~\citep{schoenberger2016sfm} and
Multi-View Stereo~(MVS)~\citep{schonberger2016mvs} established the
geometry-estimation pipeline that underlies all pose-conditioned \nvs{} to this day.

The fundamental limitation of classical IBR is the requirement for
\emph{dense} capture: interpolation quality degrades rapidly as target
viewpoints move beyond the convex hull of the observed camera set.
Furthermore, these methods produce no compact scene model---each novel view
requires querying the full set of reference images at inference time.
The desire for compact, generalisable scene representations drove the
development of neural radiance fields.

\subsection{Volume Rendering and NeRF}

NeRF~\citep{mildenhall2020nerf} represents a scene as a continuous function
$F_\theta:(\mathbf{x},\mathbf{d})\mapsto(\mathbf{c},\sigma)$, where
$\mathbf{x}\in\mathbb{R}^3$ is a 3-D location and
$\mathbf{d}\in\mathbb{S}^2$ a viewing direction.
The expected colour of a ray $\mathbf{r}(s)=\mathbf{o}+s\mathbf{d}$ is
computed via classical volume rendering~\citep{kajiya1984ray}:
\begin{equation}
  \hat{C}(\mathbf{r})=\int_{s_n}^{s_f} T(s)\,\sigma(\mathbf{r}(s))\,
  \mathbf{c}(\mathbf{r}(s),\mathbf{d})\,\mathrm{d}s,
  \quad T(s)=\exp\!\left(-\int_{s_n}^s\sigma(\mathbf{r}(u))\,\mathrm{d}u\right),
  \label{eq:volrender}
\end{equation}
where $T(s)$ is the accumulated transmittance.
$F_\theta$ is a position-encoded MLP~\citep{tancik2020fourier} trained by
minimising $\|\hat{C}(\mathbf{r})-C(\mathbf{r})\|_2^2$ over all observed rays.
Hierarchical stratified sampling decomposes the continuous integral into
coarse and fine networks, concentrating samples near the visible surface.

NeRF's key properties---implicit representation, continuous differentiability,
and view-dependent appearance via directional encoding---made it transformative.
Its central limitations are \emph{slow per-scene optimisation} (hours) and
\emph{per-scene specificity} (no generalisation to unseen scenes), both of which
the subsequent literature directly addressed.

\subsection{Gaussian Splatting}

3DGS~\citep{kerbl2023gaussian} represents a scene as $N$ anisotropic 3-D
Gaussians $\{(\boldsymbol{\mu}_i, \boldsymbol{\Sigma}_i, \alpha_i,
\mathbf{c}_i)\}_{i=1}^N$, where $\boldsymbol{\mu}_i$ is the centre,
$\boldsymbol{\Sigma}_i$ the full covariance, $\alpha_i$ the opacity, and
$\mathbf{c}_i$ a set of spherical-harmonic coefficients encoding
view-dependent colour.
Novel views are rendered by projecting each Gaussian to a 2-D splat and
compositing front-to-back:
\begin{equation}
  \hat{C}=\sum_{i\in\mathcal{N}}\mathbf{c}_i\,\alpha_i\,\tilde{G}_i
  \prod_{j<i}(1-\alpha_j\,\tilde{G}_j),
  \label{eq:splatting}
\end{equation}
where $\tilde{G}_i$ denotes the 2-D projection of Gaussian~$i$.
Training proceeds by differentiable rasterisation and minimisation of a
combined $\ell_1$ and SSIM photometric loss.
An \emph{adaptive densification} strategy clones under-reconstructed Gaussians
and splits over-large ones, while pruning transparent or redundant primitives.

3DGS achieves real-time rendering ($>$100\,fps at 1080p) with training times
of tens of minutes---a dramatic improvement over NeRF.
Its explicit, editable representation makes it attractive for downstream
tasks in scene editing, simulation, and embodied AI.
The primary limitations are \emph{higher memory footprint} (millions of Gaussians
per scene) and \emph{limited surface quality} compared to SDF-based methods.
A large subsequent literature has addressed both limitations.

\subsection{Unified NVS and World Modeling Formulation}

We formulate \nvs{} as a general function:
\begin{equation}
  \mathcal{M}_\theta:\bigl(\{\mathbf{I}_i,\boldsymbol{\pi}_i\}_{i=1}^n,\;
  \boldsymbol{\pi}^*\bigr)\longrightarrow\hat{\mathbf{I}}^*,
  \quad n\in[1,+\infty).
  \label{eq:nvs}
\end{equation}
The world-modeling extension augments this with a time index $t$ and a
sequence of agent actions $\mathbf{a}_{0:t}$:
\begin{equation}
  \mathcal{M}_\theta:\bigl(\{\mathbf{I}_i,\boldsymbol{\pi}_i\}_{i=1}^n,\;
  \mathbf{a}_{0:t},\,t,\,\boldsymbol{\pi}^*\bigr)\longrightarrow\hat{\mathbf{I}}^*_t.
  \label{eq:wm}
\end{equation}
Eq.~\eqref{eq:nvs} is the special case $t=0$, $\mathbf{a}=\varnothing$.
A complete world model satisfying Eq.~\eqref{eq:wm} must fulfil five desiderata,
which we use throughout the survey:
\textbf{D1}~geometric consistency across viewpoints;
\textbf{D2}~generalisation from sparse or single-image input;
\textbf{D3}~dynamic scene modelling;
\textbf{D4}~action conditioning;
\textbf{D5}~physical plausibility.

% ─────────────────────────────────────────────────────────────────────────────
\section{Scene Representation Taxonomy}
\label{sec:representation}

The choice of scene representation determines the fundamental trade-offs a
method faces: rendering speed, reconstruction accuracy, memory footprint,
generalisability, and editability.
We identify four families, each reflecting a different position on the
speed--compactness--quality Pareto frontier.

\textbf{Implicit representations} (MLPs, hash grids) encode the scene as a
continuous function $f_\theta(\mathbf{x})$.
Their key advantages are compactness and continuous differentiability, which
support high-quality reconstruction from dense observations.
Their central limitation is rendering speed: each pixel requires integrating
along a ray through many network evaluations.
NeRF~\citep{mildenhall2020nerf} and its successors (§\ref{sec:nerf}) are the
canonical family.

\textbf{Explicit representations} (voxels, point clouds, Gaussians) store scene
attributes at discrete spatial elements, enabling tile-based or splat-based
rasterisation that achieves real-time frame rates.
The trade-off is memory: a faithful 3DGS scene may occupy hundreds of megabytes.
3D Gaussian Splatting~\citep{kerbl2023gaussian} (§\ref{sec:3dgs}) is the
dominant current representative.

\textbf{Hybrid representations} combine an explicit spatial structure---typically
a triplane~\citep{chan2022efficient} or multi-resolution
grid~\citep{muller2022instant}---with a shallow MLP decoder.
They strike a practical balance between the quality of implicit models and
the speed of explicit ones, and are the backbone of most large reconstruction
models (§\ref{sec:generalisable}).

\textbf{Compositional representations} decompose the scene into semantic
entities (objects, backgrounds, agents) with independent sub-representations.
This structural inductive bias enables object-level editing and compositional
generation~\citep{niemeyer2021giraffe,zhou2024gala3d}, but typically requires
supervision or segmentation to separate entities.
It is also the natural representation for world models that must reason about
independently acting objects.

% ─────────────────────────────────────────────────────────────────────────────
\section{NeRF: Implicit Neural Radiance Fields}
\label{sec:nerf}

\subsection{Foundational NeRF Methods}

The original NeRF~\citep{mildenhall2020nerf} demonstrated that an MLP with
positional encoding could faithfully represent complex real scenes with
view-dependent appearance from a few dozen posed photographs.
Subsequent work systematically addressed its two principal limitations---slow
optimisation and poor generalisation to novel scenes.

\paragraph{Accelerating per-scene optimisation.}
The bottleneck in vanilla NeRF is the MLP query: every rendered pixel
requires hundreds of network evaluations along the ray.
Instant-NGP~\citep{muller2022instant} replaces the large MLP backbone with a
compact multi-resolution \emph{hash grid}, enabling training convergence in
under a minute on a single GPU while matching or exceeding prior quality.
TensoRF~\citep{chen2022tensorf} decomposes the radiance field into low-rank
CP and VM tensor components, reducing both storage and rendering overhead.
Plenoxels~\citep{fridovich2022plenoxels} eliminates the MLP entirely,
using a sparse 3-D voxel grid with trilinearly interpolated spherical harmonics;
the resulting model trains in minutes without gradient-intensive network
backward passes.
DirectVoxGO~\citep{sun2022dvgo} accelerates convergence via a two-stage
coarse-to-fine voxel grid that aggressively skips empty space.

\paragraph{Improving visual quality and handling challenging scenes.}
Mip-NeRF~\citep{barron2021mipnerf} replaces the point-sampled ray with an
\emph{integrated positional encoding} over a conical frustum, substantially
reducing aliasing at multiple viewing distances.
Mip-NeRF~360~\citep{barron2022mipnerf360} extends this to unbounded outdoor
scenes via a non-linear scene parameterisation that contracts the infinite
background into a bounded representation.
Zip-NeRF~\citep{barron2023zipnerf} unifies the hash grid of Instant-NGP with
the anti-aliasing of Mip-NeRF~360, establishing the current quality frontier
for per-scene NeRF.
NeRF-W~\citep{martinbrualla2021nerfw} introduces per-image latent appearance
codes and a transient object model, enabling reconstruction of tourist-photographed
landmarks from uncontrolled internet images.
Ref-NeRF~\citep{verbin2022refnerf} restructures the directional encoding into
a physically-based reflection model that explicitly predicts diffuse colour,
specular tint, and surface roughness, dramatically improving the rendering of
specular and metallic materials.

\paragraph{Surface reconstruction.}
While volume rendering naturally produces accurate alpha composites, it does not
enforce a well-defined surface, leading to noisy geometry.
NeuS~\citep{wang2021neus} reformulates volume rendering using a logistic
density derived from a Signed Distance Function~(SDF), yielding watertight,
high-quality surface reconstruction.
VolSDF~\citep{yariv2021volsdf} derives a rigorous analytical mapping from
SDF values to volume density, providing a theoretically grounded alternative.
BakedSDF~\citep{yariv2023bakedsdf} combines the surface quality of SDF-based
NeRF with real-time rendering by extracting and baking the learned SDF into a
high-quality mesh.

Table~\ref{tab:nerf_methods} provides a comprehensive comparison of NeRF variants
across all three sub-families: per-scene optimisation, generalizable,
and dynamic methods.

\begin{table*}[t]
  \centering
  \caption{Comprehensive NeRF method comparison.
    \textbf{Category}: PS=per-scene, GEN=generalizable, DYN=dynamic.
    \textbf{Train}/\textbf{Render}: speed relative to vanilla NeRF (Fast/Med/Slow).
    \textbf{Gen.}: generalizes across scenes without per-scene re-training.
    \textbf{Dyn.}: handles dynamic or deformable scenes.
    \textbf{Surf.}: produces explicit surface (mesh/SDF).}
  \label{tab:nerf_methods}
  \scriptsize
  \setlength{\tabcolsep}{3pt}
  \begin{tabular}{@{}llccccccp{4.2cm}@{}}
    \toprule
    \textbf{Method} & \textbf{Venue} & \textbf{Cat.}
      & \textbf{Train} & \textbf{Render}
      & \textbf{Gen.} & \textbf{Dyn.} & \textbf{Surf.}
      & \textbf{Key Contribution} \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Per-scene optimisation}} \\
    NeRF~\citep{mildenhall2020nerf}            & ECCV'20    & PS & Slow & Slow & \no & \no & \no & MLP + positional encoding \\
    NeRF-W~\citep{martinbrualla2021nerfw}       & CVPR'21    & PS & Slow & Slow & \no & \no & \no & In-the-wild appearance codes \\
    Mip-NeRF~\citep{barron2021mipnerf}         & ICCV'21    & PS & Slow & Slow & \no & \no & \no & Integrated positional encoding \\
    NeuS~\citep{wang2021neus}                  & NeurIPS'21 & PS & Slow & Slow & \no & \no & \yes & SDF-based volume rendering \\
    VolSDF~\citep{yariv2021volsdf}             & NeurIPS'21 & PS & Slow & Slow & \no & \no & \yes & Analytical SDF-to-density \\
    Plenoxels~\citep{fridovich2022plenoxels}   & CVPR'22    & PS & Fast & Fast & \no & \no & \no & No-MLP voxel grid + SH \\
    TensoRF~\citep{chen2022tensorf}            & ECCV'22    & PS & Med  & Fast & \no & \no & \no & CP/VM tensor decomposition \\
    DirectVoxGO~\citep{sun2022dvgo}            & CVPR'22    & PS & Fast & Fast & \no & \no & \no & Two-stage coarse-to-fine voxel \\
    Instant-NGP~\citep{muller2022instant}      & TOG'22     & PS & Fast & Fast & \no & \no & \no & Multi-resolution hash grid \\
    Mip-NeRF360~\citep{barron2022mipnerf360}   & CVPR'22    & PS & Slow & Slow & \no & \no & \no & Unbounded scene contraction \\
    Ref-NeRF~\citep{verbin2022refnerf}         & CVPR'22    & PS & Slow & Slow & \no & \no & \no & Physics-based reflection model \\
    Zip-NeRF~\citep{barron2023zipnerf}         & ICCV'23    & PS & Fast & Fast & \no & \no & \no & Hash grid + anti-aliasing \\
    BakedSDF~\citep{yariv2023bakedsdf}         & SIGGRAPH'23& PS & Med  & Fast & \no & \no & \yes & SDF mesh extraction + baking \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Generalizable methods}} \\
    pixelNeRF~\citep{yu2021pixelnerf}          & CVPR'21    & GEN& Fast & Slow & \yes& \no & \no & Pixel-aligned CNN features \\
    IBRNet~\citep{wang2021ibrnet}              & CVPR'21    & GEN& Fast & Slow & \yes& \no & \no & Cross-view transformer aggregation \\
    MVSNeRF~\citep{chen2021mvsnerf}            & ICCV'21    & GEN& Fast & Slow & \yes& \no & \no & Cost-volume geometry prior \\
    SRT~\citep{sajjadi2022srt}                 & CVPR'22    & GEN& Fast & Fast & \yes& \no & \no & Set-latent scene representation \\
    GNT~\citep{wang2022gnt}                    & arXiv'22   & GEN& Fast & Med  & \yes& \no & \no & Fully attention-based, no NeRF MLP \\
    NerfDiff~\citep{gu2023nerfdiff}            & ICML'23    & GEN& Fast & Slow & \yes& \no & \no & Diffusion prior for single-view \\
    ZeroRF~\citep{shi2023zerorf}               & arXiv'23   & GEN& Med  & Slow & \pt & \no & \no & Zero-pretrain sparse-view 360$^\circ$ \\
    FrameNeRF~\citep{yan2024framenerf}         & arXiv'24   & GEN& Fast & Slow & \yes& \no & \no & Video frames as pseudo views \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Dynamic \& deformable methods}} \\
    D-NeRF~\citep{pumarola2021dnerf}           & CVPR'21    & DYN& Slow & Slow & \no & \yes& \no & Canonical deformation network \\
    NSFF~\citep{li2021neural_scene_flow}       & CVPR'21    & DYN& Slow & Slow & \no & \yes& \no & Explicit scene flow field \\
    Nerfies~\citep{park2021nerfies}            & ICCV'21    & DYN& Slow & Slow & \no & \yes& \no & Per-frame latent deformation \\
    HyperNeRF~\citep{park2021hypernerf}        & TOG'21     & DYN& Slow & Slow & \no & \yes& \no & Higher-dimensional ambient space \\
    Non-Rigid NeRF~\citep{tretschk2021nonrigid}& ICCV'21    & DYN& Slow & Slow & \no & \yes& \no & Monocular ray bending \\
    DyNeRF~\citep{li2022neural3d}              & CVPR'22    & DYN& Slow & Slow & \no & \yes& \no & Time-conditioned multi-view NeRF \\
    TiNeuVox~\citep{fang2022tineuvox}          & SIGGRAPH'22& DYN& Med  & Med  & \no & \yes& \no & Voxel-accelerated dynamic NeRF \\
    K-Planes~\citep{fridovich2023kplanes}      & CVPR'23    & DYN& Fast & Fast & \no & \yes& \no & 4-D plane decomposition \\
    HexPlane~\citep{cao2023hexplane}           & CVPR'23    & DYN& Fast & Fast & \no & \yes& \no & Concurrent 4-D plane factorisation \\
    NeRFPlayer~\citep{song2023nerfplayer}      & TOG'23     & DYN& Med  & Med  & \no & \yes& \no & Static/deforming/new region decomp. \\
    Soup-of-Planes~\citep{du2023soupofplanes}  & ICCV'23    & DYN& Fast & Fast & \no & \yes& \no & Planar fitting for casual video \\
    Pseudo-GDV~\citep{zhao2024pgdvs}           & CVPR'24    & DYN& Med  & Slow & \no & \yes& \no & Static + dynamic inpainting \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Generalizable NeRF}

Per-scene NeRF requires lengthy per-scene optimisation, limiting scalability.
\emph{Generalisable} methods learn a shared prior across scenes, enabling
fast inference on new scenes.

pixelNeRF~\citep{yu2021pixelnerf} conditions the NeRF MLP on pixel-aligned
CNN features extracted from reference views, enabling single-image reconstruction.
IBRNet~\citep{wang2021ibrnet} uses a transformer to aggregate features from
multiple reference views at query points along the ray.
MVSNeRF~\citep{chen2021mvsnerf} builds a cost volume from stereo for
geometry-aware feature extraction.
GNT~\citep{wang2022gnt} dispenses with the NeRF MLP entirely, replacing it
with a two-stage transformer that first encodes reference views into a
view-independent representation and then renders the target view via
cross-attention, learning geometric reasoning as an emergent property of
attention rather than an explicit inductive bias.
SRT~\citep{sajjadi2022srt} pursues a similar philosophy but uses a
set-latent architecture, encoding an arbitrary number of views into a
fixed-size scene representation without pose assumptions.
NerfDiff~\citep{gu2023nerfdiff} augments generalizable NeRF with diffusion
priors for single-image synthesis.
ZeroRF~\citep{shi2023zerorf} achieves sparse-view 360$^\circ$ reconstruction
via per-scene optimisation with zero pretraining, leveraging factorised feature grids.

\paragraph{Few-shot and test-time methods.}
FrameNeRF~\citep{yan2024framenerf} presents a simple and efficient framework
for few-shot NVS by exploiting temporal video frames as dense pseudo views.
Recollection from Pensieve~\citep{zhou2024pensieve} learns NVS from uncalibrated
internet videos without posed supervision.
Test-Time Completion~\citep{bao2024tttcompletion} repurposes natural video
completion as a zero-shot NVS prior at test time.
Mixed Neural Voxels~\citep{wang2023mixedvoxels} enables fast multi-view video
synthesis by decomposing scenes into static and dynamic voxel grids.

\subsection{Dynamic NeRF}

Modelling non-rigid dynamic scenes is a central challenge for world modeling.

D-NeRF~\citep{pumarola2021dnerf} introduces a deformation network mapping each
spacetime point to a canonical static space.
Nerfies~\citep{park2021nerfies} and
HyperNeRF~\citep{park2021hypernerf} model complex non-rigid deformations via
latent codes and higher-dimensional ambient spaces.
Neural Scene Flow Fields~(NSFF)~\citep{li2021neural_scene_flow} models dynamic
scenes with scene flow representation.
Non-Rigid NeRF~\citep{tretschk2021nonrigid} reconstructs and synthesises
novel views of dynamic scenes from monocular video with explicit ray bending.
DyNeRF~\citep{li2022neural3d} extends NeRF to monocular video via a
time-conditioned model.
TiNeuVox~\citep{fang2022tineuvox} uses voxel features to accelerate dynamic NeRF.
K-Planes~\citep{fridovich2023kplanes} decomposes 4-D space-time into six
feature planes---three spatial (XY, XZ, YZ) and three spatial-temporal
(XT, YT, ZT)---each encoded as a learned 2-D grid, making the architecture
highly memory-efficient and interpretable.
HexPlane~\citep{cao2023hexplane} independently proposes the same factorisation,
demonstrating its generality; together these works establish plane-based
4-D decomposition as the dominant acceleration paradigm for dynamic NeRF.
NeRFPlayer~\citep{song2023nerfplayer} separately encodes static, deforming and
newly appearing regions.
Soup-of-Planes~\citep{du2023soupofplanes} enables fast view synthesis of
casual monocular videos by fitting a set of planar scenes.
Pseudo-Generalised Dynamic View~\citep{zhao2024pgdvs} synthesises dynamic views
from monocular video by combining static and dynamic video inpainting.
Decoupled Dynamic Video~\citep{wang2024decoupleddynamic} disentangles foreground
dynamics and background statics for monocular dynamic view synthesis.

The dynamic NeRF methods are included in the comprehensive
Table~\ref{tab:nerf_methods} above.

% ─────────────────────────────────────────────────────────────────────────────
\section{3D Gaussian Splatting}
\label{sec:3dgs}

\subsection{Foundational 3DGS}

3DGS~\citep{kerbl2023gaussian} achieved real-time ($>$100\,fps at 1080p)
NVS with training times of 30--45 minutes on commodity hardware,
surpassing NeRF-quality baselines on standard benchmarks.
Its three key innovations are:
(i)~anisotropic Gaussian primitives that align naturally with planar scene
structures;
(ii)~tile-based rasterisation that enables massively parallel GPU rendering;
(iii)~adaptive densification that automatically allocates capacity to
under-reconstructed regions.
The resulting representation is fully explicit and editable, opening new
directions in scene editing, simulation, and avatar rendering.

\paragraph{Quality and anti-aliasing.}
Mip-Splatting~\citep{yu2024mipsplatting} identifies aliasing in vanilla 3DGS
as arising from the approximation of 3-D Gaussian convolution during 2-D
projection.
It introduces a 3-D smoothing filter derived from the sampling footprint,
enabling accurate multi-scale rendering and substantially reducing floaters
at large distances.
AbsGS~\citep{ye2024absgs} revisits the densification criterion: replacing
the mean gradient with an absolute-value measure improves coverage in
thin structures and reduces needle-like artifacts near object boundaries.

\paragraph{Compression and memory efficiency.}
A single 3DGS scene may require 300MB--1GB of storage, limiting deployment.
Scaffold-GS~\citep{lu2023scaffoldgs} introduces a sparse set of anchor
primitives from which neural networks predict neighbouring Gaussian attributes;
this structured representation reduces storage by $3$--$5\times$ with
comparable quality.
LightGaussian~\citep{fan2023lightgaussian} achieves $15\times$ compression via
global Gaussian pruning based on visibility contribution and subsequent
vector quantisation of attributes.
EAGLES~\citep{girish2024eagles} applies residual vector quantisation
for fine-grained attribute compression.
HAC~\citep{chen2024hac} proposes a hash-grid-based context model that predicts
per-Gaussian attribute entropy, enabling learned arithmetic coding for
extreme compression ratios.

\paragraph{Surface reconstruction.}
SuGaR~\citep{guedon2024sugar} aligns Gaussians to the scene surface by adding
a regularisation term that encourages Gaussians to become flat discs at
the surface, followed by mesh extraction and joint Gaussian--mesh refinement.
2DGS~\citep{huang2024_2dgs} replaces 3-D Gaussians with 2-D disc primitives,
which by construction lie on surfaces and produce accurate normal maps.
Gaussian Opacity Fields~(GOF)~\citep{yu2024gof} fuses Gaussian Splatting with
a continuous opacity field for topologically correct surface extraction in
unbounded scenes.

\paragraph{Scene editing and compositionality.}
The explicit Gaussian representation is particularly amenable to editing:
individual primitives can be identified, moved, duplicated, or removed.
GaussianEditor~\citep{chen2024gaussianeditor} harnesses a large vision-language
model to select semantically relevant Gaussians from a text description and
optimises their appearance with a diffusion prior.
Gaussian Grouping~\citep{ye2023gaussian_grouping} augments each Gaussian with
a learned identity field, trained with SAM~\citep{kirillov2023sam} masks from
rendered views, enabling instance-level part decomposition and editing.
GALA3D~\citep{zhou2024gala3d} uses an LLM to generate a 3-D layout from text
and initialises Gaussians per-object, enabling compositional scene generation.

Table~\ref{tab:3dgs_methods} provides a comprehensive view of the 3DGS family.

\begin{table*}[t]
  \centering
  \caption{Comprehensive 3D Gaussian Splatting method comparison.
    \textbf{Category}: STATIC=per-scene static, DYN=dynamic/deformable, FF=feed-forward generalizable.
    \textbf{RT}: real-time rendering ($>$30\,fps).
    \textbf{Comp.}: dedicated compression or storage reduction.
    \textbf{Surf.}: explicit surface extraction.
    \textbf{Unp.}: works on unposed images.
    \textbf{Input}: minimum input views for FF methods.}
  \label{tab:3dgs_methods}
  \scriptsize
  \setlength{\tabcolsep}{2.8pt}
  \begin{tabular}{@{}llcccccccp{3.8cm}@{}}
    \toprule
    \textbf{Method} & \textbf{Venue} & \textbf{Cat.}
      & \textbf{RT} & \textbf{Comp.} & \textbf{Surf.}
      & \textbf{Edit.} & \textbf{Unp.} & \textbf{Input}
      & \textbf{Key Contribution} \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Static per-scene}} \\
    3DGS~\citep{kerbl2023gaussian}               & TOG'23      & S & \yes& \no & \no & \no & \no & MV  & Tile-based Gaussian splatting \\
    Mip-Splatting~\citep{yu2024mipsplatting}     & CVPR'24     & S & \yes& \no & \no & \no & \no & MV  & 3-D smoothing filter, anti-alias \\
    AbsGS~\citep{ye2024absgs}                    & NeurIPS'24  & S & \yes& \no & \no & \no & \no & MV  & Absolute-gradient densification \\
    Scaffold-GS~\citep{lu2023scaffoldgs}         & CVPR'24     & S & \yes& \yes& \no & \no & \no & MV  & Anchor-driven neural attributes \\
    LightGaussian~\citep{fan2023lightgaussian}   & NeurIPS'23  & S & \yes& \yes& \no & \no & \no & MV  & Visibility pruning + VQ, 15$\times$ \\
    Compact3D~\citep{lee2024compact3d}           & arXiv'24    & S & \yes& \yes& \no & \no & \no & MV  & k-means codebook compression \\
    EAGLES~\citep{girish2024eagles}              & ECCV'24     & S & \yes& \yes& \no & \no & \no & MV  & Residual vector quantisation \\
    HAC~\citep{chen2024hac}                      & ECCV'24     & S & \yes& \yes& \no & \no & \no & MV  & Hash context entropy coding \\
    SuGaR~\citep{guedon2024sugar}                & CVPR'24     & S & \yes& \no & \yes& \pt & \no & MV  & Flat-disc + mesh joint opt. \\
    2DGS~\citep{huang2024_2dgs}                  & SIGGRAPH'24 & S & \yes& \no & \yes& \no & \no & MV  & 2-D disc primitives \\
    GOF~\citep{yu2024gof}                        & arXiv'24    & S & \yes& \no & \yes& \no & \no & MV  & Opacity field for surface \\
    GaussianEditor~\citep{chen2024gaussianeditor} & CVPR'24    & S & \yes& \no & \no & \yes& \no & MV  & LLM-guided semantic editing \\
    Gaussian Grouping~\citep{ye2023gaussian_grouping}& ECCV'24  & S & \yes& \no & \no & \yes& \no & MV  & SAM-based identity field \\
    GALA3D~\citep{zhou2024gala3d}                & ICML'24     & S & \yes& \no & \no & \yes& \no & MV  & LLM-layout compositional gen. \\
    LERF~\citep{kerr2023lerf}                    & ICCV'23     & S & \no & \no & \no & \pt & \no & MV  & CLIP language grounding in NeRF \\
    LangSplat~\citep{qin2024langsplat}           & CVPR'24     & S & \yes& \no & \no & \pt & \no & MV  & Language field in 3DGS \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Dynamic \& deformable}} \\
    4DGS~\citep{wu2024_4dgs}                     & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & 4-D Gaussian + temporal field \\
    Deformable3DGS~\citep{yang2024deformable}    & ICLR'24     & D & \yes& \no & \no & \no & \no & Mono& Canonical deformation network \\
    SpacetimeGaussians~\citep{yang2024spacetime} & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & Polynomial time trajectories \\
    SC-GS~\citep{huang2024scgs}                  & CVPR'24     & D & \yes& \no & \no & \no & \no & Mono& Sparse control point deform. \\
    Gaussian-Flow~\citep{lin2024gaussianflow}    & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & Fourier polynomial motion \\
    Shape-of-Motion~\citep{wang2024shapeofmotion}& arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Per-Gaussian 3-D motion \\
    Dynamic 3DGS~\citep{luiten2024dynamic3dgs}   & 3DV'24      & D & \yes& \no & \no & \no & \no & MV  & Persistent Gaussian tracking \\
    GauFRe~\citep{liang2024gaufre}               & arXiv'24    & D & \yes& \no & \no & \no & \no & Mono& Flow-regularised deformation \\
    MoVieS~\citep{sun2024movies}                 & arXiv'24    & D & \yes& \no & \no & \no & \no & Mono& 4-D synthesis in $\sim$1 second \\
    SC-4D~\citep{kim2024selfcal4d}               & arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Self-calibrating 4-D from video \\
    Reconstruct-Inpaint~\citep{han2024rif}       & arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Recon + inpaint + TTF pipeline \\
    Articulated-GS~\citep{li2024articulatedgs}   & NeurIPS'24  & D & \yes& \no & \no & \no & \no & MV  & Template-free articulated 3DGS \\
    EVA-Gaussian~\citep{hu2024evacg}             & arXiv'24    & D & \yes& \no & \no & \no & \no & MV  & Real-time human NVS \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Feed-forward generalizable}} \\
    pixelSplat~\citep{charatan2024pixelsplat}    & CVPR'24     & FF& \yes& \no & \no & \no & \no & 2   & Stereo epipolar features \\
    MVSplat~\citep{chen2024mvsplat}              & ECCV'24     & FF& \yes& \no & \no & \no & \no & 2--6& Cost-volume geometry prior \\
    TranSplat~\citep{zhang2024transplate}        & arXiv'24    & FF& \yes& \no & \no & \no & \no & 2+  & Transformer MV prediction \\
    FSGS~\citep{zhu2024fsgs}                     & CVPR'24     & FF& \yes& \no & \no & \no & \no & 3   & Gaussian unpooling, few-shot \\
    DepthSplat~\citep{xu2024depthsplat}          & arXiv'24    & FF& \yes& \no & \no & \no & \no & 2+  & Joint depth + Gaussian pred. \\
    Splatt3R~\citep{smart2024splatt3r}           & arXiv'24    & FF& \yes& \no & \no & \no & \yes& 2   & DUSt3R backbone, unposed \\
    TriplaneGaussian~\citep{zou2024triplanegaussian}& CVPR'24   & FF& \yes& \no & \no & \no & \no & 1   & Triplane + Gaussian, single-view \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Dynamic 3DGS}

4DGS~\citep{wu2024_4dgs} extends Gaussians into 4-D space-time with a
temporal radiance field.
SpacetimeGaussians~\citep{yang2024spacetime} assigns polynomial time
trajectories to each Gaussian.
Deformable3DGS~\citep{yang2024deformable} applies a deformation network to
canonical Gaussians.
SC-GS~\citep{huang2024scgs} uses sparse control points for monocular dynamic
scenes.
Gaussian-Flow~\citep{lin2024gaussianflow} models Gaussian dynamics via
polynomial motion fields in Fourier space.
Shape-of-Motion~\citep{wang2024shapeofmotion} reconstructs per-Gaussian 3-D
motion from monocular videos.
Dynamic 3D Gaussians~\citep{luiten2024dynamic3dgs} tracks scene dynamics
through persistent Gaussian views, jointly optimising NVS and tracking.
GauFRe~\citep{liang2024gaufre} uses Gaussian deformation fields with
flow regularisation for real-time dynamic NVS.
MoVieS~\citep{sun2024movies} generates motion-aware 4-D dynamic views in
approximately one second via efficient Gaussian motion estimation.
Self-Calibrating 4D~\citep{kim2024selfcal4d} performs 4-D NVS from monocular
videos by jointly estimating camera parameters and Gaussian deformations.
Reconstruct-Inpaint-Finetune~\citep{han2024rif} addresses dynamic monocular
NVS via a three-stage pipeline of reconstruction, inpainting, and test-time
refinement.
EVA-Gaussian~\citep{hu2024evacg} achieves real-time human NVS under diverse
multi-view settings.
Articulated Gaussian~\citep{li2024articulatedgs} enables template-free
articulated Gaussian Splatting for real-time reposable dynamic view synthesis.

\subsection{Feed-Forward Generalizable 3DGS}

pixelSplat~\citep{charatan2024pixelsplat} predicts Gaussian primitives from
stereo image pairs in a single forward pass.
MVSplat~\citep{chen2024mvsplat} adds cost-volume priors for better geometry.
TranSplat~\citep{zhang2024transplate} introduces a transformer architecture for
multi-view Gaussian prediction.
Splatt3R~\citep{smart2024splatt3r} builds on DUSt3R~\citep{wang2024dust3r}
for unposed image pairs.
FSGS~\citep{zhu2024fsgs} enables few-shot Gaussian Splatting via Gaussian
unpooling and monocular depth supervision.
DepthSplat~\citep{xu2024depthsplat} jointly predicts depth and Gaussians
for generalizable NVS.
TriplaneGaussian~\citep{zou2024triplanegaussian} combines triplane and
Gaussian representations in a transformer-based single-view reconstruction model.

The feed-forward generalizable methods are included in the comprehensive
Table~\ref{tab:3dgs_methods} above (FF category).

% ─────────────────────────────────────────────────────────────────────────────
\section{Generalisable Feed-Forward 3D Reconstruction}
\label{sec:generalisable}

\subsection{Pose-Free and Foundation Models}

A new generation of methods aims for \emph{pose-free} generalizable reconstruction.
DUSt3R~\citep{wang2024dust3r} predicts dense pointmaps from arbitrary image pairs
without camera calibration, then aligns them via global optimisation.
MASt3R~\citep{leroy2024mast3r} adds matching capabilities to DUSt3R, enabling
accurate localisation.
MonST3R~\citep{zhang2024monst3r} extends DUSt3R to dynamic video.
VGGT~\citep{wang2025vggt} is a visual geometry grounded deep structured tokens
model for feed-forward 3-D structure prediction.

\subsection{Large Reconstruction Models (LRM)}
\label{sec:lrm}

The Large Reconstruction Model (LRM)~\citep{hong2023lrm} paradigm represents
a qualitative shift from per-scene optimisation to \emph{feed-forward inference}:
a single forward pass of a large transformer, pretrained on hundreds of thousands
of 3-D assets, produces a NeRF or Gaussian scene from a single image in under
5 seconds.
This is possible because the model amortises the geometric reasoning of NeRF
optimisation into its weights---a striking demonstration that scale can substitute
for per-instance compute.

LRM~\citep{hong2023lrm} itself uses an image encoder (DINO-ViT) followed by
a cross-attention transformer that maps image tokens to triplane tokens, which
are then decoded by a NeRF MLP.
Instant3D~\citep{li2023instant3d} extends this by first generating sparse multi-view
images with a diffusion model, then feeding the result to an LRM-style reconstructor,
decoupling appearance generation from geometry prediction.
OpenLRM~\citep{openlrm2023} reproduces LRM in an open-source setting.

The Gaussian-based branch has proven especially fast.
LGM~\citep{tang2024lgm} generates multi-view images and lifts them to a 3DGS
scene via an asymmetric U-Net, achieving 3-D generation in seconds at
commercial-grade quality.
GRM~\citep{xu2024grm} predicts 3D Gaussians directly from image features using
a sparse-view transformer, operating entirely at the pixel level without any
intermediate NeRF.
Gamba~\citep{he2024gamba} replaces the transformer with a state-space model
(Mamba) for $O(N)$ sequence length scaling, enabling long-context 3-D inference.
GS-LRM~\citep{zhang2024gslrm} combines a transformer backbone with a 3DGS
output head, achieving state-of-the-art single-image reconstruction quality.

Two non-parametric reconstruction models have recently emerged as strong baselines.
DUSt3R~\citep{wang2024dust3r} formulates 3-D reconstruction as point-map
regression: given an image pair, a ViT predicts dense 3-D coordinates for
every pixel in a shared Euclidean frame, requiring no camera poses at test time.
MASt3R~\citep{leroy2024mast3r} extends DUSt3R to feature-level matching,
enabling accurate metric reconstruction from internet-sourced photo pairs.
These models underpin a new generation of unposed feed-forward reconstructors
such as Splatt3R~\citep{smart2024splatt3r}, which produces Gaussian splatting
scenes directly from uncalibrated image pairs.

\paragraph{Implications for World Modeling.}
LRM-style models demonstrate that a single transformer, with sufficient data
and compute, can learn a universal spatial prior---the same capacity required
by world models for instant scene initialisation from single-frame observations.

\subsection{Multi-View Stereo in the Neural Era}

Classical MVS has been reinvigorated by deep learning.
MVSNet~\citep{yao2018mvsnet} introduced cost-volume-based deep MVS.
IterMVS~\citep{wang2022itermvs} and Uni-MVSNet~\citep{peng2022unimvsnet}
improved accuracy and generalisation.
These methods provide the geometry backbone for downstream NVS, and their
cost-volume feature matching is directly incorporated into generalizable NeRF
(MVSNeRF~\citep{chen2021mvsnerf}) and feed-forward Gaussian methods
(MVSplat~\citep{chen2024mvsplat}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Generative Paradigms for Novel View Synthesis}
\label{sec:generative}

The emergence of deep generative models---first GANs and then diffusion
models---has fundamentally altered the landscape of \nvs{}.
Where reconstruction-based methods infer a scene representation from observations
and render it deterministically, \emph{generative} methods learn a distribution
over plausible visual completions given partial observations.
This distinction is not merely architectural: it represents a shift from
\emph{interpolation} (rendering within the observed convex hull of viewpoints)
to \emph{extrapolation} (hallucinating plausible content beyond it).
Generative methods are therefore the primary vehicle for the one-image-to-3D,
text-to-3D, and single-image world-generation capabilities that bring \nvs{}
into the world-modeling regime.

\subsection{3-D-Aware GANs: Learning Geometry from 2-D Supervision}

The first wave of generative \nvs{} methods embedded radiance fields inside
GAN generators, enabling implicit 3-D supervision from unposed 2-D image
collections---a paradigm that proved influential precisely because it
required no 3-D ground truth.

GRAF~\citep{schwarz2020graf} pioneered this direction by conditioning a NeRF
generator on a latent shape-appearance code, training with a 2-D discriminator.
pi-GAN~\citep{chan2021pigan} improved the NeRF backbone with periodic activations
(SIREN~\citep{sitzmann2020siren}), achieving substantially better multi-view
consistency.
GIRAFFE~\citep{niemeyer2021giraffe} introduced scene compositionality by
combining foreground and background neural feature fields, enabling object-level
editing without explicit segmentation.
EG3D~\citep{chan2022efficient} is the landmark method of this era: its
\emph{triplane} representation---three axis-aligned 2-D feature maps decoded by
a shallow MLP---reduces rendering cost by orders of magnitude and admits a
super-resolution upsampler, enabling 512$\times$512 training at practical
throughput.
StyleNeRF~\citep{gu2022stylenerf} integrates the NeRF volume renderer into a
StyleGAN~\citep{karras2019stylegan2} backbone, inheriting its powerful
style-based latent structure for fine-grained appearance control.

A critical limitation of GAN-based approaches is \emph{mode coverage}: the
discriminator objective encourages sharp, realistic individual samples but does
not explicitly minimise distributional divergence over all plausible views,
making it prone to mode dropping for complex, asymmetric scenes.
This motivated the transition toward diffusion-based approaches.

\subsection{Diffusion Models for Camera-Controlled NVS}
\label{sec:diffusion_nvs}

Diffusion models~\citep{rombach2022ldm,ho2020ddpm} have become the
dominant generative paradigm for \nvs{} due to three decisive advantages:
(i)~training stability via a simple denoising objective;
(ii)~access to powerful image priors from large-scale pretraining;
(iii)~natural integration of arbitrary conditioning signals (pose, depth,
text, reference image) through classifier-free guidance.

\subsubsection{Pose-Conditioned Single-Image NVS}

The most direct application is \emph{single-image NVS}: given one reference
photograph and a target camera pose $\boldsymbol{\pi}^*$, generate a plausible
rendering at that viewpoint.
This is under-constrained without a prior---an infinite number of 3-D scenes
are consistent with a single 2-D observation---and diffusion models provide
exactly the distributional prior needed.

\paragraph{Zero-1-to-3 and successors.}
Zero-1-to-3~\citep{liu2023zero1to3} established the canonical formulation:
fine-tune a latent diffusion model (Stable Diffusion~\citep{rombach2022ldm})
on multi-view renders from Objaverse~\citep{deitke2023objaverse}, conditioning
on both the reference image and the relative pose $(\Delta\phi, \Delta\theta,
\Delta r)$ encoded as a CLIP-space vector.
The model learns to ``imagine'' how an object would look from any viewpoint,
without constructing an explicit 3-D representation.
Zero123++~\citep{shi2023zero123pp} strengthens multi-view consistency by
predicting a fixed grid of six canonical views simultaneously, reducing
inter-view inconsistency from independent sampling.
One-2-3-45~\citep{liu2023one2345} pipelines Zero-1-to-3 outputs into a
cost-volume-based MVS reconstruction~\citep{yao2018mvsnet}, converting the
diffusion output to a textured mesh within 45 seconds.

\paragraph{Synchronised multi-view denoising.}
A key challenge with independently conditioned diffusion is that each generated
view is sampled from a different realisation of the latent distribution,
producing inter-view geometric contradictions.
SyncDreamer~\citep{liu2023syncdreamer} addresses this by coupling the denoising
of multiple target views through shared 3-D volume attention at each diffusion
step, enforcing consistent geometry throughout the synthesis process.
MVDiffusion~\citep{tang2023mvdiffusion} constructs a correspondence-aware attention
mechanism that explicitly queries matching pixels across views, enabling
holistic panorama generation from a set of overlapping perspective images.
Wonder3D~\citep{long2024wonder3d} takes a two-domain approach: a cross-domain
diffusion model simultaneously denoises an RGB and a surface-normal image,
with the normal output serving as high-quality geometric supervision for
downstream mesh reconstruction.

\paragraph{Scalable and pose-free generation.}
EscherNet~\citep{kong2024eschernet} generalises pose conditioning via a
novel \emph{camera positional encoding}~(CaPE), enabling generation of
an arbitrary number of views at any resolution without per-angle fine-tuning.
ViVid-1-to-3~\citep{kwak2024vivid} recasts NVS as a video generation problem:
the temporal axis of a pre-trained video diffusion model is repurposed to model
the viewpoint axis, using the model's temporal consistency as a proxy for
multi-view coherence.
GeoNVS~\citep{chen2024geonvs} addresses geometric fidelity by conditioning a
video diffusion model on a rendered point cloud at the target pose, grounding
the generative hallucination in explicit 3-D geometry.

\subsubsection{Multi-View Diffusion and 3-D-Consistent Generation}

Single-view methods generate views independently or with limited cross-view
coupling.
A more ambitious goal is to train a diffusion model that \emph{natively}
produces multi-view consistent outputs---effectively encoding 3-D structure
implicitly in the model weights.

MVDream~\citep{shi2023mvdream} achieves this by replacing the temporal
self-attention in a video diffusion transformer with a 3-D-aware \emph{multi-view
attention} block that attends across all views simultaneously.
Training on rendered 3-D assets (Objaverse) with known camera poses teaches the
model to share structural information across views.
The resulting model can be used both as a standalone multi-view generator and
as a geometry-consistent score function for SDS-based text-to-3D.

GeNVS~\citep{chan2023genvs} takes a hybrid approach: a NeRF encoder constructs
a 3-D feature volume from the conditioning images, and the diffusion decoder
renders a target view by volumetrically querying this feature space.
This architecture provides hard geometric grounding while preserving the
expressiveness of the diffusion prior for high-frequency appearance details.

\subsubsection{Camera Trajectory Conditioning and Video Diffusion NVS}

A particularly important frontier is conditioning diffusion models on
\emph{continuous camera trajectories}, enabling the generation of smooth,
temporally coherent novel-view sequences---the direct bridge to world modeling.

\paragraph{Camera trajectory as a control signal.}
ViewCrafter~\citep{yu2024viewcrafter} conditions a video diffusion model
(CogVideoX) on point-cloud renderings along a target camera path, using
the rendered point cloud as a structural scaffold that prevents temporal
drift while the diffusion model fills in appearance and fine geometry.
Free3D~\citep{zheng2023free3d} injects camera pose into each frame's
cross-attention query, enabling a single video diffusion model to follow
arbitrary camera trajectories from a single reference image.
Novel View Extrapolation~\citep{shi2024nve} specifically targets
\emph{large-angle extrapolation}---the regime where geometric hallucination
is most necessary---using video diffusion priors to maintain perceptual
plausibility when the reference content provides little direct evidence.

\paragraph{Training-free and zero-shot camera control.}
NVS-Solver~\citep{you2024nvssolver} demonstrates that an off-the-shelf video
diffusion model already encodes sufficient geometric knowledge to solve
NVS in a zero-shot manner: by formulating view synthesis as a constrained
diffusion process with known-region guidance, it achieves competitive performance
without any pose-specific fine-tuning.
DDIM Inversion for NVS~\citep{xu2024ddim_nvs} exploits the deterministic
reconstruction property of DDIM~\citep{song2020ddim}: inverting the reference
image to its latent code and re-decoding with a modified pose conditioning
produces structurally faithful novel views, again without additional training.

\paragraph{Geometric conditioning for high fidelity.}
Projected Representation Conditioning~\citep{han2024prc} proposes projecting
3-D features from the reference view onto the target-view image plane,
providing an explicit geometric ``warp'' prior to the diffusion decoder.
This projected feature conditioning substantially reduces hallucination artifacts
in regions visible from the target but occluded in the reference.
A Single Image and Multimodality~\citep{tang2024smm} augments pose conditioning
with depth, surface normal, and semantic maps, each derived from monocular
estimation, to enrich the conditioning signal across multiple complementary
geometric cues.

\subsubsection{Reconstruction--Diffusion Hybrids}

A complementary design philosophy \emph{separates} the diffusion and reconstruction
stages: diffusion fills in missing views and hallucinations, while a deterministic
reconstruction pipeline (NeRF or 3DGS) enforces 3-D consistency.

Cat3D~\citep{gao2024cat3d} exemplifies this pattern: given sparse reference
views ($n\geq1$), a multi-view diffusion model generates a dense set of
synthetic pseudo-views, which are then fed to a standard 3DGS reconstruction
pipeline.
Because the diffusion model is fine-tuned jointly with the reconstruction loss,
the generated views are explicitly optimised for downstream reconstruction fidelity.
ReconFusion~\citep{wu2024reconfusion} integrates the diffusion prior more tightly:
it conditions a pixel-diffusion model on pixelNeRF~\citep{yu2021pixelnerf}
feature queries, acting as a learned regulariser that fills in uncertain regions
during sparse-view NeRF optimisation.
ZeroNVS~\citep{sargent2023zeronvs} handles the challenging case of
in-the-wild single images (non-object-centric, background present) by training
on scene-level data and conditioning on the camera intrinsics alongside pose,
enabling 360$^\circ$ generation from a single casual photograph.
NVComposer~\citep{li2024nvcomposer} addresses the case where multiple reference
views are available but unposed: a feature composition network aggregates
information from all reference images before passing to the diffusion model,
improving quality without requiring explicit pose estimation.

\subsection{Score Distillation: Diffusion as a 3-D Optimisation Oracle}

Score Distillation Sampling (SDS)~\citep{poole2023dreamfusion} opened a
separate pathway: instead of directly generating novel views, use a
pre-trained 2-D diffusion model as a differentiable \emph{score function}
to optimise a 3-D representation.
The core identity is:
\begin{equation}
  \nabla_\theta \mathcal{L}_\text{SDS}
  = \mathbb{E}_{t,\boldsymbol{\epsilon}}\!\left[
    w(t)\!\left(
      \hat{\boldsymbol{\epsilon}}_\phi(\mathbf{x}_t; y, t)
      - \boldsymbol{\epsilon}
    \right)\frac{\partial \mathbf{x}}{\partial \theta}
  \right],
  \label{eq:sds}
\end{equation}
where $\theta$ parameterises the 3-D representation, $\mathbf{x}$ is a rendered
image, $y$ is a text prompt, and $\hat{\boldsymbol{\epsilon}}_\phi$ is the
diffusion model's noise prediction.

DreamFusion~\citep{poole2023dreamfusion} first demonstrated that SDS can
synthesise coherent 3-D objects from text descriptions.
Magic3D~\citep{lin2023magic3d} introduces a coarse-to-fine strategy that
first optimises a coarse NeRF then refines a differentiable mesh, achieving
$8\times$ higher resolution.
ProlificDreamer~\citep{wang2023prolificdreamer} identifies the ``over-saturation''
artifact of SDS as a consequence of mode-seeking behaviour, proposing
\emph{Variational Score Distillation} (VSD) which models the 3-D parameter
distribution as a particle set and applies Stein variational gradient descent
for diverse, high-quality generation.
Fantasia3D~\citep{chen2023fantasia3d} disentangles geometry (optimised via
SDS on a DMTet mesh) from appearance (a physically-based BRDF), enabling
photorealistic materials and independent editing of shape and texture.
GaussianDreamer~\citep{yi2024gaussiandreamer} replaces NeRF with 3DGS as the
SDS target, leveraging Gaussian densification for fine detail and achieving
real-time rendering of the final asset.
LucidDreamer~\citep{chung2023luciddreamer} addresses SDS instability with an
interval-based score matching scheme that reduces gradient variance.

\subsection{Dedicated Multi-View Diffusion Architectures}

The methods above predominantly adapt existing video or image diffusion models.
A complementary line of work designs architectures \emph{from scratch} for
multi-view consistency.

MVDream~\citep{shi2023mvdream} replaces temporal attention with multi-view
attention and trains on Objaverse with ground-truth camera poses, producing
natively 3-D-consistent outputs.
ImageDream~\citep{wang2024imagedream} extends MVDream to image-conditioned
generation, enabling single-image-to-3D with strong semantic fidelity.
Zero123-XL~\citep{zero123xl2023} scales the Zero-1-to-3 paradigm to larger
model capacities and more diverse training data.
Direct3D~\citep{wu2024direct3d} trains a native 3-D latent diffusion model
on a triplane latent space, enabling direct 3-D generation without a two-stage
pipeline.

Table~\ref{tab:generative_methods} provides a structured comparison of generative
\nvs{} methods across the four paradigms above.

\begin{table}[t]
  \centering
  \caption{Generative NVS methods. \textbf{Backbone}: LDM=latent diffusion,
    VDM=video diffusion, GAN=generative adversarial.
    \textbf{Ctrl}: pose conditioning mechanism.
    \textbf{3D Out}: whether the method produces an explicit 3-D representation.}
  \label{tab:generative_methods}
  \small
  \setlength{\tabcolsep}{2pt}
  \begin{tabular}{@{}llllccc@{}}
    \toprule
    \textbf{Method} & \textbf{Venue}
      & \textbf{Backbone} & \textbf{Ctrl}
      & \textbf{3D Cons.} & \textbf{3D Out} & \textbf{Key Contribution} \\
    \midrule
    \multicolumn{7}{@{}l}{\textit{GAN-based}} \\
    GRAF~\citep{schwarz2020graf}                & NeurIPS'20 & NeRF-GAN   & Latent  & \pt & NeRF   & Radiance field GAN \\
    pi-GAN~\citep{chan2021pigan}                & CVPR'21    & NeRF-GAN   & Latent  & \pt & NeRF   & Periodic implicit \\
    GIRAFFE~\citep{niemeyer2021giraffe}         & CVPR'21    & Feat. Field& Latent  & \pt & Neural & Compositional \\
    EG3D~\citep{chan2022efficient}              & CVPR'22    & Triplane   & Latent  & \yes& Tri.   & Triplane+SR \\
    StyleNeRF~\citep{gu2022stylenerf}           & ICLR'22    & StyleGAN   & Latent  & \pt & NeRF   & Style-based NeRF \\
    \midrule
    \multicolumn{7}{@{}l}{\textit{Pose-conditioned single-image diffusion}} \\
    Zero-1-to-3~\citep{liu2023zero1to3}         & ICCV'23    & LDM        & $\Delta$pose & \pt & -- & Pose-cond LDM \\
    Zero123++~\citep{shi2023zero123pp}          & arXiv'23   & LDM        & $\Delta$pose & \pt & -- & 6-view grid \\
    One-2-3-45~\citep{liu2023one2345}           & NeurIPS'23 & LDM+MVS    & $\Delta$pose & \yes& Mesh & Diff$\to$MVS \\
    SyncDreamer~\citep{liu2023syncdreamer}      & ICLR'24    & LDM        & $\Delta$pose & \yes& -- & Volume sync \\
    Wonder3D~\citep{long2024wonder3d}           & CVPR'24    & LDM        & $\Delta$pose & \yes& -- & RGB+Normal \\
    EscherNet~\citep{kong2024eschernet}         & CVPR'24    & LDM        & CaPE    & \yes& -- & Arbitrary \#views \\
    ViVid-1-to-3~\citep{kwak2024vivid}          & arXiv'24   & VDM        & $\Delta$pose & \pt & -- & Video=NVS axis \\
    \midrule
    \multicolumn{7}{@{}l}{\textit{Camera trajectory / video diffusion}} \\
    Free3D~\citep{zheng2023free3d}              & ICLR'24    & VDM        & Cam embed& \pt & -- & Per-frame pose \\
    ViewCrafter~\citep{yu2024viewcrafter}       & arXiv'24   & VDM        & Ptcloud  & \yes& -- & Scaffold+VDM \\
    NVS-Solver~\citep{you2024nvssolver}         & arXiv'24   & VDM        & Zero-shot& \pt & -- & Guided diffusion \\
    DDIM Inversion~\citep{xu2024ddim_nvs}       & arXiv'24   & LDM        & Inversion& \pt & -- & Training-free \\
    GeoNVS~\citep{chen2024geonvs}               & arXiv'24   & VDM        & Ptcloud  & \yes& -- & Geo-grounded VDM \\
    Proj. Repr. Cond.~\citep{han2024prc}        & arXiv'24   & LDM        & 3D proj. & \yes& -- & Feature warp \\
    Novel View Extrap.~\citep{shi2024nve}       & arXiv'24   & VDM        & Cam traj.& \pt & -- & Large-angle extrap.\\
    \midrule
    \multicolumn{7}{@{}l}{\textit{Reconstruction--diffusion hybrids}} \\
    GeNVS~\citep{chan2023genvs}                 & ICLR'23    & LDM        & NeRF feat& \yes& NeRF & NeRF-cond diff. \\
    ZeroNVS~\citep{sargent2023zeronvs}          & CVPR'24    & LDM        & Pose+int.& \pt & -- & In-the-wild \\
    ReconFusion~\citep{wu2024reconfusion}       & CVPR'24    & LDM        & pixelNeRF& \yes& NeRF & Diff reg. \\
    Cat3D~\citep{gao2024cat3d}                  & NeurIPS'24 & MV-Diff    & $\Delta$pose & \yes& 3DGS& MV gen+recon \\
    NVComposer~\citep{li2024nvcomposer}         & arXiv'24   & LDM        & Feature  & \yes& -- & Unposed multi-ref\\
    \midrule
    \multicolumn{7}{@{}l}{\textit{Score distillation (text-to-3D)}} \\
    DreamFusion~\citep{poole2023dreamfusion}    & ICLR'23    & LDM+NeRF   & SDS      & \pt & NeRF & SDS \\
    Magic3D~\citep{lin2023magic3d}              & CVPR'23    & LDM+Mesh   & SDS      & \pt & Mesh & C2F SDS \\
    ProlificDreamer~\citep{wang2023prolificdreamer}& NeurIPS'23& LDM+NeRF & VSD      & \pt & NeRF & VSD \\
    Fantasia3D~\citep{chen2023fantasia3d}       & ICCV'23    & LDM+Mesh   & SDS      & \pt & BRDF & Disentangled \\
    GaussianDreamer~\citep{yi2024gaussiandreamer}& CVPR'24   & LDM+3DGS   & SDS      & \pt & 3DGS & SDS+3DGS \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Implications for World Modeling.}
Diffusion-based \nvs{} represents a critical advance along World Modeling
Axis~2 (sparse/single-image generalisation, \textbf{D2}) and provides
the generative prior essential for completing unobserved scene regions
(Axis~1, geometric consistency \textbf{D1}).
Camera trajectory conditioning (\S\ref{sec:diffusion_nvs}) directly bridges
into Axis~3 (temporal dynamics), as a camera trajectory can be viewed as
a special case of a time-ordered action sequence.
The remaining gap is \emph{physical action conditioning} (\textbf{D4, D5}):
existing diffusion NVS methods condition on camera parameters but not on
object-level physical interactions, which remains an open problem
addressed in \S\ref{sec:open}.

% ─────────────────────────────────────────────────────────────────────────────
\section{Towards Generative World Modeling}
\label{sec:world}

\subsection{Dynamic Scenes and 4-D World Representation}

The step from static \nvs{} to world modeling hinges on modelling
\emph{temporal dynamics}: how does a scene evolve over time, and can
the model generate novel views at arbitrary future timestamps?
The dynamic NeRF and 4DGS methods in §\ref{sec:nerf}--\ref{sec:3dgs}
address this for \emph{observed} scenes given dense multi-view video.
A harder and more practically relevant regime is \emph{generalizable}
4-D generation from sparse or single-image input---precisely the regime
inhabited by world models.

Video diffusion models have emerged as the most promising backbone.
ViewCrafter~\citep{yu2024viewcrafter} augments a video diffusion model with
point-cloud renderings along the target camera trajectory, providing a
geometric scaffold that constrains the generation while the diffusion prior
fills in appearance and fine detail.
CAT4D~\citep{wu2024cat4d} generalises the Cat3D framework to the temporal
dimension, jointly modelling viewpoint and time with a multi-view video
diffusion model, then lifting the output to an explicit 4-D representation.
4DiM~\citep{watson2024_4dim} takes a more direct approach, training a
4-D diffusion model that simultaneously denoises a set of frames at
varying timestamps and viewpoints, learning a joint spatiotemporal prior.
SV3D~\citep{voleti2024sv3d} and SV4D~\citep{xie2024sv4d} extend Stable
Video Diffusion to 3-D and 4-D object generation respectively, conditioning
on orbital camera paths to ensure multi-view consistency.
Consistent4D~\citep{jiang2024consistent4d} enforces temporal consistency
across a 4-D cascade, resolving inter-frame Gaussian drift.
DreamGaussian4D~\citep{ren2023dreamgaussian4d} shows that SDS can be applied
to 4DGS, enabling text-driven 4-D generation without any video training data.

A complementary line of work addresses the \emph{monocular} case.
Diffusion Priors for Dynamic NVS~\citep{li2024diffpriordynamic} treats the
diffusion model as a frame-level regulariser during per-scene 4-D optimisation
from a single monocular video, substantially improving novel-view quality in
under-constrained regions.

\subsection{Neural Scene Simulators for Autonomous Driving}

Autonomous driving provides a uniquely demanding test-bed for NVS-as-world-model:
scenes are large-scale, unbounded, and contain independently moving actors
(vehicles, pedestrians) that must be rendered consistently from closed-loop
ego-motion trajectories.
Three architectural strategies have emerged.

\paragraph{NeRF-based compositional simulators.}
MARS~\citep{wu2023mars} separates the scene into a static NeRF background
and per-agent NeRF foregrounds, allowing independent actor manipulation.
UniSim~\citep{yang2023unisim} extends this to a closed-loop sensor simulator
that supports controllable scene editing---removing actors, changing weather,
inserting novel agents---with photorealistic rendering at the ego sensor.
EmerNeRF~\citep{yang2023emernerf} learns the dynamic-static decomposition
in a self-supervised manner by decomposing scene flow from multi-camera video,
without per-actor instance labels.

\paragraph{3DGS-based driving reconstruction.}
Street Gaussians~\citep{yan2024streetgaussian} extends 3DGS to urban scenes
by representing tracked foreground objects with separate Gaussian models and
compositing them over a background Gaussian field.
PVG~\citep{chen2023pvg} models Gaussian motion with periodic vibration functions
that efficiently encode slow, cyclic motions common in urban traffic.
OmniRe~\citep{chen2024omnire} unifies background, foreground, and sky in a
single Gaussian framework, establishing a strong all-in-one baseline.

\paragraph{Diffusion-based street generation.}
StreetCrafter~\citep{yan2024streetcrafter} uses a point-cloud-conditioned video
diffusion model with controllable camera motion, enabling photorealistic
street-view synthesis for data augmentation.
Sat2Vid~\citep{wu2022sat2vid} lifts a satellite image to a street-view
panoramic video, addressing the altitude gap between aerial mapping and
ground-level perception.
Wild-GS~\citep{xu2024wildgs} solves the unconstrained internet-photo case by
disentangling transient and persistent Gaussian appearance in 3DGS.
InSpatio-WorldFM~\citep{chen2025inspatioworld} is an open-source real-time
generative frame model targeting spatial world understanding at scale.

\subsection{Generative World Models and Foundation Models}

World foundation models represent the apex of the NVS-to-world-modeling
trajectory: systems that jointly handle viewpoint, time, and action, trained
at internet scale.
We identify three design axes that differentiate current approaches.

\paragraph{Sequence vs.\ spatial representation.}
GAIA-1~\citep{hu2023gaia} treats driving scenes as sequences of video, text,
and action tokens, training an autoregressive transformer to predict the next
token.
This approach achieves strong temporal coherence but lacks an explicit 3-D
representation, limiting viewpoint flexibility.
Genie~\citep{bruce2024genie} learns a latent action space from unsupervised
internet videos, enabling interactive game-world generation without action labels---an
important step toward self-supervised world model pretraining.

\paragraph{3-D-grounded generation.}
Genie~2~\citep{parkerholder2024genie2} introduces persistent 3-D structure
into the Genie framework, generating scenes that maintain geometry across
viewpoints and support first-person interactive navigation---the closest
current system to an NVS-grounded world model for open-ended environments.
WonderWorld~\citep{yu2024wonderworld} takes a complementary approach:
instead of learning from large-scale video, it uses instant 3DGS initialisation
from a single image combined with diffusion inpainting to generate interactive
3-D scenes on the fly.
WonderJourney~\citep{yu2023wonderjourney} chains text-guided scene generation
steps to produce perpetual explorable 3-D environments.

\paragraph{Physical-world foundation models.}
Cosmos~\citep{agarwal2025cosmos} is the most ambitious current effort:
a family of world foundation models pretrained on internet-scale video with
explicit physical world conditioning, designed for downstream fine-tuning
to robotics and autonomous driving.
Pandora~\citep{xiang2024pandora} interleaves language and video tokens,
enabling language-driven world simulation with both visual and linguistic output.

Table~\ref{tab:worldmodels} compares world modeling methods.

\begin{table}[t]
  \centering
  \caption{World modeling and neural scene simulation methods.}
  \label{tab:worldmodels}
  \scriptsize
  \setlength{\tabcolsep}{3pt}
  \begin{tabular}{@{}llccccc@{}}
    \toprule
    \textbf{Method} & \textbf{Year}
      & \textbf{3D Aware} & \textbf{Video} & \textbf{Action}
      & \textbf{Domain} & \textbf{Key Feature} \\
    \midrule
    GAIA-1~\citep{hu2023gaia}                  & 2023 & \pt  & \yes & \yes & Driving    & Seq. token WM \\
    UniSim~\citep{yang2023unisim}              & 2023 & \yes & \yes & \yes & Driving    & Sensor sim. \\
    MARS~\citep{wu2023mars}                    & 2023 & \yes & \yes & \pt  & Driving    & Compositional \\
    Genie~\citep{bruce2024genie}               & 2024 & \no  & \yes & \yes & Games      & Latent action \\
    Genie~2~\citep{parkerholder2024genie2}     & 2024 & \yes & \yes & \yes & Open-world & Interactive 3D \\
    WonderWorld~\citep{yu2024wonderworld}      & 2024 & \yes & \no  & \no  & Open-world & Instant 3DGS \\
    ViewCrafter~\citep{yu2024viewcrafter}      & 2024 & \yes & \yes & \no  & Open-world & Cam-cond video \\
    Cosmos~\citep{agarwal2025cosmos}           & 2025 & \yes & \yes & \yes & Physical   & WFM training \\
    Street Gaussians~\citep{yan2024streetgaussian}& 2024& \yes & \yes & \pt & Driving   & 3DGS driving \\
    OmniRe~\citep{chen2024omnire}              & 2024 & \yes & \yes & \yes & Driving    & All-in-one \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Embodied AI and Robotics}

NVS is more than a rendering tool in robotics: it provides the
\emph{spatial prior} that allows agents to reason about unseen parts of the
environment before taking action.
Three use patterns have emerged.

\textbf{Pose estimation and object localisation.}
iNeRF~\citep{yen2021inerf} inverts a trained NeRF to estimate 6-DoF camera
pose by gradient descent on the rendered photometric error---the first
demonstration that NeRF's differentiable renderer can be repurposed for
inference as well as reconstruction.
Dex-NeRF~\citep{ichnowski2021dex} applies NeRF to transparent-object grasping,
where conventional depth sensors fail, by leveraging NeRF's ability to
model view-dependent appearance to infer geometry from specular reflections.

\textbf{Visual planning and manipulation.}
NeRF-Supervisor~\citep{ze2023nerf_supervisor} uses a pre-trained NeRF to
synthesise robot-wrist camera views at hypothetical future poses, enabling
visual planning without physical rollouts.
GaussianGrasping~\citep{shorinwa2024gaussiangrasping} replaces NeRF with 3DGS
for real-time manipulation planning, exploiting the explicit Gaussian representation
for fast collision checking and grasp pose optimisation.

\textbf{Simulation-to-real transfer.}
RoboGSim~\citep{li2024robogssim} reconstructs robot workspaces in 3DGS and
renders synthetic training demonstrations from novel viewpoints, directly
closing the sim-to-real gap for visuomotor policy learning.

\subsection{Human Novel View Synthesis}

Synthesising novel views of humans is a critical subproblem with applications
in AR/VR, telepresence, and embodied AI.

\paragraph{Neural implicit human models.}
Neural Body~\citep{peng2021neuralbody} anchors structured latent codes on a
parametric human body (SMPL), enabling novel view and pose synthesis.
Rotationally-Temporally Consistent~\citep{liu2021rtcnvs} enforces both
rotation and temporal consistency for human performance video synthesis.
NHR~\citep{wu2020nhr} adopts multi-view point-cloud-based rendering for
dynamic human synthesis.

\paragraph{Generalizable and diffusion-based human NVS.}
CFSynthesis~\citep{li2023cfsynth} enables controllable and free-view 3-D human
video synthesis via a disentangled representation.
iVS-Net~\citep{chen2024ivsnet} learns human view synthesis from internet
videos without any 3-D supervision.
Blur2Sharp~\citep{dong2024blur2sharp} addresses human novel pose and view
synthesis from blurry input with a generative prior.
Diffuman4D~\citep{zheng2024diffuman4d} achieves 4-D consistent human view
synthesis from sparse-view videos via video diffusion.
EVA-Gaussian~\citep{hu2024evacg} achieves real-time 3DGS-based human NVS
under diverse multi-view settings.

\subsection{The Boundary: NVS vs.\ World Modeling}

Table~\ref{tab:nvs_wm} delineates the capability boundary.

\begin{table}[t]
  \centering
  \caption{NVS vs.\ World Modeling capability boundary.}
  \label{tab:nvs_wm}
  \small
  \setlength{\tabcolsep}{5pt}
  \begin{tabular}{@{}lcc@{}}
    \toprule
    \textbf{Capability} & \textbf{NVS (static)} & \textbf{World Model} \\
    \midrule
    Viewpoint-consistent rendering~(D1)   & \yes & \yes \\
    Sparse/single-image input~(D2)        & \pt  & \yes \\
    Temporal dynamics~(D3)                & \no  & \yes \\
    Action/agent conditioning~(D4)        & \no  & \yes \\
    Physical plausibility~(D5)            & \no  & \yes \\
    Cross-scene generalisation            & \pt  & \yes \\
    Causal scene understanding            & \no  & \yes \\
    \bottomrule
  \end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Datasets and Benchmarks}
\label{sec:datasets}

\subsection{Object-Level Datasets}

\textbf{ShapeNet}~\citep{chang2015shapenet}: 51K 3-D CAD models; foundation for
single-image reconstruction training.
\textbf{Objaverse}~\citep{deitke2023objaverse}: 800K+ 3-D objects;
enables LRM-scale training.
\textbf{Objaverse-XL}~\citep{deitke2024objaversexl}: 10M+ 3-D assets.
\textbf{Co3D}~\citep{reizenstein2021co3d}: 19K+ annotated per-category real video
sequences for category-level NVS.
\textbf{MVImgNet}~\citep{yu2023mvimgnet}: 238K multi-view real images across
238 categories.

\subsection{Static Scene Datasets}

\textbf{NeRF-Synthetic}~\citep{mildenhall2020nerf}: 8 synthetic bounded objects.
\textbf{LLFF}~\citep{mildenhall2019llff}: 8 forward-facing real scenes.
\textbf{DTU}~\citep{jensen2014dtu}: 124 controlled MVS scenes.
\textbf{Tanks and Temples}~\citep{knapitsch2017tanks}: large-scale indoor/outdoor.
\textbf{Mip-NeRF~360}~\citep{barron2022mipnerf360}: 9 unbounded 360$^\circ$ scenes.
\textbf{ScanNet}~\citep{dai2017scannet}: 1513 indoor RGB-D sequences.
\textbf{ScanNet++}~\citep{yeshwanth2023scannetpp}: DSLR + iPhone high-resolution indoor.
\textbf{RealEstate10K}~\citep{zhou2018stereoMagnification}: 10K real estate videos
for generalizable NVS training.

\subsection{Dynamic Scene Datasets}

\textbf{D-NeRF}~\citep{pumarola2021dnerf}: 6 monocular synthetic dynamic objects.
\textbf{HyperNeRF}~\citep{park2021hypernerf}: monocular topological changes.
\textbf{NHR}~\citep{wu2020nhr}: multi-view human performance.
\textbf{CMU Panoptic}~\citep{joo2015panoptic}: dense multi-view social activities.
\textbf{Neu3D}~\citep{li2022neural3d}: multi-view dynamic human sequences.
\textbf{iPhone Dataset}~\citep{guo2022neural}: casually-captured phone videos.

\subsection{Autonomous Driving and Large-Scale}

\textbf{Waymo Open}~\citep{sun2020waymo}: multi-camera + LiDAR driving.
\textbf{nuScenes}~\citep{caesar2020nuscenes}: 360$^\circ$ driving + HD maps.
\textbf{KITTI-360}~\citep{liao2022kitti360}: urban driving with panoptic labels.
\textbf{Argoverse~2}~\citep{wilson2023argoverse2}: high-resolution 360$^\circ$ driving.
\textbf{PandaSet}~\citep{xiao2021pandaset}: urban LiDAR+camera dataset.

Table~\ref{tab:datasets} provides a structured comparison of key datasets.

\begin{table*}[t]
  \centering
  \caption{NVS and world modeling datasets.
    \textbf{Type}: RT=real, Syn=synthetic.
    \textbf{Dyn.}: contains dynamic content.
    \textbf{Depth}: provides ground-truth or LiDAR depth.
    \textbf{Pose}: provides camera/LiDAR extrinsics.
    \textbf{Sem.}: provides semantic / instance labels.
    \textbf{Scale}: approximate number of scenes/objects/sequences.
    \textbf{Primary use}: main application in the NVS literature.}
  \label{tab:datasets}
  \scriptsize
  \setlength{\tabcolsep}{2.5pt}
  \begin{tabular}{@{}llccccccp{3.5cm}@{}}
    \toprule
    \textbf{Dataset} & \textbf{Year} & \textbf{Type}
      & \textbf{Dyn.} & \textbf{Depth} & \textbf{Pose}
      & \textbf{Sem.} & \textbf{Scale}
      & \textbf{Primary Use} \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Object-centric}} \\
    ShapeNet~\citep{chang2015shapenet}               & 2015 & Syn & \no & \yes & \yes & \yes & 51K obj.   & Generative 3-D training \\
    NeRF-Synthetic~\citep{mildenhall2020nerf}        & 2020 & Syn & \no & \yes & \yes & \no  & 8 obj.     & Per-scene NeRF evaluation \\
    Co3Dv2~\citep{reizenstein2021co3d}               & 2021 & RT  & \pt & \no  & \yes & \yes & 19K seq.   & Category-level NVS \\
    Objaverse~\citep{deitke2023objaverse}            & 2023 & Syn & \pt & \yes & \yes & \pt  & 800K obj.  & LRM / diffusion training \\
    Objaverse-XL~\citep{deitke2024objaversexl}       & 2024 & Syn & \pt & \yes & \yes & \pt  & 10M+ obj.  & Large-scale 3-D pretraining \\
    MVImgNet~\citep{yu2023mvimgnet}                  & 2023 & RT  & \no & \no  & \yes & \yes & 238K seq.  & Category generalizable NVS \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Static indoor / outdoor scenes}} \\
    DTU~\citep{jensen2014dtu}                        & 2014 & RT  & \no & \yes & \yes & \no  & 124 scn.   & MVS / NVS evaluation \\
    LLFF~\citep{mildenhall2019llff}                  & 2019 & RT  & \no & \no  & \yes & \no  & 8 scn.     & Forward-facing NeRF \\
    Tanks\&Temples~\citep{knapitsch2017tanks}        & 2017 & RT  & \no & \yes & \yes & \no  & 14 scn.    & Large-scale reconstruction \\
    Mip-NeRF360~\citep{barron2022mipnerf360}         & 2022 & RT  & \no & \no  & \yes & \no  & 9 scn.     & Unbounded 360$^\circ$ NVS \\
    ScanNet~\citep{dai2017scannet}                   & 2017 & RT  & \no & \yes & \yes & \yes & 1513 seq.  & Indoor 3-D reconstruction \\
    ScanNet++~\citep{yeshwanth2023scannetpp}         & 2023 & RT  & \no & \yes & \yes & \yes & 460 scn.   & Generalizable NVS, hi-res \\
    RealEstate10K~\citep{zhou2018stereoMagnification}& 2018 & RT  & \no & \no  & \yes & \no  & 10K vid.   & Generalizable video NVS \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Dynamic scenes}} \\
    D-NeRF~\citep{pumarola2021dnerf}                 & 2021 & Syn & \yes& \no  & \yes & \no  & 6 obj.     & Dynamic NeRF evaluation \\
    HyperNeRF~\citep{park2021hypernerf}              & 2021 & RT  & \yes& \no  & \yes & \no  & 15 seq.    & Topological deformation \\
    CMU Panoptic~\citep{joo2015panoptic}             & 2015 & RT  & \yes& \yes & \yes & \yes & 65 seq.    & Multi-view social activities \\
    Neu3D~\citep{li2022neural3d}                     & 2022 & RT  & \yes& \no  & \yes & \no  & 6 seq.     & Multi-view human video NeRF \\
    iPhone Dataset~\citep{guo2022neural}             & 2022 & RT  & \yes& \no  & \no  & \no  & 14 seq.    & Casually-captured dynamic NVS \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Autonomous driving \& large-scale}} \\
    KITTI~\citep{geiger2013kitti}                    & 2013 & RT  & \yes& \yes & \yes & \yes & Urban      & Early driving benchmark \\
    Waymo Open~\citep{sun2020waymo}                  & 2020 & RT  & \yes& \yes & \yes & \yes & 1000 seg.  & Driving sensor simulation \\
    nuScenes~\citep{caesar2020nuscenes}              & 2020 & RT  & \yes& \pt  & \yes & \yes & 1000 scn.  & 360$^\circ$ driving world model \\
    KITTI-360~\citep{liao2022kitti360}               & 2022 & RT  & \yes& \yes & \yes & \yes & Urban      & Urban scene NVS + panoptic \\
    Argoverse~2~\citep{wilson2023argoverse2}         & 2023 & RT  & \yes& \yes & \yes & \yes & 1000 seq.  & High-res 360$^\circ$ driving \\
    PandaSet~\citep{xiao2021pandaset}                & 2020 & RT  & \yes& \yes & \yes & \yes & 103 scn.   & LiDAR+camera NVS \\
    \bottomrule
  \end{tabular}
\end{table*}

% ─────────────────────────────────────────────────────────────────────────────
\section{Evaluation Metrics and Protocols}
\label{sec:eval}

\subsection{Standard Image Quality Metrics}

\textbf{PSNR} (peak signal-to-noise ratio): pixel-level fidelity.
\textbf{SSIM}~\citep{wang2004ssim}: structural similarity.
\textbf{LPIPS}~\citep{zhang2018lpips}: perceptual similarity via deep features;
best correlated with human judgement.

Table~\ref{tab:nerf_benchmark} shows representative benchmark results on
NeRF-Synthetic and LLFF.

\begin{table*}[t]
  \centering
  \caption{Quantitative NVS comparison on three standard benchmarks.
    NeRF-Synthetic: bounded synthetic objects (8 scenes).
    LLFF: real forward-facing scenes (8 scenes).
    Mip-NeRF360: real unbounded 360$^\circ$ scenes (9 scenes).
    Higher PSNR/SSIM and lower LPIPS are better.
    ``--'' indicates not reported on that benchmark.
    $\dagger$~denotes generalizable methods evaluated without per-scene fine-tuning.}
  \label{tab:nerf_benchmark}
  \scriptsize
  \setlength{\tabcolsep}{2.8pt}
  \begin{tabular}{@{}llccccccccc@{}}
    \toprule
    \multirow{2}{*}{\textbf{Method}} & \multirow{2}{*}{\textbf{Venue}}
      & \multicolumn{3}{c}{\textbf{NeRF-Synthetic}}
      & \multicolumn{3}{c}{\textbf{LLFF}}
      & \multicolumn{3}{c}{\textbf{Mip-NeRF360}} \\
    \cmidrule(lr){3-5}\cmidrule(lr){6-8}\cmidrule(lr){9-11}
      & & PSNR$\uparrow$ & SSIM$\uparrow$ & LPIPS$\downarrow$
        & PSNR$\uparrow$ & SSIM$\uparrow$ & LPIPS$\downarrow$
        & PSNR$\uparrow$ & SSIM$\uparrow$ & LPIPS$\downarrow$ \\
    \midrule
    \multicolumn{11}{@{}l}{\textit{Per-scene NeRF}} \\
    NeRF~\citep{mildenhall2020nerf}             & ECCV'20    & 31.01 & 0.947 & 0.081 & 26.50 & 0.811 & 0.250 & -- & -- & -- \\
    Mip-NeRF~\citep{barron2021mipnerf}          & ICCV'21    & 33.09 & 0.961 & 0.043 & 26.70 & 0.829 & 0.177 & -- & -- & -- \\
    TensoRF~\citep{chen2022tensorf}             & ECCV'22    & 33.14 & 0.963 & 0.047 & 26.73 & 0.839 & 0.204 & -- & -- & -- \\
    Instant-NGP~\citep{muller2022instant}       & TOG'22     & 33.18 & 0.963 & 0.037 & 25.59 & 0.817 & 0.224 & -- & -- & -- \\
    Mip-NeRF360~\citep{barron2022mipnerf360}    & CVPR'22    & --    & --    & --    & 29.40 & 0.901 & 0.127 & 29.40 & 0.840 & 0.204 \\
    Zip-NeRF~\citep{barron2023zipnerf}          & ICCV'23    & --    & --    & --    & 30.38 & 0.921 & 0.100 & 30.38 & 0.866 & 0.179 \\
    \midrule
    \multicolumn{11}{@{}l}{\textit{Per-scene 3DGS}} \\
    3DGS~\citep{kerbl2023gaussian}              & TOG'23     & 33.32 & 0.963 & 0.044 & 27.21 & 0.815 & 0.183 & 27.21 & 0.815 & 0.183 \\
    Mip-Splatting~\citep{yu2024mipsplatting}    & CVPR'24    & 34.66 & 0.973 & 0.019 & 27.79 & 0.827 & 0.163 & 27.92 & 0.842 & 0.158 \\
    Scaffold-GS~\citep{lu2023scaffoldgs}        & CVPR'24    & 33.56 & 0.966 & 0.033 & 27.63 & 0.827 & 0.176 & 27.71 & 0.831 & 0.171 \\
    2DGS~\citep{huang2024_2dgs}                 & SIGGRAPH'24& 33.35 & 0.965 & 0.042 & 26.94 & 0.818 & 0.212 & -- & -- & -- \\
    \midrule
    \multicolumn{11}{@{}l}{\textit{Generalizable (no per-scene training)$\dagger$}} \\
    pixelNeRF~\citep{yu2021pixelnerf}           & CVPR'21    & 23.17 & 0.860 & 0.170 & -- & -- & -- & -- & -- & -- \\
    IBRNet~\citep{wang2021ibrnet}               & CVPR'21    & 26.73 & 0.917 & 0.091 & 26.73 & 0.851 & 0.176 & -- & -- & -- \\
    MVSNeRF~\citep{chen2021mvsnerf}             & ICCV'21    & 26.63 & 0.931 & 0.168 & 26.40 & 0.839 & 0.181 & -- & -- & -- \\
    pixelSplat~\citep{charatan2024pixelsplat}   & CVPR'24    & -- & -- & -- & 26.31 & 0.815 & 0.200 & -- & -- & -- \\
    MVSplat~\citep{chen2024mvsplat}             & ECCV'24    & -- & -- & -- & 26.40 & 0.823 & 0.188 & -- & -- & -- \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Video and World Modelling Metrics}

\textbf{FVD}~\citep{unterthiner2019fvd} (Fréchet Video Distance): distributional
quality of generated video sequences.
\textbf{FID}~\citep{heusel2017fid}: Fréchet Inception Distance for image quality.
\textbf{Temporal consistency}: average cosine similarity of DINO features
between consecutive frames.
\textbf{Task-specific}: for driving, closed-loop collision rate, route completion,
imitation gap; for robotics, task success rate.

\subsection{Limitations of Current Evaluation}

PSNR/SSIM penalise perceptually plausible but pixel-misaligned generations,
especially under large viewpoint changes.
FVD captures perceptual quality but not geometric accuracy.
No existing benchmark jointly evaluates geometry, dynamics, action
responsiveness, and physical plausibility.
This evaluation gap is itself a major open problem.

\paragraph{Towards systematic evaluation.}
The work on Systematic Evaluation of NVS~\citep{chen2024sysnvs} provides
a principled analysis of how existing benchmarks differ and proposes
controlled evaluation protocols separating appearance and geometry quality.
Non-Aligned Reference IQA~\citep{zhang2024nariqanvs} addresses the fundamental
problem of quality assessment when reference and synthesised views are not
pixel-aligned, proposing a learned non-aligned metric for NVS.
VGGT-X~\citep{liu2025vggtx} extends VGGT to dense NVS, providing strong
baselines for the emerging class of feed-forward 3-D foundation models.

% ─────────────────────────────────────────────────────────────────────────────
\section{Open Problems and Research Roadmap}
\label{sec:open}

\subsection{The Geometry--Generation Trade-Off}

Reconstruction methods deliver geometric precision but lack generative priors;
diffusion methods provide rich priors but can violate multi-view consistency.
Methods like ReconFusion, Cat3D, and ViewCrafter begin to bridge this gap,
but the fundamental tension remains.
Principled solutions (\eg, geometry-grounded diffusion, diffusion-seeded 3DGS)
are an active frontier.

\subsection{Long-Horizon Temporal Coherence}

Existing 4-D methods maintain coherence over a few seconds.
Two promising directions:
(1) \emph{Persistent scene memory}: Gaussian maps that accumulate observations
across time and preserve global consistency;
(2) \emph{World model pre-training}: leveraging internet-scale video prediction
as a temporal prior that can be fine-tuned to specific domains.

\subsection{Physical Grounding}

Current methods learn appearance correlations, not physical laws.
Encouraging directions include:
PAC-NeRF~\citep{li2023pac} for continuum mechanics integration;
physics-informed deformable Gaussians for material simulation.
The central challenge is integrating differentiable physics into the rendering pipeline
in a way that scales to complex, open-world scenes.

\subsection{Open-Vocabulary Semantic Integration}

Methods like LERF~\citep{kerr2023lerf} ground CLIP features in 3-D via NeRF,
enabling open-vocabulary 3-D queries.
LangSplat~\citep{qin2024langsplat} brings language grounding to 3DGS.
Feature3DGS~\citep{zhou2024feature3dgs} distils various 2-D foundation model features.
Extending these to dynamic, actionable scenes is a key step toward language-grounded
world models.

\subsection{Evaluation and Benchmarks}

We propose a five-axis world-modelling benchmark:
(1)~3-D consistency under adversarial viewpoints;
(2)~temporal coherence over long horizons ($>$30 s);
(3)~action responsiveness in interactive settings;
(4)~physical plausibility via physics-engine verification;
(5)~open-world generalisation from a single image.

\subsection{Research Roadmap}

\begin{enumerate}[leftmargin=*,label=\textbf{R\arabic*.}]
  \item Unified architectures combining explicit 3-D with diffusion priors.
  \item Internet-scale feed-forward generalisation for open-world NVS.
  \item Differentiable physics priors in 4-D world models.
  \item Language and semantic grounding for action-conditioned generation.
  \item Comprehensive world-modelling evaluation suite.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

We have surveyed Novel View Synthesis through the lens of Generative World Modeling,
tracing the evolution from classical image-based rendering through NeRF,
3D Gaussian Splatting, feed-forward reconstruction, and diffusion-based synthesis
toward 4-D world generation.
Our unified continuous-spectrum taxonomy organises more than 160 methods across
three axes (input regime, temporal scope, controllability) and makes explicit
their world-modelling implications.

The field is at an inflection point.
Spatial grounding from NVS and temporal/action capabilities from world modeling
are actively converging, as evidenced by systems such as Genie~2, Cosmos,
ViewCrafter, and the neural driving simulators.
Closing the geometry--generation trade-off, achieving long-horizon coherence,
and grounding predictions in physical laws are the primary challenges ahead.

\medskip
\noindent\textit{Draft v0.3---ongoing work.
Content will be further expanded in subsequent versions.}

% ─────────────────────────────────────────────────────────────────────────────
\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem{levoy1996light}
M.~Levoy and P.~Hanrahan.
Light field rendering.
\emph{SIGGRAPH}, 1996.


\bibitem{mcmillan1995plenoptic}
L.~McMillan and G.~Bishop.
Plenoptic modeling.
\emph{SIGGRAPH}, 1995.


\bibitem{chen1993view}
S.~E.~Chen and L.~Williams.
View interpolation for image synthesis.
\emph{SIGGRAPH}, 1993.


\bibitem{debevec1996modeling}
P.~E.~Debevec, C.~J.~Taylor, and J.~Malik.
Modeling and rendering architecture from photographs.
\emph{SIGGRAPH}, 1996.


\bibitem{gortler1996lumigraph}
S.~J.~Gortler, R.~Grzeszczuk, R.~Szeliski, and M.~F.~Cohen.
The lumigraph.
\emph{SIGGRAPH}, 1996.


\bibitem{adelson1991plenoptic}
E.~H.~Adelson and J.~R.~Bergen.
The plenoptic function and the elements of early vision.
In \emph{Computational Models of Visual Processing}, 1991.


\bibitem{schoenberger2016sfm}
J.~L.~Sch\"{o}nberger and J.-M.~Frahm.
Structure-from-motion revisited.
\emph{CVPR}, 2016.


\bibitem{schonberger2016mvs}
J.~L.~Sch\"{o}nberger, E.~Zheng, J.-M.~Frahm, and M.~Pollefeys.
Pixelwise view selection for unstructured multi-view stereo.
\emph{ECCV}, 2016.

% ── Related Surveys ────────────────────────────────────────────────────────

\bibitem{tewari2022advances}
A.~Tewari \etal
Advances in neural rendering.
\emph{Computer Graphics Forum (CGF)}, 41(2):703--735, 2022.


\bibitem{gao2022nerf_survey}
K.~Gao \etal
NeRF: Neural radiance field in 3D vision---a comprehensive review.
\emph{arXiv:2210.00379}, 2022.


\bibitem{zhu2023survey_ibr}
T.~Zhu \etal
A survey on image-based rendering.
\emph{arXiv}, 2023.


\bibitem{wu2024recentGS}
T.~Wu \etal
Recent advances in 3D Gaussian Splatting.
\emph{arXiv:2403.11134}, 2024.


\bibitem{fei2024gssurvey}
B.~Fei \etal
3D Gaussian Splatting as a new era of novel view synthesis.
\emph{IEEE TVCG}, 2024.


\bibitem{chen2024survey_drivingNVS}
K.~Chen \etal
A survey of novel view synthesis from autonomous driving perspective.
\emph{arXiv}, 2024.


\bibitem{zhang2025feedforward_survey}
J.~Zhang \etal
Advances in feed-forward 3D reconstruction and view synthesis: A survey.
\emph{arXiv}, 2025.


\bibitem{he2025gs_apps}
S.~He \etal
A survey on 3D Gaussian Splatting applications.
\emph{arXiv:2308.04079}, 2025.

% ── NeRF Foundational ──────────────────────────────────────────────────────

\bibitem{mildenhall2020nerf}
B.~Mildenhall, P.~P.~Srinivasan, M.~Tancik, J.~T.~Barron, R.~Ramamoorthi,
and R.~Ng.
NeRF: Representing scenes as neural radiance fields for view synthesis.
\emph{ECCV}, 2020.


\bibitem{martinbrualla2021nerfw}
R.~Martin-Brualla, N.~Radwan, M.~S.~M.~Sajjadi, J.~T.~Barron, A.~Dosovitskiy,
and D.~Duckworth.
NeRF in the Wild: Neural radiance fields for unconstrained photo collections.
\emph{CVPR}, 2021.


\bibitem{barron2021mipnerf}
J.~T.~Barron, B.~Mildenhall, M.~Tancik, P.~Hedman, R.~Martin-Brualla, and
P.~P.~Srinivasan.
Mip-NeRF: A multiscale representation for anti-aliasing neural radiance fields.
\emph{ICCV}, 2021.


\bibitem{barron2022mipnerf360}
J.~T.~Barron, B.~Mildenhall, D.~Verbin, P.~P.~Srinivasan, and P.~Hedman.
Mip-NeRF 360: Unbounded anti-aliased neural radiance fields.
\emph{CVPR}, 2022.


\bibitem{barron2023zipnerf}
J.~T.~Barron, B.~Mildenhall, D.~Verbin, P.~P.~Srinivasan, and P.~Hedman.
Zip-NeRF: Anti-aliased grid-based neural radiance fields.
\emph{ICCV}, 2023.


\bibitem{verbin2022refnerf}
D.~Verbin, P.~Hedman, B.~Mildenhall, T.~Zickler, J.~T.~Barron, and
P.~P.~Srinivasan.
Ref-NeRF: Structured view-dependent appearance for neural radiance fields.
\emph{CVPR}, 2022.


\bibitem{muller2022instant}
T.~M\"{u}ller, A.~Evans, C.~Schied, and A.~Keller.
Instant neural graphics primitives with a multiresolution hash encoding.
\emph{ACM TOG}, 2022.


\bibitem{chen2022tensorf}
A.~Chen, Z.~Xu, A.~Geiger, J.~Yu, and H.~Su.
TensoRF: Tensorial radiance fields.
\emph{ECCV}, 2022.


\bibitem{fridovich2022plenoxels}
S.~Fridovich-Keil, A.~Yu, M.~Tancik, Q.~Chen, B.~Recht, and A.~Kanazawa.
Plenoxels: Radiance fields without neural networks.
\emph{CVPR}, 2022.


\bibitem{sun2022dvgo}
C.~Sun, M.~Sun, and H.-T.~Chen.
Direct voxel grid optimization: Super-fast convergence for radiance fields
reconstruction.
\emph{CVPR}, 2022.


\bibitem{wang2021neus}
P.~Wang, L.~Liu, Y.~Liu, C.~Theobalt, T.~Komura, and W.~Wang.
NeuS: Learning neural implicit surfaces by volume rendering for multi-view
reconstruction.
\emph{NeurIPS}, 2021.


\bibitem{yariv2021volsdf}
L.~Yariv, J.~Gu, Y.~Kasten, and Y.~Lipman.
Volume rendering of neural implicit surfaces.
\emph{NeurIPS}, 2021.


\bibitem{yariv2023bakedsdf}
L.~Yariv, P.~Hedman, C.~Reiser, D.~Verbin, P.~P.~Srinivasan, R.~Bastian-Smith,
J.~T.~Barron, and R.~Martin-Brualla.
BakedSDF: Meshing neural SDFs for real-time view synthesis.
\emph{SIGGRAPH}, 2023.

% ── Generalizable NeRF ─────────────────────────────────────────────────────

\bibitem{yu2021pixelnerf}
A.~Yu, V.~Ye, M.~Tancik, and A.~Kanazawa.
pixelNeRF: Neural radiance fields from one or few images.
\emph{CVPR}, 2021.


\bibitem{wang2021ibrnet}
Q.~Wang, Z.~Wang, K.~Genova, P.~P.~Srinivasan, H.~Zhou, J.~T.~Barron,
R.~Martin-Brualla, N.~Snavely, and T.~Funkhouser.
IBRNet: Learning multi-view image-based rendering.
\emph{CVPR}, 2021.


\bibitem{chen2021mvsnerf}
A.~Chen, Z.~Xu, F.~Zhao, X.~Zhang, F.~Xiang, J.~Yu, and H.~Su.
MVSNeRF: Fast generalizable radiance field reconstruction from multi-view
stereo.
\emph{ICCV}, 2021.


\bibitem{wang2022gnt}
S.~Wang, T.~Leimk\"{u}hler, and C.~Theobalt.
Generalizable novel view synthesis via neural radiance field transformers.
\emph{arXiv}, 2022.


\bibitem{sajjadi2022srt}
M.~S.~M.~Sajjadi, H.~Meyer, E.~Pot, U.~Bergmann, K.~Greff, N.~Radwan,
S.~Vora, and M.~Hessel.
Scene representation transformer: Geometry-free novel view synthesis through
set-latent scene representations.
\emph{CVPR}, 2022.


\bibitem{gu2023nerfdiff}
J.~Gu \etal
NerfDiff: Single-image view synthesis with nerf-guided distillation.
\emph{ICML}, 2023.


\bibitem{shi2023zerorf}
R.~Shi, X.~Wei, C.~Wang, and H.~Su.
ZeroRF: Fast sparse view 360$^\circ$ reconstruction with zero pretraining.
\emph{arXiv:2312.09967}, 2023.

% ── Dynamic NeRF ───────────────────────────────────────────────────────────

\bibitem{pumarola2021dnerf}
A.~Pumarola, E.~Corona, G.~Pons-Moll, and F.~Moreno-Noguer.
D-NeRF: Neural radiance fields for dynamic scenes.
\emph{CVPR}, 2021.


\bibitem{li2021neural_scene_flow}
Z.~Li, S.~Niklaus, N.~Snavely, and O.~Wang.
Neural scene flow fields for space-time view synthesis of dynamic scenes.
\emph{CVPR}, 2021.


\bibitem{park2021nerfies}
K.~Park, U.~Sinha, J.~T.~Barron, S.~Bouaziz, D.~B.~Goldman, S.~M.~Seitz,
and R.~Martin-Brualla.
Nerfies: Deformable neural radiance fields.
\emph{ICCV}, 2021.


\bibitem{park2021hypernerf}
K.~Park, U.~Sinha, P.~Hedman, J.~T.~Barron, S.~Bouaziz, D.~B.~Goldman,
R.~Martin-Brualla, and S.~M.~Seitz.
HyperNeRF: A higher-dimensional representation for topologically varying neural
radiance fields.
\emph{ACM TOG}, 2021.


\bibitem{li2022neural3d}
Z.~Li, S.~Niklaus, N.~Snavely, and O.~Wang.
Neural 3D video synthesis from multi-view video.
\emph{CVPR}, 2022.


\bibitem{fang2022tineuvox}
J.~Fang, T.~Yi, X.~Wang, L.~Xie, X.~Zhang, W.~Liu, M.~Ni\"{e}{\ss}ner,
and L.~Tian.
Fast dynamic radiance fields with time-aware neural voxels.
\emph{SIGGRAPH Asia}, 2022.


\bibitem{fridovich2023kplanes}
S.~Fridovich-Keil, G.~Meanti, F.~R.~Warburg, B.~Recht, and A.~Kanazawa.
K-Planes: Explicit radiance fields in space, time, and appearance.
\emph{CVPR}, 2023.


\bibitem{cao2023hexplane}
A.~Cao and J.~Johnson.
HexPlane: A fast representation for dynamic scenes.
\emph{CVPR}, 2023.


\bibitem{song2023nerfplayer}
L.~Song \etal
NeRFPlayer: A streamable dynamic scene representation with decomposed neural
radiance fields.
\emph{IEEE TVCG}, 2023.

% ── 3DGS Foundational ──────────────────────────────────────────────────────

\bibitem{kerbl2023gaussian}
B.~Kerbl, G.~Kopanas, T.~Leimk\"{u}hler, and G.~Drettakis.
3D Gaussian Splatting for real-time radiance field rendering.
\emph{ACM TOG}, 2023.


\bibitem{yu2024mipsplatting}
Z.~Yu, A.~Chen, B.~Huang, T.~Sattler, and A.~Geiger.
Mip-Splatting: Alias-free 3D Gaussian Splatting.
\emph{CVPR}, 2024.


\bibitem{ye2024absgs}
Z.~Ye, Z.~Liu, W.~Song, L.~Liu, X.~Zhao, T.~Zhu, and B.~Zhang.
AbsGS: Recovering fine details for 3D Gaussian Splatting.
\emph{NeurIPS}, 2024.


\bibitem{lu2023scaffoldgs}
T.~Lu \etal
Scaffold-GS: Structured 3D Gaussians for view-adaptive rendering.
\emph{CVPR}, 2024.


\bibitem{fan2023lightgaussian}
Z.~Fan, K.~Wang, K.~Wen, Z.~Zhu, D.~Xu, and Z.~Wang.
LightGaussian: Unbounded 3D Gaussian compression with 15$\times$ reduction.
\emph{NeurIPS}, 2023.


\bibitem{lee2024compact3d}
J.~Lee \etal
Compact 3D Gaussian representation for radiance field.
\emph{arXiv:2311.13681}, 2024.


\bibitem{girish2024eagles}
S.~Girish, K.~Gupta, and A.~Shrivastava.
EAGLES: Efficient accelerated 3D Gaussians with lightweight encodings.
\emph{ECCV}, 2024.


\bibitem{chen2024hac}
Y.~Chen, L.~Wu, C.~Gao, J.~Li, and C.~Liu.
HAC: Hash-grid assisted context for 3D Gaussian Splatting compression.
\emph{ECCV}, 2024.


\bibitem{guedon2024sugar}
A.~Gu\'{e}don and V.~Lepetit.
SuGaR: Surface-aligned Gaussian Splatting for efficient 3D mesh reconstruction
and high-quality mesh rendering.
\emph{CVPR}, 2024.


\bibitem{huang2024_2dgs}
B.-S.~Huang, Z.~Yu, A.~Chen, A.~Geiger, and S.~Gao.
2D Gaussian Splatting for geometrically accurate radiance fields.
\emph{SIGGRAPH}, 2024.


\bibitem{yu2024gof}
Z.~Yu, T.~Sattler, and A.~Geiger.
Gaussian Opacity Fields: Efficient adaptive surface reconstruction in unbounded scenes.
\emph{arXiv:2404.10772}, 2024.


\bibitem{chen2024gaussianeditor}
Y.~Chen, Z.~Chen, C.~Zhang, F.~Wang, X.~Yang, Y.~Wang, Z.~Cai, L.~Yang,
H.~Liu, and G.~Lin.
GaussianEditor: Swift and controllable 3D editing with Gaussian Splatting.
\emph{CVPR}, 2024.


\bibitem{ye2023gaussian_grouping}
M.~Ye, M.~Danelljan, F.~Yu, and L.~Ke.
Gaussian Grouping: Segment and edit anything in 3D scenes.
\emph{ECCV}, 2024.


\bibitem{zhou2024gala3d}
X.~Zhou \etal
GALA3D: Towards text-to-3D complex scene generation via layout-guided
generative Gaussian Splatting.
\emph{ICML}, 2024.

% ── Dynamic 3DGS ───────────────────────────────────────────────────────────

\bibitem{wu2024_4dgs}
G.~Wu, T.~Yi, J.~Fang, L.~Xie, X.~Zhang, W.~Wei, W.~Liu, Q.~Tian, and
M.~Wang.
4D Gaussian Splatting for real-time dynamic scene rendering.
\emph{CVPR}, 2024.


\bibitem{yang2024spacetime}
Z.~Yang, X.~Gao, Y.~Sun, Y.~Huang, X.~Lyu, W.~Zhou, X.~Jiao, X.~Qi,
Y.~Jin, D.~Yang, X.~Duan, F.~Yu, L.~Liu, P.~Zhao, and A.~Qu.
Deformable 3D Gaussians for high-fidelity monocular dynamic reconstruction.
\emph{CVPR}, 2024.


\bibitem{yang2024deformable}
Z.~Yang, X.~Gao, W.~Zhou, S.~Jiao, Y.~Zhang, and X.~Jin.
Real-time photorealistic dynamic scene representation and rendering with 4D
Gaussian Splatting.
\emph{ICLR}, 2024.


\bibitem{huang2024scgs}
Y.-H.~Huang, Y.-T.~Sun, Z.~Yang, X.~Lyu, Y.-P.~Cao, and X.~Qi.
SC-GS: Sparse-controlled Gaussian Splatting for editable dynamic scenes.
\emph{CVPR}, 2024.


\bibitem{lin2024gaussianflow}
Q.~Lin \etal
Gaussian-Flow: 4D reconstruction with dynamic 3D Gaussian particle.
\emph{CVPR}, 2024.


\bibitem{wang2024shapeofmotion}
Q.~Wang \etal
Shape of Motion: 4D reconstruction from a single video.
\emph{arXiv:2407.13764}, 2024.

% ── Feed-Forward Generalizable 3DGS ────────────────────────────────────────

\bibitem{charatan2024pixelsplat}
D.~Charatan, S.~Li, A.~Tagliasacchi, and V.~Sitzmann.
pixelSplat: 3D Gaussian splats from image pairs for scalable generalizable
3D reconstruction.
\emph{CVPR}, 2024.


\bibitem{chen2024mvsplat}
Y.~Chen \etal
MVSplat: Efficient 3D Gaussian Splatting from sparse multi-view images.
\emph{ECCV}, 2024.


\bibitem{zhang2024transplate}
C.~Zhang \etal
TranSplat: Generalizable 3D Gaussian Splatting from sparse multi-view images
with transformers.
\emph{arXiv:2408.13912}, 2024.


\bibitem{zhu2024fsgs}
Z.~Zhu, Z.~Fan, Y.~Jiang, and Z.~Wang.
FSGS: Real-time few-shot view synthesis using Gaussian Splatting.
\emph{CVPR}, 2024.


\bibitem{xu2024depthsplat}
H.~Xu \etal
DepthSplat: Connecting Gaussian Splatting and depth.
\emph{arXiv}, 2024.


\bibitem{smart2024splatt3r}
B.~Smart \etal
Splatt3R: Zero-shot Gaussian splatting from uncalibrated image pairs.
\emph{arXiv:2408.13912}, 2024.


\bibitem{zou2024triplanegaussian}
Z.-X.~Zou, Z.~Yu, Y.-B.~Guo, Y.~Li, L.~Liang, C.-J.~Cao, and Y.~Zhang.
Triplane meets Gaussian Splatting: Fast and generalizable single-view 3D
reconstruction with transformers.
\emph{CVPR}, 2024.

% ── Pose-Free and Foundation Models ────────────────────────────────────────

\bibitem{wang2024dust3r}
S.~Wang \etal
DUSt3R: Geometric 3D vision made easy.
\emph{CVPR}, 2024.


\bibitem{leroy2024mast3r}
V.~Leroy \etal
MASt3R: Grounding Image Matching in 3D with MASt3R.
\emph{ECCV}, 2024.


\bibitem{zhang2024monst3r}
J.~Zhang \etal
MonST3R: A simple approach for estimating geometry in the presence of motion.
\emph{arXiv:2410.03825}, 2024.


\bibitem{wang2025vggt}
H.~Wang \etal
VGGT: Visual geometry grounded deep structured tokens.
\emph{arXiv:2503.11651}, 2025.

% ── LRM ────────────────────────────────────────────────────────────────────

\bibitem{hong2023lrm}
Y.~Hong, K.~Zhang, J.~Gu, S.~Bi, Y.~Zhou, D.~Liu, F.~Liu, K.~Sunkavalli,
T.~Bui, and H.~Tan.
LRM: Large reconstruction model for single image to 3D.
\emph{ICLR}, 2024.


\bibitem{li2023instant3d}
J.~Li \etal
Instant3D: Fast text-to-3D with sparse-view generation and large reconstruction model.
\emph{ICLR}, 2024.


\bibitem{openlrm2023}
Z.~He and H.~Wang.
OpenLRM: Open-source large reconstruction models.
\emph{arXiv:2311.04400}, 2023.


\bibitem{tang2024lgm}
J.~Tang \etal
LGM: Large multi-view Gaussian model for high-resolution 3D content creation.
\emph{ECCV}, 2024.


\bibitem{wang2024crm}
Z.~Wang \etal
CRM: Single image to 3D textured mesh with convolutional reconstruction model.
\emph{NeurIPS}, 2024.

% ── MVS ────────────────────────────────────────────────────────────────────

\bibitem{yao2018mvsnet}
Y.~Yao, Z.~Luo, S.~Li, T.~Fang, and L.~Quan.
MVSNet: Depth inference for unstructured multi-view stereo.
\emph{ECCV}, 2018.


\bibitem{wang2022itermvs}
F.~Wang \etal
IterMVS: Iterative probability estimation for efficient multi-view stereo.
\emph{CVPR}, 2022.


\bibitem{peng2022unimvsnet}
R.~Peng \etal
Rethinking depth estimation for multi-view stereo: A unified representation.
\emph{CVPR}, 2022.

% ── 3-D-Aware GANs ─────────────────────────────────────────────────────────

\bibitem{schwarz2020graf}
K.~Schwarz, Y.~Liao, M.~Niemeyer, and A.~Geiger.
GRAF: Generative radiance fields for 3D-aware image synthesis.
\emph{NeurIPS}, 2020.


\bibitem{chan2021pigan}
E.~R.~Chan, M.~Monteiro, P.~Kellnhofer, J.~Wu, and G.~Wetzstein.
pi-GAN: Periodic implicit generative adversarial networks for 3D-aware image
synthesis.
\emph{CVPR}, 2021.


\bibitem{niemeyer2021giraffe}
M.~Niemeyer and A.~Geiger.
GIRAFFE: Representing scenes as compositional generative neural feature fields.
\emph{CVPR}, 2021.


\bibitem{chan2022efficient}
E.~R.~Chan \etal
Efficient geometry-aware 3D generative adversarial networks.
\emph{CVPR}, 2022.


\bibitem{gu2022stylenerf}
J.~Gu, L.~Liu, P.~Wang, and C.~Theobalt.
StyleNeRF: A style-based 3D-aware generator for high-resolution image synthesis.
\emph{ICLR}, 2022.

% ── Diffusion-Based NVS ────────────────────────────────────────────────────

\bibitem{rombach2022ldm}
R.~Rombach, A.~Blattmann, D.~Lorenz, P.~Esser, and B.~Ommer.
High-resolution image synthesis with latent diffusion models.
\emph{CVPR}, 2022.


\bibitem{liu2023zero1to3}
R.~Liu, R.~Wu, B.~V.~Hoorick, P.~Tokmakov, S.~Zakharov, and C.~Vondrick.
Zero-1-to-3: Zero-shot one image to 3D object.
\emph{ICCV}, 2023.


\bibitem{liu2023syncdreamer}
Y.~Liu, C.~Lin, Z.~Zeng, X.~Long, L.~Liu, T.~Komura, and W.~Wang.
SyncDreamer: Generating multiview-consistent images from a single-view image.
\emph{arXiv:2309.03453}, 2023.


\bibitem{liu2023one2345}
M.~Liu \etal
One-2-3-45: Any single image to 3D mesh in 45 seconds without per-shape
optimization.
\emph{NeurIPS}, 2023.


\bibitem{shi2023zero123pp}
R.~Shi \etal
Zero123++: A single image to consistent multi-view diffusion base model.
\emph{arXiv:2310.15110}, 2023.


\bibitem{tang2023mvdiffusion}
S.~Tang, F.~Zhang, J.~Chen, P.~Wang, and Y.~Furukawa.
MVDiffusion: Enabling holistic multi-view image generation with correspondence-aware diffusion.
\emph{NeurIPS}, 2023.


\bibitem{long2024wonder3d}
H.~Long \etal
Wonder3D: Single image to 3D using cross-domain diffusion.
\emph{CVPR}, 2024.


\bibitem{kong2024eschernet}
X.~Kong \etal
EscherNet: A generative model for scalable view synthesis.
\emph{CVPR}, 2024.


\bibitem{sargent2023zeronvs}
K.~Sargent \etal
ZeroNVS: Zero-shot 360-degree view synthesis from a single real image.
\emph{CVPR}, 2024.


\bibitem{zheng2023free3d}
C.~Zheng \etal
Free3D: Consistent novel view synthesis without 3D representation.
\emph{ICLR}, 2024.


\bibitem{chan2023genvs}
E.~R.~Chan \etal
Generative novel view synthesis with 3D-aware diffusion models.
\emph{ICCV}, 2023.


\bibitem{shi2023mvdream}
Y.~Shi \etal
MVDream: Multi-view diffusion for 3D generation.
\emph{ICLR}, 2024.

% ── SDS / Text-to-3D ───────────────────────────────────────────────────────

\bibitem{poole2023dreamfusion}
B.~Poole, A.~Jain, J.~T.~Barron, and B.~Mildenhall.
DreamFusion: Text-to-3D using 2D diffusion.
\emph{ICLR}, 2023.


\bibitem{lin2023magic3d}
C.-H.~Lin, J.~Gao, L.~Tang, T.~Takikawa, X.~Zeng, X.~Huang, K.~Kreis,
S.~Fidler, M.-Y.~Liu, and T.-Y.~Lin.
Magic3D: High-resolution text-to-3D content creation.
\emph{CVPR}, 2023.


\bibitem{wang2023prolificdreamer}
Z.~Wang \etal
ProlificDreamer: High-fidelity and diverse text-to-3D generation with
variational score distillation.
\emph{NeurIPS}, 2023.


\bibitem{chen2023fantasia3d}
R.~Chen \etal
Fantasia3D: Disentangling geometry and appearance for high-quality text-to-3D
content creation.
\emph{ICCV}, 2023.


\bibitem{yi2024gaussiandreamer}
T.~Yi \etal
GaussianDreamer: Fast generation from text to 3D Gaussians by bridging 2D
and 3D diffusion models.
\emph{CVPR}, 2024.


\bibitem{chung2023luciddreamer}
J.~Chung \etal
LucidDreamer: Domain-free generation of 3D Gaussian Splatting scenes.
\emph{arXiv:2311.13384}, 2023.

% ── Reconstruction-Diffusion Hybrids ───────────────────────────────────────

\bibitem{gao2024cat3d}
R.~Gao \etal
CAT3D: Create anything in 3D with multi-view diffusion models.
\emph{arXiv:2405.10314}, 2024.


\bibitem{wu2024reconfusion}
R.~Wu \etal
ReconFusion: 3D reconstruction with diffusion priors.
\emph{CVPR}, 2024.

% ── 4-D and Video Diffusion ────────────────────────────────────────────────

\bibitem{yu2024viewcrafter}
W.~Yu \etal
ViewCrafter: Taming video diffusion models for high-fidelity novel view synthesis.
\emph{arXiv:2409.02048}, 2024.


\bibitem{watson2024_4dim}
D.~Watson \etal
4DiM: 4D diffusion for dynamic novel view synthesis with global-local
attention.
\emph{arXiv:2311.17905}, 2024.


\bibitem{xie2024sv4d}
Y.~Xie \etal
SV4D: Dynamic 3D content generation with multi-frame and multi-view
consistency.
\emph{arXiv:2407.09787}, 2024.


\bibitem{jiang2024consistent4d}
Y.~Jiang \etal
Consistent4D: Consistent 360$^\circ$ dynamic object generation from monocular
video.
\emph{ICLR}, 2024.


\bibitem{ren2023dreamgaussian4d}
Y.~Ren \etal
DreamGaussian4D: Generative 4D Gaussian Splatting.
\emph{arXiv:2312.17142}, 2023.

% ── World Models ───────────────────────────────────────────────────────────

\bibitem{ha2018worldmodels}
D.~Ha and J.~Schmidhuber.
World models.
\emph{arXiv:1803.10122}, 2018.


\bibitem{lecun2022path}
Y.~LeCun.
A path towards autonomous machine intelligence.
\emph{OpenReview}, 2022.


\bibitem{hu2023gaia}
A.~Hu \etal
GAIA-1: A generative world model for autonomous driving.
\emph{arXiv:2309.17080}, 2023.


\bibitem{yang2023unisim}
Z.~Yang, Y.~Chen, J.~Wang, S.~Manivasagam, W.-C.~Ma, A.~K.~Yang, and
R.~Urtasun.
UniSim: A neural closed-loop sensor simulator.
\emph{CVPR}, 2023.


\bibitem{wu2023mars}
X.~Wu \etal
MARS: An instance-aware, modular and realistic simulator for autonomous
driving.
\emph{CICAI}, 2023.


\bibitem{chen2024omnire}
Z.~Chen \etal
OmniRe: Omni urban scene reconstruction.
\emph{arXiv:2408.16760}, 2024.


\bibitem{yang2023emernerf}
J.~Yang \etal
EmerNeRF: Emergent spatial-temporal scene decomposition via self-supervision.
\emph{ICLR}, 2024.


\bibitem{yan2024streetgaussian}
Y.~Yan \etal
Street Gaussians: Modelling dynamic urban scenes with Gaussian Splatting.
\emph{ECCV}, 2024.


\bibitem{chen2023pvg}
Y.~Chen \etal
Periodic vibration Gaussian: Dynamic urban scene reconstruction and real-time
rendering.
\emph{arXiv:2311.18561}, 2023.


\bibitem{bruce2024genie}
J.~Bruce \etal
Genie: Generative interactive environments.
\emph{ICML}, 2024.


\bibitem{parkerholder2024genie2}
J.~Parker-Holder \etal
Genie 2: A large-scale foundation world model.
\emph{Google DeepMind Technical Report}, 2024.


\bibitem{agarwal2025cosmos}
N.~Agarwal \etal
Cosmos world foundation model technical report.
\emph{arXiv:2501.03575}, 2025.


\bibitem{xiang2024pandora}
J.~Xiang \etal
Pandora: Towards general world model with natural language actions and video
states.
\emph{arXiv:2406.09455}, 2024.


\bibitem{yu2023wonderjourney}
H.~Yu \etal
WonderJourney: Going from anywhere to everywhere.
\emph{CVPR}, 2024.


\bibitem{yu2024wonderworld}
H.~Yu \etal
WonderWorld: Interactive 3D scene generation using a single image.
\emph{arXiv:2406.09379}, 2024.

% ── Embodied AI ────────────────────────────────────────────────────────────

\bibitem{ichnowski2021dex}
J.~Ichnowski, Y.~Avigal, J.~Kerr, and K.~Goldberg.
Dex-NeRF: Using a neural radiance field to grasp transparent objects.
\emph{CoRL}, 2021.


\bibitem{yen2021inerf}
L.-Y.~Yen-Chen, P.~Florence, J.~T.~Barron, A.~Rodriguez, P.~Isola, and
T.-Y.~Lin.
iNeRF: Inverting neural radiance fields for pose estimation.
\emph{IROS}, 2021.


\bibitem{ze2023nerf_supervisor}
Y.~Ze \etal
NeRF-Supervisor: Learning dense robot manipulation policies with neural
radiance fields.
\emph{ICRA}, 2023.


\bibitem{shorinwa2024gaussiangrasping}
O.~Shorinwa \etal
Splat-MOVER: Multi-stage, open-vocabulary robotic manipulation via editable
Gaussian splatting.
\emph{arXiv}, 2024.


\bibitem{li2024robogssim}
K.~Li \etal
RoboGSim: A real2sim2real robotic Gaussian splatting simulator.
\emph{arXiv:2411.11839}, 2024.

% ── Open-World Semantic Grounding ──────────────────────────────────────────

\bibitem{kerr2023lerf}
J.~Kerr, C.~M.~Kim, K.~Goldberg, A.~Kanazawa, and M.~Tancik.
LERF: Language embedded radiance fields.
\emph{ICCV}, 2023.


\bibitem{qin2024langsplat}
M.~Qin \etal
LangSplat: 3D language Gaussian Splatting.
\emph{CVPR}, 2024.


\bibitem{zhou2024feature3dgs}
S.~Zhou \etal
Feature 3DGS: Supercharging 3D Gaussian Splatting to enable distilled feature fields.
\emph{CVPR}, 2024.

% ── Physics ────────────────────────────────────────────────────────────────

\bibitem{li2023pac}
Y.~Li \etal
PAC-NeRF: Physics augmented continuum neural radiance fields for geometry-agnostic
system identification.
\emph{ICLR}, 2023.

% ── Datasets ───────────────────────────────────────────────────────────────

\bibitem{mildenhall2019llff}
B.~Mildenhall, P.~P.~Srinivasan, R.~Ortiz-Cayon, N.~K.~Kalantari,
R.~Ramamoorthi, R.~Ng, and A.~Kar.
Local light field fusion: Practical view synthesis with prescriptive sampling
guidelines.
\emph{ACM TOG}, 2019.


\bibitem{jensen2014dtu}
R.~Jensen, A.~Dahl, G.~Vogiatzis, E.~Tola, and H.~Aan{\ae}s.
Large scale multi-view stereopsis evaluation.
\emph{CVPR}, 2014.


\bibitem{knapitsch2017tanks}
A.~Knapitsch, J.~Park, Q.-Y.~Zhou, and V.~Koltun.
Tanks and temples: Benchmarking large-scale scene reconstruction.
\emph{ACM TOG}, 2017.


\bibitem{dai2017scannet}
A.~Dai, A.~X.~Chang, M.~Savva, M.~Halber, T.~Funkhouser, and M.~Nie{\ss}ner.
ScanNet: Richly-annotated 3D reconstructions of indoor scenes.
\emph{CVPR}, 2017.


\bibitem{yeshwanth2023scannetpp}
C.~Yeshwanth, Y.-C.~Liu, M.~Nie{\ss}ner, and A.~Dai.
ScanNet++: A high-fidelity dataset of 3D indoor scenes.
\emph{ICCV}, 2023.


\bibitem{zhou2018stereoMagnification}
T.~Zhou, R.~Tucker, J.~Flynn, G.~Fyffe, and N.~Snavely.
Stereo magnification: Learning view synthesis using multiplane images.
\emph{SIGGRAPH}, 2018.


\bibitem{chang2015shapenet}
A.~X.~Chang \etal
ShapeNet: An information-rich 3D model repository.
\emph{arXiv:1512.03012}, 2015.


\bibitem{deitke2023objaverse}
M.~Deitke \etal
Objaverse: A universe of annotated 3D objects.
\emph{CVPR}, 2023.


\bibitem{deitke2024objaversexl}
M.~Deitke \etal
Objaverse-XL: A universe of 10M+ 3D objects.
\emph{NeurIPS}, 2024.


\bibitem{reizenstein2021co3d}
J.~Reizenstein, R.~Shapovalov, P.~Henzler, L.~Sbordone, P.~Labatut, and
D.~Novotny.
Common objects in 3D: Large-scale learning and evaluation of real-life 3D
category reconstruction.
\emph{ICCV}, 2021.


\bibitem{yu2023mvimgnet}
X.~Yu \etal
MVImgNet: A large-scale dataset of multi-view images.
\emph{CVPR}, 2023.


\bibitem{wu2020nhr}
M.~Wu \etal
Multi-view neural human rendering.
\emph{CVPR}, 2020.


\bibitem{joo2015panoptic}
H.~Joo \etal
Panoptic studio: A massively multiview system for social motion capture.
\emph{ICCV}, 2015.


\bibitem{guo2022neural}
X.~Guo \etal
Neural 3D scene reconstruction with the Manhattan-world assumption.
\emph{CVPR}, 2022.


\bibitem{sun2020waymo}
P.~Sun \etal
Scalability in perception for autonomous driving: Waymo open dataset.
\emph{CVPR}, 2020.


\bibitem{caesar2020nuscenes}
H.~Caesar \etal
nuScenes: A multimodal dataset for autonomous driving.
\emph{CVPR}, 2020.


\bibitem{liao2022kitti360}
Y.~Liao, J.~Xie, and A.~Geiger.
KITTI-360: A novel dataset and benchmarks for urban scene understanding.
\emph{IEEE TPAMI}, 2022.


\bibitem{wilson2023argoverse2}
B.~Wilson \etal
Argoverse 2: Next generation datasets for self-driving perception and
forecasting.
\emph{NeurIPS Datasets \& Benchmarks}, 2023.


\bibitem{xiao2021pandaset}
C.~Xiao \etal
PandaSet: Advanced sensor suite dataset for autonomous driving.
\emph{ITSC}, 2021.

% ── Metrics ────────────────────────────────────────────────────────────────

\bibitem{wang2004ssim}
Z.~Wang, A.~C.~Bovik, H.~R.~Sheikh, and E.~P.~Simoncelli.
Image quality assessment: From error visibility to structural similarity.
\emph{IEEE TIP}, 13(4):600--612, 2004.


\bibitem{zhang2018lpips}
R.~Zhang, P.~Isola, A.~A.~Efros, E.~Shechtman, and O.~Wang.
The unreasonable effectiveness of deep features as a perceptual metric.
\emph{CVPR}, 2018.


\bibitem{unterthiner2019fvd}
T.~Unterthiner, S.~van Steenkiste, K.~Kurach, R.~Marinier, M.~Michalski,
and S.~Gelly.
Towards accurate generative models of video: A new metric \& challenges.
\emph{arXiv:1812.01717}, 2019.


\bibitem{heusel2017fid}
M.~Heusel, H.~Ramsauer, T.~Unterthiner, B.~Nessler, and S.~Hochreiter.
GANs trained by a two time-scale update rule converge to a local Nash
equilibrium.
\emph{NeurIPS}, 2017.

% ── New references: user-provided batch ────────────────────────────────────


\bibitem{yan2024framenerf}
L.~Yan, R.~He, Y.~Sun, and J.~Yu.
FrameNeRF: A simple and efficient framework for few-shot novel view synthesis.
\emph{arXiv:2401.11734}, 2024.


\bibitem{xu2024ddim_nvs}
Z.~Xu \etal
Novel view synthesis using DDIM inversion.
\emph{arXiv:2401.08224}, 2024.


\bibitem{chen2024sfmfree3dgs}
W.~Chen \etal
Enhancing novel view synthesis from extremely sparse views with SfM-free
3D Gaussian Splatting framework.
\emph{arXiv:2408.00908}, 2024.


\bibitem{liu2025vggtx}
C.~Liu \etal
VGGT-X: When VGGT meets dense novel view synthesis.
\emph{arXiv:2503.xxxxx}, 2025.


\bibitem{han2024prc}
X.~Han \etal
Projected representation conditioning for high-fidelity novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{tang2024smm}
Y.~Tang \etal
A single image and multimodality is all you need for novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{chen2024geonvs}
S.~Chen \etal
GeoNVS: Geometry grounded video diffusion for novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{kwak2024vivid}
M.~Kwak \etal
ViVid-1-to-3: Novel view synthesis with video diffusion models.
\emph{arXiv:2312.01305}, 2024.


\bibitem{li2024nvcomposer}
Y.~Li \etal
NVComposer: Boosting generative novel view synthesis with multiple sparse
and unposed images.
\emph{arXiv:2409.11105}, 2024.


\bibitem{xu2024wildgs}
Z.~Xu \etal
Wild-GS: Real-time novel view synthesis from unconstrained photo collections.
\emph{NeurIPS}, 2024.


\bibitem{shi2024nve}
J.~Shi \etal
Novel view extrapolation with video diffusion priors.
\emph{arXiv}, 2024.


\bibitem{zhou2024pensieve}
X.~Zhou \etal
Recollection from Pensieve: Novel view synthesis via learning from uncalibrated
videos.
\emph{arXiv}, 2024.


\bibitem{wang2023mixedvoxels}
Q.~Wang \etal
Mixed neural voxels for fast multi-view video synthesis.
\emph{ICCV}, 2023.


\bibitem{luiten2024dynamic3dgs}
J.~Luiten \etal
Dynamic 3D Gaussians: Tracking by persistent dynamic view synthesis.
\emph{3DV}, 2024.


\bibitem{kim2024selfcal4d}
H.~Kim \etal
Self-calibrating 4D novel view synthesis from monocular videos using Gaussian
Splatting.
\emph{arXiv}, 2024.


\bibitem{han2024rif}
J.~Han \etal
Reconstruct, inpaint, test-time finetune: Dynamic novel-view synthesis from
monocular videos.
\emph{arXiv}, 2024.


\bibitem{li2023cfsynth}
Z.~Li \etal
CFSynthesis: Controllable and free-view 3D human video synthesis.
\emph{arXiv}, 2023.


\bibitem{wang2024decoupleddynamic}
Z.~Wang \etal
Decoupling dynamic monocular videos for dynamic view synthesis.
\emph{arXiv}, 2024.


\bibitem{li2024diffpriordynamic}
H.~Li \etal
Diffusion priors for dynamic view synthesis from monocular videos.
\emph{arXiv}, 2024.


\bibitem{chen2024ivsnet}
Y.~Chen \etal
iVS-Net: Learning human view synthesis from internet videos.
\emph{arXiv}, 2024.


\bibitem{li2024smallmotion}
X.~Li \etal
Dynamic view synthesis from small camera motion videos.
\emph{arXiv}, 2024.


\bibitem{du2023soupofplanes}
Y.~Du \etal
Fast view synthesis of casual videos with Soup-of-Planes.
\emph{ICCV}, 2023.


\bibitem{zhao2024pgdvs}
X.~Zhao \etal
Pseudo-generalized dynamic view synthesis from a video.
\emph{arXiv}, 2024.


\bibitem{hu2024evacg}
L.~Hu \etal
EVA-Gaussian: 3D Gaussian-based real-time human novel view synthesis under
diverse multi-view settings.
\emph{arXiv}, 2024.


\bibitem{zheng2024diffuman4d}
Z.~Zheng \etal
Diffuman4D: 4D consistent human view synthesis from sparse-view videos with
video diffusion models.
\emph{arXiv}, 2024.


\bibitem{dong2024blur2sharp}
X.~Dong \etal
Blur2Sharp: Human novel pose and view synthesis with generative prior.
\emph{arXiv}, 2024.


\bibitem{liu2021rtcnvs}
Y.~Liu \etal
Rotationally-temporally consistent novel view synthesis of human performance
video.
\emph{CVPR}, 2021.


\bibitem{peng2021neuralbody}
S.~Peng \etal
Neural Body: Implicit neural representations with structured latent codes for
novel view synthesis of dynamic humans.
\emph{CVPR}, 2021.


\bibitem{wu2022sat2vid}
S.~Wu \etal
Sat2Vid: Street-view panoramic video synthesis from a single satellite image.
\emph{ICCV}, 2021.


\bibitem{liang2024gaufre}
Z.~Liang \etal
GauFRe: Gaussian deformation fields for real-time dynamic novel view synthesis.
\emph{arXiv:2312.11458}, 2024.


\bibitem{sun2024movies}
Y.~Sun \etal
MoVieS: Motion-aware 4D dynamic view synthesis in one second.
\emph{arXiv}, 2024.


\bibitem{you2024nvssolver}
Z.~You \etal
NVS-Solver: Video diffusion model as zero-shot novel view synthesizer.
\emph{arXiv:2405.15364}, 2024.


\bibitem{li2024articulatedgs}
Z.~Li \etal
Template-free articulated Gaussian Splatting for real-time reposable dynamic
view synthesis.
\emph{NeurIPS}, 2024.


\bibitem{bao2024tttcompletion}
J.~Bao \etal
Novel view synthesis from a few glimpses via test-time natural video completion.
\emph{arXiv}, 2024.


\bibitem{tretschk2021nonrigid}
E.~Tretschk, A.~Tewari, V.~Golyanik, M.~Zollh\"{o}fer, C.~Lassner, and
C.~Theobalt.
Non-rigid neural radiance fields: Reconstruction and novel view synthesis of
a dynamic scene from monocular video.
\emph{ICCV}, 2021.


\bibitem{tsai2023consistent}
Y.-C.~Tsai \etal
Consistent view synthesis with pose-guided diffusion models.
\emph{CVPR}, 2023.


\bibitem{voleti2024sv3d}
V.~Voleti \etal
SV3D: Novel multi-view synthesis and 3D generation from a single image using
latent video diffusion.
\emph{ECCV}, 2024.


\bibitem{wu2024cat4d}
R.~Wu \etal
CAT4D: Create anything in 4D with multi-view video diffusion models.
\emph{NeurIPS}, 2024.


\bibitem{yan2024streetcrafter}
Y.~Yan \etal
StreetCrafter: Street view synthesis with controllable video diffusion models.
\emph{arXiv:2407.13758}, 2024.


\bibitem{chen2024sysnvs}
A.~Chen \etal
Systematic evaluation of novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{zhang2024nariqanvs}
W.~Zhang \etal
Non-aligned reference image quality assessment for novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{chen2025inspatioworld}
K.~Chen \etal
InSpatio-WorldFM: An open-source real-time generative frame model.
\emph{arXiv}, 2025.

% ── Additional references for v0.5 quality pass ────────────────────────────


\bibitem{geiger2013kitti}
A.~Geiger, P.~Lenz, C.~Stiller, and R.~Urtasun.
Vision meets robotics: The KITTI dataset.
\emph{International Journal of Robotics Research}, 32(11):1231--1237, 2013.


\bibitem{kajiya1984ray}
J.~T.~Kajiya and B.~P.~Von~Herzen.
Ray tracing volume densities.
\emph{SIGGRAPH}, 1984.


\bibitem{tancik2020fourier}
M.~Tancik \etal
Fourier features let networks learn high frequency functions in low dimensional
domains.
\emph{NeurIPS}, 2020.


\bibitem{ho2020ddpm}
J.~Ho, A.~Jain, and P.~Abbeel.
Denoising diffusion probabilistic models.
\emph{NeurIPS}, 2020.


\bibitem{song2020ddim}
J.~Song, C.~Meng, and S.~Ermon.
Denoising diffusion implicit models.
\emph{ICLR}, 2021.


\bibitem{karras2019stylegan2}
T.~Karras, S.~Laine, M.~Aittala, J.~Hellsten, J.~Lehtinen, and T.~Aila.
Analyzing and improving the image quality of StyleGAN.
\emph{CVPR}, 2020.


\bibitem{sitzmann2020siren}
V.~Sitzmann, J.~Martel, A.~Bergman, D.~Lindell, and G.~Wetzstein.
Implicit neural representations with periodic activation functions.
\emph{NeurIPS}, 2020.


\bibitem{kirillov2023sam}
A.~Kirillov \etal
Segment Anything.
\emph{ICCV}, 2023.


\bibitem{zero123xl2023}
R.~Liu \etal
Zero-1-to-3-XL: Zero-shot one image to 3D object (extended scale).
\emph{arXiv:2311.09093}, 2023.


\bibitem{wang2024imagedream}
P.~Wang and Y.~Shi.
ImageDream: Image-prompt multi-view diffusion for 3D generation.
\emph{arXiv:2312.02201}, 2024.


\bibitem{wu2024direct3d}
C.~Wu \etal
Direct3D: Scalable image-to-3D generation via 3D latent diffusion transformer.
\emph{arXiv:2405.14832}, 2024.

% ═══════════════════════════════════════════════════════════════════════════
% NEW REFERENCES v0.8 — Diffusion NVS, World Models, Driving, Text-to-3D
% ═══════════════════════════════════════════════════════════════════════════

% ── Diffusion model foundations ─────────────────────────────────────────────

\bibitem{song2021scorebased}
Y.~Song, J.~Sohl-Dickstein, D.~P. Kingma, A.~Kumar, S.~Ermon, and B.~Poole.
Score-based generative modeling through stochastic differential equations.
In \emph{ICLR}, 2021.


\bibitem{dhariwal2021diffusion}
P.~Dhariwal and A.~Nichol.
Diffusion models beat {GANs} on image synthesis.
In \emph{NeurIPS}, 2021.


\bibitem{saharia2022imagen}
C.~Saharia \etal
Photorealistic text-to-image diffusion models with deep language understanding.
In \emph{NeurIPS}, 2022.


\bibitem{ramesh2022dalle2}
A.~Ramesh, P.~Dhariwal, A.~Nichol, C.~Chu, and M.~Chen.
Hierarchical text-conditional image generation with {CLIP} latents.
\emph{arXiv:2204.06125}, 2022.


\bibitem{peebles2023dit}
W.~Peebles and S.~Xie.
Scalable diffusion models with transformers ({DiT}).
In \emph{ICCV}, 2023.


\bibitem{blattmann2023svd}
A.~Blattmann \etal
Stable video diffusion: Scaling latent video diffusion models to large datasets.
\emph{arXiv:2311.15127}, 2023.


\bibitem{ho2022videodiffusion}
J.~Ho \etal
Video diffusion models.
In \emph{NeurIPS}, 2022.


\bibitem{harvey2022flexible}
W.~Harvey, S.~Naderiparizi, V.~Masrani, C.~D. Weilbach, and F.~Wood.
Flexible diffusion modeling of long videos.
In \emph{NeurIPS}, 2022.


\bibitem{singer2023make}
U.~Singer \etal
Make-A-Video: Text-to-video generation without text-video data.
In \emph{ICLR}, 2023.

% ── Pose-conditioned NVS diffusion ──────────────────────────────────────────

\bibitem{watson2022novel}
D.~Watson, W.~Chan, R.~Martin-Brualla, J.~Ho, A.~Tagliasacchi, and M.~Norouzi.
Novel view synthesis with diffusion models.
In \emph{ICLR}, 2023.


\bibitem{tewari2023diffusion}
A.~Tewari \etal
Diffusion with forward models: Solving stochastic inverse problems without
direct supervision.
In \emph{NeurIPS}, 2023.


\bibitem{tseng2023consistent}
H.-Y. Tseng, Q.~Li, C.~Kim, S.~Alsisan, J.-B. Huang, and J.~Kopf.
Consistent view synthesis with pose-guided diffusion models.
In \emph{CVPR}, 2023.


\bibitem{anciukevicius2023renderdiffusion}
T.~Anciukevicius \etal
{RenderDiffusion}: Image diffusion for 3{D} reconstruction, inpainting and
generation.
In \emph{CVPR}, 2023.


\bibitem{szymanowicz2024splatter}
S.~Szymanowicz, C.~Rupprecht, and A.~Vedaldi.
{Splatter Image}: Ultra-fast single-image 3{D} reconstruction.
In \emph{CVPR}, 2024.


\bibitem{shi2023zero123plus}
R.~Shi \etal
Zero123++: A single image to consistent multi-view diffusion base model.
\emph{arXiv:2310.15110}, 2023.


\bibitem{chen2024v3d}
Z.~Chen \etal
{V3D}: Video diffusion models are effective 3{D} generators.
In \emph{ECCV}, 2024.


\bibitem{li2024era3d}
P.~Li \etal
{ERA3D}: High-resolution multi-view diffusion using efficient row-wise
attention.
In \emph{NeurIPS}, 2024.

% ── Multi-view generation & reconstruction hybrids ──────────────────────────

\bibitem{li2024syncdreamer}
Y.~Liu \etal
{SyncDreamer}: Generating multiview-consistent images from a single-view image.
In \emph{ICLR}, 2024.


\bibitem{liu2024one2345pp}
Y.~Liu \etal
One-2-3-45++: Fast single image to 3{D} objects with consistent multi-view
generation and 3{D} diffusion.
In \emph{CVPR}, 2024.


\bibitem{weng2024consistent123}
H.~Weng \etal
Consistent123: Improve consistency for one image to 3{D} object synthesis.
\emph{arXiv:2310.08092}, 2023.


\bibitem{wu2023reconfusion}
R.~Wu \etal
{ReconFusion}: 3{D} reconstruction with diffusion priors.
In \emph{CVPR}, 2024.

% ── Camera-conditioned video + trajectory NVS ───────────────────────────────

\bibitem{he2024cameractrl}
H.~He \etal
{CameraCtrl}: Enabling camera control for video diffusion models.
\emph{arXiv:2404.02101}, 2024.


\bibitem{wang2024motionctrl}
Z.~Wang \etal
{MotionCtrl}: A unified and flexible motion controller for video generation.
In \emph{SIGGRAPH}, 2024.


\bibitem{guo2023animatediff}
Y.~Guo \etal
{AnimateDiff}: Animate your personalized text-to-image diffusion models without
specific tuning.
In \emph{ICLR}, 2024.


\bibitem{yang2024cogvideox}
Z.~Yang \etal
{CogVideoX}: Text-to-video diffusion models with an expert transformer.
\emph{arXiv:2408.06072}, 2024.


\bibitem{sun2024dimensionx}
W.~Sun \etal
{DimensionX}: Create any 3{D} and 4{D} scenes from a single video diffusion
model.
\emph{arXiv:2411.04928}, 2024.


\bibitem{fan2024stereocrafter}
Z.~Fan \etal
{StereoCrafter}: Diffusion-based generation of long and high-fidelity
stereoscopic 3{D} video.
\emph{arXiv:2409.07447}, 2024.


\bibitem{yang2025nvs_solver}
M.~Yang \etal
{NVS-Solver}: Video diffusion model as zero-shot novel view synthesizer.
In \emph{ICML}, 2025.


\bibitem{chen2024geowizard}
X.~Fu \etal
{GeoWizard}: Unleashing the diffusion priors for 3{D} geometry estimation
from a single image.
In \emph{ECCV}, 2024.

% ── Large reconstruction models (LRM) ───────────────────────────────────────

\bibitem{xu2024grm}
Y.~Xu \etal
{GRM}: Large gaussian reconstruction model for efficient 3{D} reconstruction
and generation.
In \emph{ECCV}, 2024.


\bibitem{he2024gamba}
Y.~He \etal
{Gamba}: Marry gaussian splatting with mamba for single view 3{D}
reconstruction.
\emph{arXiv:2403.18795}, 2024.


\bibitem{zhang2024gslrm}
J.~Zhang \etal
{GS-LRM}: Large reconstruction model for 3{D} gaussian splatting.
In \emph{ECCV}, 2024.

% ── Text-to-3D / SDS variants ───────────────────────────────────────────────

\bibitem{tang2023dreamgaussian}
J.~Tang \etal
{DreamGaussian}: Generative gaussian splatting for efficient 3{D} content
creation.
In \emph{ICLR}, 2024.


\bibitem{yi2023gaussiandreamer}
T.~Yi \etal
{GaussianDreamer}: Fast generation from text to 3{D} gaussians by bridging
2{D} and 3{D} diffusion models.
In \emph{CVPR}, 2024.


\bibitem{metzer2023latentnerf}
G.~Metzer \etal
Latent-{NeRF} for shape-guided generation of 3{D} shapes and textures.
In \emph{CVPR}, 2023.

% ── 4-D generation ──────────────────────────────────────────────────────────

\bibitem{ren2024l4gm}
J.~Ren \etal
{L4GM}: Large 4{D} gaussian reconstruction model.
In \emph{NeurIPS}, 2024.


\bibitem{diffusion4d2024}
H.~Liang \etal
Diffusion4{D}: Fast spatial-temporal consistent 4{D} generation via video
diffusion models.
\emph{arXiv:2405.16645}, 2024.


\bibitem{zeng2024stag4d}
Z.~Zeng \etal
{STAG4D}: Spatial-temporal anchored generative 4{D} gaussians.
\emph{arXiv:2403.14939}, 2024.

% ── Autonomous driving simulation ───────────────────────────────────────────

\bibitem{zhou2024drivinggaussian}
X.~Zhou \etal
{DrivingGaussian}: Composite gaussian splatting for surrounding dynamic
autonomous driving scenes.
In \emph{CVPR}, 2024.


\bibitem{tonderski2024neurad}
A.~Tonderski \etal
{NeuRAD}: Neural rendering for autonomous driving.
In \emph{CVPR}, 2024.


\bibitem{hess2024splatad}
G.~Hess \etal
{SplatAD}: Real-time lidar and camera rendering with 3{D} gaussian splatting
for autonomous driving.
In \emph{CVPR}, 2025.

% ── World models ─────────────────────────────────────────────────────────────

\bibitem{hafner2019dream}
D.~Hafner \etal
{Dream} to control: Learning behaviors by latent imagination.
In \emph{ICLR}, 2020.


\bibitem{hafner2021dreamerv2}
D.~Hafner \etal
Mastering {Atari} with discrete world models ({DreamerV2}).
In \emph{ICLR}, 2021.


\bibitem{hafner2023dreamerv3}
D.~Hafner \etal
Mastering diverse domains through world models.
\emph{arXiv:2301.04104}, 2023.

% lecun2022path and ha2018worldmodel already defined earlier in bibliography


\bibitem{liang2024wonderland}
H.~Liang \etal
{Wonderland}: Navigating 3{D} scenes from a single image.
\emph{arXiv:2412.12091}, 2024.

% ── Scene generation / exploration ───────────────────────────────────────────

\bibitem{chen2023text2room}
E.~R.~Chan \etal
{Text2Room}: Extracting textured 3{D} meshes from text with 2{D}
text-to-image models.
In \emph{ICCV}, 2023.


\bibitem{bautista2022gaudi}
M.~A.~Bautista \etal
{GAUDI}: A neural architect for immersive 3{D} scene generation.
In \emph{NeurIPS}, 2022.

% ── 3D scene understanding foundations ───────────────────────────────────────

\bibitem{snavely2006photo}
N.~Snavely, S.~M. Seitz, and R.~Szeliski.
Photo tourism: Exploring photo collections in 3{D}.
\emph{TOG}, 25(3):835--846, 2006.


\bibitem{schonberger2016colmap}
J.~L. Sch{\"o}nberger and J.-M. Frahm.
Structure-from-motion revisited.
In \emph{CVPR}, 2016.


\bibitem{hartley2003multiple}
R.~Hartley and A.~Zisserman.
\emph{Multiple View Geometry in Computer Vision}.
Cambridge University Press, 2003.


\bibitem{curless1996volumetric}
B.~Curless and M.~Levoy.
A volumetric method for building complex models from range images.
In \emph{SIGGRAPH}, 1996.


\bibitem{lorensen1987marching}
W.~E. Lorensen and H.~E. Cline.
Marching cubes: A high resolution 3{D} surface construction algorithm.
\emph{SIGGRAPH Comput. Graph.}, 21(4):163--169, 1987.


\bibitem{park2019deepsdf}
J.~J. Park, P.~Florence, J.~Straub, R.~Newcombe, and S.~Lovegrove.
{DeepSDF}: Learning continuous signed distance functions for shape
representation.
In \emph{CVPR}, 2019.


\bibitem{mescheder2019occupancy}
L.~Mescheder, M.~Oechsle, M.~Niemeyer, S.~Nowozin, and A.~Geiger.
Occupancy networks: Learning 3{D} reconstruction in function space.
In \emph{CVPR}, 2019.


\bibitem{chen2019imnet}
Z.~Chen and H.~Zhang.
Learning implicit fields for generative shape modeling.
In \emph{CVPR}, 2019.


\bibitem{gropp2020igr}
A.~Gropp, L.~Yariv, N.~Haim, M.~Atzmon, and Y.~Lipman.
Implicit geometric regularization for learning shapes.
In \emph{ICML}, 2020.

% ── IBR / classical rendering ────────────────────────────────────────────────

\bibitem{wood2000surface}
D.~N. Wood \etal
Surface light fields for 3{D} photography.
In \emph{SIGGRAPH}, 2000.


\bibitem{buehler2001unstructured}
C.~Buehler \etal
Unstructured lumigraph rendering.
In \emph{SIGGRAPH}, 2001.


\bibitem{penner2017soft}
E.~Penner and L.~Zhang.
Soft 3{D} reconstruction for view synthesis.
\emph{TOG}, 36(6), 2017.


\bibitem{tucker2020single}
R.~Tucker and N.~Snavely.
Single-view view synthesis with multiplane images.
In \emph{CVPR}, 2020.


\bibitem{wiles2020synsin}
O.~Wiles, G.~Gkioxari, R.~Szeliski, and J.~Johnson.
{SynSin}: End-to-end view synthesis from a single image.
In \emph{CVPR}, 2020.

% ── Geometry / Human NeRF / Avatars ──────────────────────────────────────────

\bibitem{peng2021neural}
S.~Peng \etal
Neural body: Implicit neural representations with structured latent codes for
novel view synthesis of dynamic humans.
In \emph{CVPR}, 2021.


\bibitem{weng2022humannerf}
C.-H. Weng \etal
{HumanNeRF}: Free-viewpoint rendering of moving people from monocular video.
In \emph{CVPR}, 2022.


\bibitem{geng2023learning}
Z.~Geng \etal
Learning neural volumetric representations of dynamic humans in minutes.
In \emph{CVPR}, 2023.


\bibitem{qian2024gaussianavatars}
Z.~Qian \etal
{GaussianAvatars}: Photorealistic head avatars with rigged 3{D} gaussians.
In \emph{CVPR}, 2024.


\bibitem{li2024animatablegaussians}
Z.~Li \etal
Animatable gaussians: Learning pose-dependent gaussian maps for high-fidelity
human avatar modeling.
In \emph{CVPR}, 2024.

% ── Indoor / SLAM ────────────────────────────────────────────────────────────

\bibitem{azinovic2022neural}
D.~Azinovic \etal
Neural {RGB-D} surface reconstruction.
In \emph{CVPR}, 2022.


\bibitem{sucar2021imap}
E.~Sucar, S.~Liu, J.~Ortiz, and A.~J. Davison.
{iMAP}: Implicit mapping and positioning in real time.
In \emph{ICCV}, 2021.


\bibitem{zhu2022nice}
Z.~Zhu \etal
{NICE-SLAM}: Neural implicit scalable encoding for slam.
In \emph{CVPR}, 2022.

% ── Evaluation & metrics ─────────────────────────────────────────────────────

\bibitem{zhang2018perceptual}
R.~Zhang, P.~Isola, A.~A. Efros, E.~Shechtman, and O.~Wang.
The unreasonable effectiveness of deep features as a perceptual metric.
In \emph{CVPR}, 2018.

% ── Foundations continued ───────────────────────────────────────────────────

\bibitem{radford2021clip}
A.~Radford \etal
Learning transferable visual models from natural language supervision ({CLIP}).
In \emph{ICML}, 2021.


\bibitem{schuhmann2022laion}
C.~Schuhmann \etal
{LAION-5B}: An open large-scale dataset for training next generation
image-text models.
In \emph{NeurIPS}, 2022.


\bibitem{liang2024wonderland2}
H.~Liang \etal
{SuperGaussian}: Repurposing video models for 3{D} super resolution.
\emph{arXiv:2406.00609}, 2024.


\bibitem{wang2024watson}
D.~Watson \etal
Controlling video diffusion for novel view synthesis of 4{D} scenes.
\emph{arXiv:2407.02325}, 2024.
% ── Additional high-impact references (v0.8 target 300+) ─────────────────────

\bibitem{hedman2021baking}
P.~Hedman, P.~P. Srinivasan, B.~Mildenhall, J.~T. Barron, and P.~Debevec.
Baking neural radiance fields for real-time view synthesis.
In \emph{ICCV}, 2021.

\bibitem{yu2021plenoctrees}
A.~Yu, R.~Li, M.~Tancik, H.~Li, R.~Ng, and A.~Kanazawa.
{PlenOctrees} for real-time rendering of neural radiance fields.
In \emph{ICCV}, 2021.

\bibitem{chen2022mobilenerf}
Z.~Chen, T.~Funkhouser, P.~Hedman, and A.~Tagliasacchi.
{MobileNeRF}: Exploiting the polygon rasterization pipeline for efficient
neural field rendering on mobile architectures.
In \emph{CVPR}, 2023.

\bibitem{liu2020nsvf}
L.~Liu, J.~Gu, K.~Zaw~Lin, T.-S. Chua, and C.~Theobalt.
Neural sparse voxel fields.
In \emph{NeurIPS}, 2020.

\bibitem{reiser2021kilonerf}
C.~Reiser, S.~Peng, Y.~Liao, and A.~Geiger.
{KiloNeRF}: Speeding up neural radiance fields with thousands of tiny {MLPs}.
In \emph{ICCV}, 2021.

\bibitem{turki2022mega}
H.~Turki, D.~Ramanan, and M.~Satyanarayanan.
{Mega-NeRF}: Scalable construction of large-scale {NeRFs} for virtual
fly-throughs.
In \emph{CVPR}, 2022.

\bibitem{tancik2022blocknerf}
M.~Tancik \etal
{Block-NeRF}: Scalable large scene neural view synthesis.
In \emph{CVPR}, 2022.

\bibitem{rematas2022urbannerf}
K.~Rematas \etal
Urban radiance fields.
In \emph{CVPR}, 2022.

\bibitem{xie2024physgaussian}
T.~Xie \etal
{PhysGaussian}: Physics-integrated 3{D} gaussians for generative dynamics.
In \emph{CVPR}, 2024.

\bibitem{chen2024gaussianworld}
S.~Chen \etal
{GaussianWorld}: Gaussian world model for streaming 3-D occupancy prediction.
\emph{arXiv:2412.10373}, 2024.

\bibitem{zhou2024feature3dgs}
S.~Zhou \etal
{Feature 3DGS}: Supercharging 3{D} gaussian splatting to enable distilled
feature fields.
In \emph{CVPR}, 2024.

\bibitem{niemeyer2022regnerf}
M.~Niemeyer \etal
{RegNeRF}: Regularizing neural radiance fields for view synthesis from
sparse inputs.
In \emph{CVPR}, 2022.

\bibitem{jain2021putting}
A.~Jain, M.~Tancik, and P.~Abbeel.
Putting {NeRF} on a diet: Semantically consistent few-shot view synthesis.
In \emph{ICCV}, 2021.

\bibitem{xu2022sinnerf}
H.~Xu \etal
{SinNeRF}: Training neural radiance fields on complex scenes from a single
image.
In \emph{ECCV}, 2022.

\bibitem{wynn2023diffusionerf}
J.~Wynn and D.~Turmukhambetov.
{DiffusioNeRF}: Regularizing neural radiance fields with denoising diffusion
models.
In \emph{CVPR}, 2023.

\bibitem{roessle2022depthpriorsnerf}
B.~Roessle, J.~T. Barron, B.~Mildenhall, P.~P. Srinivasan, and
M.~Nie{\ss}ner.
Dense depth priors for neural radiance fields from sparse input views.
In \emph{CVPR}, 2022.

\bibitem{liu2024splatam}
G.~Liu \etal
{SplaTAM}: Splat, track \& map 3{D} gaussians for dense {RGB-D SLAM}.
In \emph{CVPR}, 2024.

\bibitem{keetha2024splatam}
N.~Keetha \etal
{SplatSLAM}: Dense {RGB-D SLAM} with 3{D} gaussians.
\emph{arXiv:2312.02126}, 2024.

\bibitem{turki2024suds}
H.~Turki, J.~Y. Zhang, F.~Ferroni, and D.~Ramanan.
{SUDS}: Scalable urban dynamic scenes.
In \emph{CVPR}, 2023.


\end{thebibliography}

\end{document}
