\section{3D Gaussian Splatting}
\label{sec:3dgs}

Despite the quality advances of the NeRF family, a fundamental bottleneck
remained: volumetric ray marching requires hundreds of sequential MLP
evaluations per ray, making real-time rendering on commodity hardware essentially
unachievable.
3D Gaussian Splatting~(3DGS)~\citep{kerbl2023gaussian} resolved this constraint
decisively, achieving over 100\,fps at 1080p resolution by replacing the
continuous radiance field with a collection of explicit, anisotropic Gaussian
primitives that are rendered via tile-based rasterisation.
Beyond raw speed, the representation is fully explicit and directly editable:
individual Gaussians can be identified, moved, duplicated, and removed without
re-optimisation.
These properties have opened a new chapter in \nvs{} research, generating a
large and rapidly expanding literature that we organise here into four
thematic directions: quality and anti-aliasing improvements, compression and
memory efficiency, surface reconstruction, and scene editing and compositionality.
Table~\ref{tab:3dgs_methods} provides a comprehensive cross-family comparison.

\subsection{Foundational 3DGS}

3DGS~\citep{kerbl2023gaussian} achieved real-time ($>$100\,fps at 1080p)
NVS with training times of 30--45 minutes on commodity hardware,
surpassing NeRF-quality baselines on standard benchmarks.
Its three key innovations are:
(i)~anisotropic Gaussian primitives that align naturally with planar scene
structures;
(ii)~tile-based rasterisation that enables massively parallel GPU rendering;
(iii)~adaptive densification that automatically allocates capacity to
under-reconstructed regions.
The resulting representation is fully explicit and editable, opening new
directions in scene editing, simulation, and avatar rendering.

\paragraph{Quality and anti-aliasing.}
With the rendering speed bottleneck resolved, the focus shifted immediately
to improving the visual quality of Gaussian-based rendering, particularly in
regimes that exposed the approximations made by the original method.

Mip-Splatting~\citep{yu2024mipsplatting} identifies aliasing in vanilla 3DGS
as arising from the approximation of 3-D Gaussian convolution during 2-D
projection.
It introduces a 3-D smoothing filter derived from the sampling footprint,
enabling accurate multi-scale rendering and substantially reducing floaters
at large distances.
This is the direct Gaussian-space analogue of the integrated positional
encoding insight from Mip-NeRF, and it similarly establishes the anti-aliasing
baseline for the 3DGS literature.
AbsGS~\citep{ye2024absgs} revisits the densification criterion: replacing
the mean gradient with an absolute-value measure improves coverage in
thin structures and reduces needle-like artifacts near object boundaries,
demonstrating that the adaptive densification heuristic of vanilla 3DGS
leaves systematic quality gaps that a principled gradient measure can address.

\paragraph{Compression and memory efficiency.}
A single 3DGS scene may require 300\,MB--1\,GB of storage, limiting deployment
on mobile or embedded platforms and making it impractical to maintain multiple
scenes in GPU memory simultaneously.
Addressing this bottleneck has motivated a dedicated compression sub-field
that seeks to reduce storage cost while preserving rendering quality.

Scaffold-GS~\citep{lu2023scaffoldgs} introduces a sparse set of anchor
primitives from which neural networks predict neighbouring Gaussian attributes;
this structured representation reduces storage by $3$--$5\times$ with
comparable quality.
Rather than compressing an unstructured point cloud post-hoc, it imposes
architectural structure during optimisation, ensuring that each Gaussian
is efficiently described relative to its spatial context.
LightGaussian~\citep{fan2023lightgaussian} achieves $15\times$ compression via
global Gaussian pruning based on visibility contribution and subsequent
vector quantisation of attributes, demonstrating that the vast majority of
Gaussians contribute negligibly to the final render and can be aggressively
pruned without perceptible quality loss.
EAGLES~\citep{girish2024eagles} applies residual vector quantisation
for fine-grained attribute compression, pushing the quantisation
of Gaussian attributes beyond scalar to multi-level residual codebooks.
HAC~\citep{chen2024hac} proposes a hash-grid-based context model that predicts
per-Gaussian attribute entropy, enabling learned arithmetic coding for
extreme compression ratios by exploiting spatial correlations between
neighbouring Gaussians that simpler quantisation methods ignore.

\paragraph{Surface reconstruction.}
The Gaussian primitive is an inherently volumetric object: its density falls
off smoothly in all directions with no well-defined surface.
This makes mesh extraction from 3DGS non-trivial, yet many downstream
applications---physical simulation, collision checking, UV texture mapping,
and export to standard graphics pipelines---require explicit surface meshes.
A dedicated line of work has addressed this by encouraging Gaussians to
adopt disc-like or planar configurations aligned with scene surfaces.

SuGaR~\citep{guedon2024sugar} aligns Gaussians to the scene surface by adding
a regularisation term that encourages Gaussians to become flat discs at
the surface, followed by mesh extraction and joint Gaussian--mesh refinement.
This two-stage approach leverages the speed of Gaussian optimisation for
initial scene reconstruction and the geometric precision of mesh refinement
for final surface quality.
2DGS~\citep{huang2024_2dgs} takes a more principled approach by replacing
3-D Gaussians with 2-D disc primitives, which by construction lie on surfaces
and produce accurate normal maps; the representation guarantees surface
alignment without post-hoc regularisation.
Gaussian Opacity Fields~(GOF)~\citep{yu2024gof} fuses Gaussian Splatting with
a continuous opacity field for topologically correct surface extraction in
unbounded scenes, addressing the failure mode of earlier methods on complex
open-world geometry.

\paragraph{Scene editing and compositionality.}
The explicit Gaussian representation is particularly amenable to editing:
individual primitives can be identified, moved, duplicated, or removed.
This editability has inspired a line of work on semantically-aware and
compositional Gaussian scenes that bridges \nvs{} and scene understanding.

GaussianEditor~\citep{chen2024gaussianeditor} harnesses a large vision-language
model to select semantically relevant Gaussians from a text description and
optimises their appearance with a diffusion prior.
The approach demonstrates that the VLM's semantic segmentation can be
transferred to the Gaussian domain via rendered-view supervision, enabling
language-guided 3-D editing without any 3-D annotation.
Gaussian Grouping~\citep{ye2023gaussian_grouping} augments each Gaussian with
a learned identity field, trained with SAM~\citep{kirillov2023sam} masks from
rendered views, enabling instance-level part decomposition and editing.
By propagating 2-D instance masks into 3-D Gaussian space via the differentiable
renderer, it achieves zero-shot part decomposition that requires no 3-D
supervision.
LERF~\citep{kerr2023lerf} and LangSplat~\citep{qin2024langsplat} both
ground language into 3-D neural representations, with LERF doing so in
NeRF space and LangSplat directly in the Gaussian representation;
together they demonstrate that open-vocabulary 3-D querying---``show me all
chairs in this scene''---is achievable with language-image pretraining.
GALA3D~\citep{zhou2024gala3d} uses an LLM to generate a 3-D layout from text
and initialises Gaussians per-object, enabling compositional scene generation
that extends the scene-level 3DGS capabilities beyond reconstruction to text-driven
synthesis of entirely novel multi-object scenes.

The synthesis of editing-focused methods reveals an important pattern:
the Gaussian representation is increasingly serving as the interface between
2-D foundation models (SAM, CLIP, LLMs) and 3-D scene understanding,
because the differentiable rasteriser provides a natural gradient bridge
from rendered-view 2-D supervision to 3-D primitive attributes.
This architectural pattern---2-D supervision $\to$ Gaussian attributes---is
likely to become even more central as 2-D foundation models continue to improve.

Table~\ref{tab:3dgs_methods} provides a comprehensive view of the 3DGS family.

\begin{table*}[t]
  \centering
  \caption{Comprehensive 3D Gaussian Splatting method comparison.
    \textbf{Category}: STATIC=per-scene static, DYN=dynamic/deformable, FF=feed-forward generalizable.
    \textbf{RT}: real-time rendering ($>$30\,fps).
    \textbf{Comp.}: dedicated compression or storage reduction.
    \textbf{Surf.}: explicit surface extraction.
    \textbf{Unp.}: works on unposed images.
    \textbf{Input}: minimum input views for FF methods.}
  \label{tab:3dgs_methods}
  \scriptsize
  \setlength{\tabcolsep}{2.8pt}
  \begin{tabular}{@{}llcccccccp{3.8cm}@{}}
    \toprule
    \textbf{Method} & \textbf{Venue} & \textbf{Cat.}
      & \textbf{RT} & \textbf{Comp.} & \textbf{Surf.}
      & \textbf{Edit.} & \textbf{Unp.} & \textbf{Input}
      & \textbf{Key Contribution} \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Static per-scene}} \\
    3DGS~\citep{kerbl2023gaussian}               & TOG'23      & S & \yes& \no & \no & \no & \no & MV  & Tile-based Gaussian splatting \\
    Mip-Splatting~\citep{yu2024mipsplatting}     & CVPR'24     & S & \yes& \no & \no & \no & \no & MV  & 3-D smoothing filter, anti-alias \\
    AbsGS~\citep{ye2024absgs}                    & NeurIPS'24  & S & \yes& \no & \no & \no & \no & MV  & Absolute-gradient densification \\
    Scaffold-GS~\citep{lu2023scaffoldgs}         & CVPR'24     & S & \yes& \yes& \no & \no & \no & MV  & Anchor-driven neural attributes \\
    LightGaussian~\citep{fan2023lightgaussian}   & NeurIPS'23  & S & \yes& \yes& \no & \no & \no & MV  & Visibility pruning + VQ, 15$\times$ \\
    Compact3D~\citep{lee2024compact3d}           & arXiv'24    & S & \yes& \yes& \no & \no & \no & MV  & k-means codebook compression \\
    EAGLES~\citep{girish2024eagles}              & ECCV'24     & S & \yes& \yes& \no & \no & \no & MV  & Residual vector quantisation \\
    HAC~\citep{chen2024hac}                      & ECCV'24     & S & \yes& \yes& \no & \no & \no & MV  & Hash context entropy coding \\
    SuGaR~\citep{guedon2024sugar}                & CVPR'24     & S & \yes& \no & \yes& \pt & \no & MV  & Flat-disc + mesh joint opt. \\
    2DGS~\citep{huang2024_2dgs}                  & SIGGRAPH'24 & S & \yes& \no & \yes& \no & \no & MV  & 2-D disc primitives \\
    GOF~\citep{yu2024gof}                        & arXiv'24    & S & \yes& \no & \yes& \no & \no & MV  & Opacity field for surface \\
    GaussianEditor~\citep{chen2024gaussianeditor} & CVPR'24    & S & \yes& \no & \no & \yes& \no & MV  & LLM-guided semantic editing \\
    Gaussian Grouping~\citep{ye2023gaussian_grouping}& ECCV'24  & S & \yes& \no & \no & \yes& \no & MV  & SAM-based identity field \\
    GALA3D~\citep{zhou2024gala3d}                & ICML'24     & S & \yes& \no & \no & \yes& \no & MV  & LLM-layout compositional gen. \\
    LERF~\citep{kerr2023lerf}                    & ICCV'23     & S & \no & \no & \no & \pt & \no & MV  & CLIP language grounding in NeRF \\
    LangSplat~\citep{qin2024langsplat}           & CVPR'24     & S & \yes& \no & \no & \pt & \no & MV  & Language field in 3DGS \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Dynamic \& deformable}} \\
    4DGS~\citep{wu2024_4dgs}                     & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & 4-D Gaussian + temporal field \\
    Deformable3DGS~\citep{yang2024deformable}    & ICLR'24     & D & \yes& \no & \no & \no & \no & Mono& Canonical deformation network \\
    SpacetimeGaussians~\citep{yang2024spacetime} & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & Polynomial time trajectories \\
    SC-GS~\citep{huang2024scgs}                  & CVPR'24     & D & \yes& \no & \no & \no & \no & Mono& Sparse control point deform. \\
    Gaussian-Flow~\citep{lin2024gaussianflow}    & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & Fourier polynomial motion \\
    Shape-of-Motion~\citep{wang2024shapeofmotion}& arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Per-Gaussian 3-D motion \\
    Dynamic 3DGS~\citep{luiten2024dynamic3dgs}   & 3DV'24      & D & \yes& \no & \no & \no & \no & MV  & Persistent Gaussian tracking \\
    GauFRe~\citep{liang2024gaufre}               & arXiv'24    & D & \yes& \no & \no & \no & \no & Mono& Flow-regularised deformation \\
    MoVieS~\citep{sun2024movies}                 & arXiv'24    & D & \yes& \no & \no & \no & \no & Mono& 4-D synthesis in $\sim$1 second \\
    SC-4D~\citep{kim2024selfcal4d}               & arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Self-calibrating 4-D from video \\
    Reconstruct-Inpaint~\citep{han2024rif}       & arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Recon + inpaint + TTF pipeline \\
    Articulated-GS~\citep{li2024articulatedgs}   & NeurIPS'24  & D & \yes& \no & \no & \no & \no & MV  & Template-free articulated 3DGS \\
    EVA-Gaussian~\citep{hu2024evacg}             & arXiv'24    & D & \yes& \no & \no & \no & \no & MV  & Real-time human NVS \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Feed-forward generalizable}} \\
    pixelSplat~\citep{charatan2024pixelsplat}    & CVPR'24     & FF& \yes& \no & \no & \no & \no & 2   & Stereo epipolar features \\
    MVSplat~\citep{chen2024mvsplat}              & ECCV'24     & FF& \yes& \no & \no & \no & \no & 2--6& Cost-volume geometry prior \\
    TranSplat~\citep{zhang2024transplate}        & arXiv'24    & FF& \yes& \no & \no & \no & \no & 2+  & Transformer MV prediction \\
    FSGS~\citep{zhu2024fsgs}                     & CVPR'24     & FF& \yes& \no & \no & \no & \no & 3   & Gaussian unpooling, few-shot \\
    DepthSplat~\citep{xu2024depthsplat}          & arXiv'24    & FF& \yes& \no & \no & \no & \no & 2+  & Joint depth + Gaussian pred. \\
    Splatt3R~\citep{smart2024splatt3r}           & arXiv'24    & FF& \yes& \no & \no & \no & \yes& 2   & DUSt3R backbone, unposed \\
    TriplaneGaussian~\citep{zou2024triplanegaussian}& CVPR'24   & FF& \yes& \no & \no & \no & \no & 1   & Triplane + Gaussian, single-view \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Dynamic 3DGS}

Static 3DGS, while powerful, is limited to scenes that remain fixed over
time.
Extending Gaussian splatting to dynamic scenes requires mechanisms to model
the \emph{motion} of Gaussian primitives over time---a challenge that has
attracted considerable attention given the representation's unique combination
of explicit, editable structure and real-time rendering speed.
The key design question is whether to model motion as a \emph{deformation field}
applied to canonical Gaussians, or as an \emph{intrinsic temporal attribute}
directly parameterising each Gaussian's trajectory.

4DGS~\citep{wu2024_4dgs} takes the second approach, extending Gaussians into
4-D space-time by pairing each primitive with a temporal radiance field that
modulates its position and appearance as a function of time; this formulation
enables real-time rendering of complex dynamic scenes such as sports and dance.
SpacetimeGaussians~\citep{yang2024spacetime} assigns polynomial time
trajectories directly to each Gaussian's position and rotation, providing
a compact and analytically differentiable motion representation that avoids
the computational cost of an implicit deformation network.
Deformable3DGS~\citep{yang2024deformable} and SC-GS~\citep{huang2024scgs}
adopt the canonical deformation approach, following the D-NeRF paradigm but
in the Gaussian domain: Deformable3DGS applies a global deformation network
to all Gaussians, while SC-GS uses sparse control points that provide a
physically interpretable, structure-aware deformation field particularly
suited for articulated and human body scenes.

Gaussian-Flow~\citep{lin2024gaussianflow} models Gaussian dynamics via
polynomial motion fields in Fourier space, enabling compact temporal
parameterisation that naturally captures periodic motions such as walking
and repetitive industrial machinery.
Shape-of-Motion~\citep{wang2024shapeofmotion} takes a more ambitious approach,
jointly estimating per-Gaussian 3-D motion and global scene flow from
monocular video, operating in the most challenging input regime.
Dynamic 3D Gaussians~\citep{luiten2024dynamic3dgs} tracks scene dynamics
through persistent Gaussian identities across views, treating the tracking
and NVS problems as jointly constrained by the photometric consistency of
the Gaussian primitives.

GauFRe~\citep{liang2024gaufre} and MoVieS~\citep{sun2024movies} both target
the \emph{efficiency} end of the dynamic 3DGS spectrum:
GauFRe regularises the deformation field with dense optical flow to produce
temporally coherent results from monocular video, while MoVieS generates
motion-aware 4-D dynamic views in approximately one second via an
efficient Gaussian motion estimation pipeline.
Self-Calibrating 4D~\citep{kim2024selfcal4d} addresses the practical challenge
that monocular dynamic videos lack calibrated camera parameters, jointly
estimating camera intrinsics and extrinsics alongside the Gaussian deformation.
Reconstruct-Inpaint-Finetune~\citep{han2024rif} takes a pragmatic three-stage
approach---first reconstructing the static scene, then inpainting dynamic regions
with a generative model, and finally refining the result at test time---that
produces high-quality dynamic novel views without an end-to-end dynamic 3DGS model.
EVA-Gaussian~\citep{hu2024evacg} achieves real-time human NVS under diverse
multi-view settings.
Articulated Gaussian~\citep{li2024articulatedgs} enables template-free
articulated Gaussian Splatting for real-time reposable dynamic view synthesis.

Surveying the dynamic 3DGS landscape reveals a common tension: deformation-field
methods provide high quality but require per-scene optimisation, while
trajectory-attribute methods are faster but less expressive for complex
non-rigid motion.
The monocular dynamic synthesis problem remains substantially harder than
the multi-view case due to the depth-motion ambiguity, and the best current
methods still rely on additional supervision from optical flow and monocular
depth estimates to constrain the otherwise ill-posed optimisation.

\subsection{Feed-Forward Generalizable 3DGS}

The feed-forward 3DGS paradigm inherits the architectural pattern established
by generalisable NeRF---amortise scene-specific computation into a pretrained
model---while retaining the rendering speed advantage of the Gaussian representation.
The result is a family of models that synthesise novel views in a single
forward pass, without any per-scene optimisation, at real-time rendering speed.
This combination of generalisation and rendering efficiency is precisely what
downstream applications in robotics and autonomous driving require.

pixelSplat~\citep{charatan2024pixelsplat} established the feed-forward 3DGS
paradigm by predicting Gaussian primitives from stereo image pairs in a single
forward pass, using epipolar attention to exploit the known geometric relationship
between the two cameras.
The model implicitly solves stereo matching, depth estimation, and Gaussian
attribute prediction in a unified architecture.
MVSplat~\citep{chen2024mvsplat} adds explicit cost-volume priors for better
geometry under challenging viewpoints, demonstrating that the multi-view stereo
geometry prior---which proved essential in the generalisable NeRF literature---is
equally beneficial in the Gaussian domain.
TranSplat~\citep{zhang2024transplate} introduces a transformer architecture for
multi-view Gaussian prediction that attends across all input views jointly,
enabling better utilisation of cross-view geometric consistency cues.

Splatt3R~\citep{smart2024splatt3r} takes a qualitatively different approach by
building on DUSt3R~\citep{wang2024dust3r}, which predicts dense 3-D point maps
directly from unposed image pairs.
By treating the DUSt3R pointmaps as Gaussian positions and learning only the
Gaussian attributes (colour, opacity, scale) via a lightweight decoder,
Splatt3R achieves unposed image reconstruction---no camera calibration
required---which dramatically simplifies the capture pipeline.
FSGS~\citep{zhu2024fsgs} enables few-shot Gaussian Splatting via Gaussian
unpooling and monocular depth supervision, specifically addressing the
case where only three or fewer reference views are available.
DepthSplat~\citep{xu2024depthsplat} jointly predicts depth and Gaussians
in a single model, sharing representations between the depth estimation
and Gaussian prediction tasks to improve geometry quality.
TriplaneGaussian~\citep{zou2024triplanegaussian} combines triplane and
Gaussian representations in a transformer-based single-view reconstruction model,
using the triplane as an intermediate 3-D feature representation that is
then decoded into Gaussian primitives.

The feed-forward generalizable methods are included in the comprehensive
Table~\ref{tab:3dgs_methods} above (FF category).

% ─────────────────────────────────────────────────────────────────────────────
