\section{Generalisable Feed-Forward 3D Reconstruction}
\label{sec:generalisable}

The per-scene paradigm that dominates the NeRF and classical 3DGS literature
carries a fundamental limitation: every new scene requires a fresh optimisation
loop lasting tens of minutes to hours.
This is incompatible with real-world applications where new scenes are encountered
continuously---autonomous vehicles observing novel road segments, robots
entering unfamiliar rooms, AR systems overlaying content on new environments.
The \emph{feed-forward reconstruction} paradigm directly addresses this bottleneck:
a single, amortised model trained on large datasets of 3-D assets or multi-view
captures learns to predict a complete 3-D scene representation from one or
a few images in a single forward pass.

This section covers two complementary approaches to this goal.
We first discuss the pose-free foundation model line of work (DUSt3R, MASt3R,
and successors), which treats reconstruction as a direct regression problem
from image pairs to point maps without requiring camera calibration.
We then examine the Large Reconstruction Model (LRM) paradigm, which uses
large transformer models pretrained on 3-D asset repositories to predict
NeRF or Gaussian scenes from a single image.
Together, these two families represent the current state-of-the-art in
generalised single-image and sparse-view 3-D understanding.

\subsection{Pose-Free and Foundation Models}

The classical 3-D reconstruction pipeline requires camera calibration as a
prerequisite: Structure-from-Motion~\citep{schoenberger2016sfm} must be run
to estimate camera poses before any NeRF or 3DGS optimisation can begin.
This prerequisite is a significant practical obstacle---internet-sourced images
rarely come with calibration information, and running SfM on every image
collection adds latency.
A new generation of \emph{pose-free} models eliminates this prerequisite by
learning to predict 3-D geometry directly from image pairs, treating camera
pose estimation as an internal computation rather than an external preprocessing
step.

DUSt3R~\citep{wang2024dust3r} is the landmark work of this direction.
Rather than predicting depth or disparity (which require calibration to convert
to metric 3-D), DUSt3R predicts \emph{dense point maps}: for each pixel in
both input images, it directly regresses the 3-D coordinate in a shared
Euclidean frame.
The key insight is that a ViT backbone with cross-image attention can jointly
reason about both images and implicitly solve the correspondence and geometry
problems in a single forward pass.
For collections of more than two images, DUSt3R aligns pairwise predictions
via a global optimisation that finds the rotation and translation that minimises
pairwise reprojection error, producing a metrically accurate 3-D reconstruction
without any explicit SfM.

MASt3R~\citep{leroy2024mast3r} extends DUSt3R by adding a matching head that
produces dense feature descriptors alongside the point map predictions.
This addition enables accurate metric reconstruction from internet-sourced
photo pairs where appearance varies substantially---a regime where vanilla
DUSt3R's purely geometric predictions can accumulate errors---and directly
enables downstream applications such as localisation and structure-aware
image retrieval.

MonST3R~\citep{zhang2024monst3r} extends DUSt3R to dynamic video sequences,
where the scene contains independently moving objects.
By jointly estimating camera motion and scene-object motion within the same
point-map regression framework, MonST3R handles the dynamic case that DUSt3R
implicitly assumes away, substantially broadening the practical applicability
of pose-free reconstruction to casually captured video.

VGGT~\citep{wang2025vggt} represents the most recent advance in this line,
proposing a Visual Geometry Grounded Transformer that predicts a comprehensive
set of geometric quantities---point maps, depth, camera parameters, and feature
descriptors---from an arbitrary number of views in a single transformer forward
pass.
Its key innovation is the integration of structured geometric tokens that
provide an explicit 3-D coordinate system for the transformer's attention
operations, enabling the model to reason jointly about all geometric quantities
rather than treating each as an independent prediction head.

The pose-free foundation models collectively demonstrate that large-scale
pretraining on diverse 3-D data can substitute for the carefully designed
geometric pipelines that classical reconstruction required, opening the door
to reconstruction from casually captured, unposed image collections.
This generalisation is a direct enabler of world modeling at scale, where
collecting calibrated sensor data for every new environment is impractical.

\subsection{Large Reconstruction Models (LRM)}
\label{sec:lrm}

While the pose-free models address multi-view reconstruction, a parallel
line of work targets an even more challenging regime: reconstructing a
complete 3-D scene from a \emph{single} image.
Single-image reconstruction is fundamentally under-constrained without a
strong prior---the imaging process collapses 3-D geometry to a 2-D projection,
discarding all depth information and occluded geometry---and requires a
model that has internalised the distribution of plausible 3-D shapes for the
objects and scenes it encounters.
The Large Reconstruction Model~(LRM) paradigm~\citep{hong2023lrm} provides this
prior through large-scale pretraining: a transformer trained on hundreds of
thousands of 3-D assets from Objaverse~\citep{deitke2023objaverse} learns
to map a single input image to a complete triplane representation of a NeRF,
achieving complete 3-D reconstruction in under five seconds.

The LRM~\citep{hong2023lrm} architecture uses a DINO-ViT encoder followed by
a cross-attention transformer that maps image tokens to triplane tokens,
which are decoded by a NeRF MLP.
The key insight is that the transformer can amortise the geometric reasoning
of per-scene NeRF optimisation into its weights via large-scale pretraining:
by training on hundreds of thousands of objects with known ground-truth geometry,
the model learns to \emph{recognise} plausible 3-D shapes rather than
\emph{optimise} them.
Instant3D~\citep{li2023instant3d} decouples appearance generation from geometry
prediction, first generating sparse multi-view images with a diffusion model
and then feeding the result to an LRM-style reconstructor.
This two-stage design allows each component to be trained on different data
and to be improved independently, and it achieves text-to-3D generation
in under five seconds---a remarkable speed relative to the hours required by
SDS-based alternatives.
OpenLRM~\citep{openlrm2023} reproduces the LRM pipeline in an open-source
setting, enabling broader research access to the paradigm.

The Gaussian-based LRM branch has proven especially popular for its additional
rendering speed advantage.
LGM~\citep{tang2024lgm} generates multi-view images and lifts them to a 3DGS
scene via an asymmetric U-Net, achieving 3-D generation in seconds at
commercial-grade quality while retaining the real-time rendering advantage of
the Gaussian representation.
GRM~\citep{xu2024grm} predicts 3D Gaussians directly from image features using
a sparse-view transformer, operating entirely at the pixel level without any
intermediate NeRF or triplane representation; this pixel-level prediction
enables finer spatial resolution in the output Gaussian set.
Gamba~\citep{he2024gamba} addresses the computational bottleneck of the
transformer's quadratic attention cost by replacing it with a state-space
model (Mamba) for $\mathcal{O}(N)$ sequence length scaling, making long-context
3-D inference from high-resolution inputs feasible.
GS-LRM~\citep{zhang2024gslrm} combines a transformer backbone with a 3DGS
output head, training on multi-view sequences from Objaverse; it achieves
state-of-the-art single-image reconstruction quality by jointly training
the visual encoder and the Gaussian decoder end-to-end on reconstruction quality.

Two non-parametric reconstruction models serve as particularly strong baselines
in the feed-forward regime and additionally underpin the next generation of
generalised Gaussian reconstructors.
DUSt3R~\citep{wang2024dust3r} and MASt3R~\citep{leroy2024mast3r}---already
discussed in the pose-free section---provide unposed pairwise point maps that
Splatt3R~\citep{smart2024splatt3r} then converts to Gaussian splatting scenes
via a lightweight attribute prediction head, combining the geometric reliability
of DUSt3R with the rendering speed and editability of 3DGS.

\paragraph{Implications for World Modeling.}
LRM-style models demonstrate that a single transformer, with sufficient data
and compute, can learn a universal spatial prior---the same capacity required
by world models for instant scene initialisation from single-frame observations.
This is directly relevant to the world-modeling agenda: a world model that
must initialise a 3-D scene representation from a single observation
(a camera frame at the start of a robot episode, or a single photo of a room
to be manipulated) can leverage the LRM prior to obtain a geometrically coherent
starting point, rather than requiring dense multi-view capture.
We return to this connection in Section~\ref{sec:world}.

\subsection{Multi-View Stereo in the Neural Era}

Classical MVS has been reinvigorated by deep learning.
MVSNet~\citep{yao2018mvsnet} introduced cost-volume-based deep MVS, using a
differentiable homography warping to build a cost volume from multiple reference
views and predicting per-pixel depth via a 3-D CNN.
IterMVS~\citep{wang2022itermvs} and Uni-MVSNet~\citep{peng2022unimvsnet}
improved accuracy and generalisation through iterative GRU-based refinement and
unified feature matching, respectively.
These methods provide the geometry backbone for downstream NVS: their
cost-volume feature matching is directly incorporated into generalisable NeRF
(MVSNeRF~\citep{chen2021mvsnerf}) and feed-forward Gaussian methods
(MVSplat~\citep{chen2024mvsplat}), establishing a productive synergy between
the classical MVS and neural NVS communities that continues to be productive.

% ─────────────────────────────────────────────────────────────────────────────
