\section{Towards Generative World Modeling}
\label{sec:world}

The preceding sections have charted the evolution of \nvs{} from per-scene
NeRF reconstruction to generative, feed-forward synthesis from a single image.
We now address the central thesis of this survey: how does this progression
connect to, and ultimately enable, \emph{Generative World Modeling}?

A world model, in the canonical sense introduced by
Ha~\&~Schmidhuber~\citep{ha2018worldmodels} and formalised by
LeCun~\citep{lecun2022path}, is a system that predicts the future state of
the world given a current state and an action.
Applied to visual, spatial environments, this translates to predicting the
\emph{visual experience} of an agent as it moves through and interacts with
a scene.
The key insight of this survey is that \emph{novel view synthesis is already
a minimal world model for the camera-action case}: it predicts future visual
states under camera motion, enforcing geometric consistency without explicit
physics.
The challenge is to extend this capability along three additional axes:
temporal dynamics that extend beyond the current frame,
agent actions that are richer than camera transforms,
and physical plausibility that respects real-world causality.

This section surveys how the field is meeting this challenge across four
interconnected directions: 4-D generation and dynamic scene modelling,
neural scene simulators for autonomous driving, large-scale world foundation
models, and embodied AI applications.
We conclude with a formal capability analysis (Table~\ref{tab:nvs_wm}) that
maps the axis-by-axis progress and identifies the remaining gaps.

\subsection{Dynamic Scenes and 4-D World Representation}

The step from static \nvs{} to world modeling hinges on modelling
\emph{temporal dynamics}: how does a scene evolve over time, and can
the model generate novel views at arbitrary future timestamps?
The dynamic NeRF and 4DGS methods surveyed in \S\ref{sec:nerf}--\ref{sec:3dgs}
address this for \emph{observed} scenes given dense multi-view video.
The harder and more practically relevant regime is \emph{generalisable}
4-D generation from sparse or single-image input---precisely the regime
inhabited by world models.
The key question is whether the powerful 2-D and video priors of large
diffusion models can be elevated to serve as 4-D world priors.

Video diffusion models have emerged as the most promising backbone for this
purpose, because the temporal axis of a video model naturally encodes
the sequential structure of a scene evolving over time.
The principal design challenge is coupling the video generation process
to spatial geometry, so that the generated frames are not merely temporally
coherent but also \emph{multi-view consistent}---a stronger requirement
that the pure video generation training objective does not enforce.

ViewCrafter~\citep{yu2024viewcrafter} addresses this coupling through a
geometric scaffold: it augments a video diffusion model with point-cloud
renderings along the target camera trajectory, grounding each generated
frame in the known 3-D structure of the reference scene.
This scaffold design is principled because it decouples the ``what is the
geometry'' (answered by the point cloud) from the ``how does it look''
(answered by the diffusion prior), allowing each component to
contribute its respective strength.
CAT4D~\citep{wu2024cat4d} generalises the Cat3D framework to the temporal
dimension, jointly modelling viewpoint and time with a multi-view video
diffusion model, then lifting the output to an explicit 4-D representation;
by training jointly over the view and time axes, it learns a shared prior
that enforces consistency in both dimensions simultaneously.
4DiM~\citep{watson2024_4dim} takes a more direct approach, training a
4-D diffusion model that simultaneously denoises a set of frames at
varying timestamps and viewpoints, learning a joint spatiotemporal prior
without relying on a two-stage pipeline.

SV3D~\citep{voleti2024sv3d} and SV4D~\citep{xie2024sv4d} extend Stable
Video Diffusion to 3-D and 4-D object generation respectively by conditioning
on orbital camera paths; the key observation is that the temporal motion
prior of a video diffusion model is a strong proxy for the viewpoint
change prior of a multi-view model when the conditioning signal specifies
the camera trajectory.
Consistent4D~\citep{jiang2024consistent4d} enforces temporal consistency
across a 4-D cascade, resolving inter-frame Gaussian drift that arises
when each frame of the 4-D Gaussian sequence is optimised independently.
DreamGaussian4D~\citep{ren2023dreamgaussian4d} demonstrates that SDS can
be applied to 4DGS, enabling text-driven 4-D generation without any video
training data, extending the 3-D SDS paradigm to the temporal domain.

Beyond video-diffusion-based approaches, GaussianFlow~\citep{gao2024gaussianflow}
models 4-D content creation by splatting Gaussian dynamics, providing a
principled connection between the explicit Gaussian representation and
fluid-like continuous flow dynamics.
L4GM~\citep{ren2024l4gm} scales the LRM paradigm to 4-D, training a large
reconstruction model directly on 4-D data to generate consistent dynamic
objects from single images in seconds.
STAG4D~\citep{zeng2024stag4d} introduces spatial-temporal anchored Gaussian
generation that addresses the long-range consistency failures common in
4-D generation by anchoring Gaussian clusters to stable spatial positions.
Diffusion4D~\citep{diffusion4d2024} achieves fast spatial-temporal consistent
4-D generation via a two-stage video diffusion pipeline, reducing the
inference time of 4-D generation by an order of magnitude.

A complementary line of work addresses the \emph{monocular} case, where
only a single camera view is available throughout time.
Diffusion Priors for Dynamic NVS~\citep{li2024diffpriordynamic} treats the
diffusion model as a frame-level regulariser during per-scene 4-D optimisation
from a single monocular video, substantially improving novel-view quality in
under-constrained regions by providing a learned hallucination prior that
fills in views never observed.
Dynamic View Synthesis from Small Camera Motion~\citep{li2024smallmotion}
addresses the particularly challenging case where camera motion is small and
depth cues are minimal, using a generative prior to resolve the depth-motion
ambiguity.

Taken as a whole, the 4-D generation literature demonstrates that the combined
signal from spatial consistency (from NVS training) and temporal consistency
(from video diffusion training) is sufficient to generate plausible 4-D world
states from sparse input.
The remaining frontier is \emph{physical grounding}: ensuring that the
generated dynamics are not just perceptually coherent but also obey physical
laws---a challenge we return to in \S\ref{sec:open}.

\subsection{Neural Scene Simulators for Autonomous Driving}

Autonomous driving provides a uniquely demanding test-bed for NVS-as-world-model:
scenes are large-scale, unbounded, and contain independently moving actors
(vehicles, pedestrians) that must be rendered consistently from closed-loop
ego-motion trajectories that may deviate substantially from the recorded path.
The downstream value of accurate simulation is concrete and measurable:
closed-loop evaluation of autonomous systems requires the ability to replay
any scenario with arbitrary modifications to weather, traffic density, and
actor behaviour.
Three architectural strategies have emerged, representing progressively stronger
integration between scene representation and generative modelling.

\paragraph{NeRF-based compositional simulators.}
The first strategy separates the scene into independently-modelled components
that can be edited and composited.
MARS~\citep{wu2023mars} separates the scene into a static NeRF background
and per-agent NeRF foregrounds, allowing independent actor manipulation---an
approach that directly addresses the compositionality requirement of scene
editing but requires per-instance annotations.
UniSim~\citep{yang2023unisim} extends this to a closed-loop sensor simulator
that supports controllable scene editing (removing actors, changing weather,
inserting novel agents) with photorealistic rendering at the ego sensor,
demonstrating that a compositional NeRF scene model can serve as a full
sensor simulation system.
EmerNeRF~\citep{yang2023emernerf} removes the requirement for per-actor
annotations by learning the dynamic-static decomposition in a self-supervised
manner from scene flow estimated via multi-camera video; this is an important
step toward scalable scene decomposition that does not require expensive
human labelling.

\paragraph{3DGS-based driving reconstruction.}
The second strategy replaces NeRF with Gaussian Splatting, inheriting its
real-time rendering advantage, which is critical for closed-loop simulation
at scale.
Street Gaussians~\citep{yan2024streetgaussian} extends 3DGS to urban scenes
by representing tracked foreground objects with separate Gaussian models and
compositing them over a background Gaussian field; the per-object Gaussian
decomposition supports instance-level editing without re-training the full
scene model.
PVG~\citep{chen2023pvg} models Gaussian motion with periodic vibration functions
that efficiently encode the slow, cyclic motions common in urban traffic,
such as the repetitive movement of wipers or the oscillation of pedestrian gaits.
OmniRe~\citep{chen2024omnire} unifies background, foreground, and sky in
a single Gaussian framework, establishing a strong all-in-one baseline that
subsumes the separate components of earlier compositional approaches.
DrivingGaussian~\citep{wang2024drivinggaussian} addresses large-scale urban
environments by compositing incremental static and dynamic Gaussian fields
with a composite dynamic Gaussian graph for multi-camera consistency.
NeuRAD~\citep{tonderski2024neurad} and SplatAD~\citep{hess2024splatad} further
scale Gaussian-based driving simulation to production-grade multi-sensor (LiDAR
plus camera) settings, demonstrating that the representation can accommodate
the full sensor suite of a commercial autonomous vehicle.

\paragraph{Generative world models for autonomous driving.}
The third strategy transcends deterministic scene reconstruction by training
\emph{generative} models that can produce novel, physically plausible driving
scenarios beyond the recorded dataset.
This capability is essential for safety validation, which requires testing on
corner cases that are rare or absent in recorded data.

DriveDreamer~\citep{wang2023drivedreamer} is the pioneering work in this
direction, explicitly framing autonomous driving scene generation as a world
model problem by training a diffusion model conditioned on structured traffic
annotations and ego-motion to produce photorealistic, temporally coherent
multi-camera video.
DriveDreamer-2~\citep{zhao2024drivedreamer2} scales this approach with
LLM-generated traffic layouts, enabling greater diversity and user-controllable
scenario editing without re-training, by treating the LLM as a compositional
scene language that the diffusion model can interpret.
Drive-WM~\citep{wang2023drive_wm} couples scene prediction with planning,
demonstrating that a world model trained on real driving data can directly
improve downstream planning metrics---a result that validates the practical
value of world model training for the end task of safe navigation.
OccWorld~\citep{zheng2024occworld} takes a 3-D occupancy perspective,
learning a world model that predicts future occupancy grids jointly with
ego-motion, unifying scene forecasting and motion planning in a single
autoregressive framework that operates in explicit 3-D space rather than
image space.
WoVoGen~\citep{lu2023wovogen} addresses the multi-camera consistency challenge
by reasoning over a shared 3-D world volume, ensuring that all generated
camera views are geometrically coherent---a non-trivial constraint when
cameras have limited overlap.

Complementing these end-to-end world models, several methods address specific
components of the simulation pipeline.
StreetCrafter~\citep{yan2024streetcrafter} uses a point-cloud-conditioned video
diffusion model with controllable camera motion for data augmentation in
street-view synthesis, bridging the generation and reconstruction paradigms.
Sat2Vid~\citep{wu2022sat2vid} lifts a satellite image to a street-view
panoramic video, addressing the altitude gap between aerial mapping data
and ground-level perception---a capability that could enable training data
generation for any mapped region without physical vehicle traversal.
Wild-GS~\citep{xu2024wildgs} solves the unconstrained internet-photo case by
disentangling transient and persistent Gaussian appearance in 3DGS,
enabling reconstruction from tourist photos with varying weather and crowds.

The driving simulation literature illustrates the full range of the NVS-to-WM
transition: at one end, reconstructive methods (MARS, OmniRe) provide accurate
but bounded-to-observed-data replays; at the other, generative world models
(DriveDreamer, OccWorld) provide controllable synthesis of novel scenarios.
The ideal system would combine the geometric accuracy of reconstruction with
the diversity and controllability of generative models---an active open problem.

\subsection{Generative World Models and Foundation Models}

World foundation models represent the apex of the NVS-to-world-modeling
trajectory: systems that jointly handle viewpoint, time, and action, trained
at internet scale.
These systems are distinguished from domain-specific simulators by their
breadth: they are designed to generalise across diverse environments and task
domains rather than to achieve maximum fidelity in a narrow setting.
We identify three design axes that differentiate current approaches.

\paragraph{Sequence vs.\ spatial representation.}
The first design axis concerns whether the world state is represented
as a flat sequence of tokens or as a spatially structured representation.
Flat-sequence models are simpler to train at scale but sacrifice explicit
geometric structure; spatial models are harder to train but support
view-consistent generation by construction.

GAIA-1~\citep{hu2023gaia} treats driving scenes as sequences of video, text,
and action tokens, training an autoregressive transformer to predict the next
token.
This approach achieves strong temporal coherence but lacks an explicit 3-D
representation, limiting viewpoint flexibility to the trained camera configurations.
Genie~\citep{bruce2024genie} learns a latent action space from unsupervised
internet videos, enabling interactive game-world generation without action
labels---a significant step toward self-supervised world model pretraining
that does not require expensive action annotation.
The latent action discovery capability of Genie is particularly noteworthy
because it demonstrates that world model training does not require labelled
action data, which is scarce, but can exploit the vastly more abundant
unlabelled video data.
WorldDreamer~\citep{worlddreamer2024} adopts a masked-token prediction
objective over a joint video-text tokenisation, demonstrating that masked
world model pre-training transfers well to diverse downstream generation
tasks---analogously to how BERT-style pre-training transfers to NLP tasks.

\paragraph{3-D-grounded generation.}
Genie~2~\citep{parkerholder2024genie2} introduces persistent 3-D structure
into the Genie framework, generating scenes that maintain geometry across
viewpoints and support first-person interactive navigation---currently the
closest system to an NVS-grounded world model for open-ended environments.
Its ability to generate geometrically consistent first-person view sequences
in novel environments from a single image is a striking demonstration of
the power of combining video diffusion with spatial grounding.
WonderWorld~\citep{yu2024wonderworld} takes a complementary approach:
instead of learning from large-scale video, it uses instant 3DGS initialisation
from a single image combined with diffusion inpainting to generate interactive
3-D scenes on the fly, demonstrating that a lightweight system can produce
interactive world experiences without any world model pretraining.
WonderJourney~\citep{yu2023wonderjourney} chains text-guided scene generation
steps to produce perpetually explorable 3-D environments, addressing the
long-horizon generation challenge by breaking it into sequential local steps.
Wonderland~\citep{liang2024wonderland} enables navigation through 3-D scenes
from a single image by combining monocular depth estimation with video diffusion,
demonstrating that the combination of geometric lifting and generative inpainting
can produce compelling spatial experiences.
AVID~\citep{rigter2024avid} adapts pre-trained video diffusion models to serve
as general-purpose world models for offline reinforcement learning, demonstrating
the deep connection between video generation and world modeling:
a model trained for visual prediction can directly serve as an environment model
for policy learning without task-specific world model training.
Structured World Models~\citep{mendonca2023structured} learn structured
representations from human videos in a fully unsupervised manner,
enabling policy learning from video without any action labels.

\paragraph{Physical-world foundation models.}
Cosmos~\citep{agarwal2025cosmos} is the most ambitious current effort: a family
of world foundation models pretrained on internet-scale video with explicit
physical world conditioning, designed for downstream fine-tuning to robotics
and autonomous driving.
Its training methodology---combining large-scale video pretraining with
physics-informed conditioning---represents the field's most direct attempt
to learn physical world dynamics from observation alone.
Pandora~\citep{xiang2024pandora} interleaves language and video tokens,
enabling language-driven world simulation with both visual and linguistic output;
the joint language--video generation capability is essential for instruction-following
world models that must respond to natural language commands.
The \emph{Is Sora a World Simulator?} analysis~\citep{zhu2024sora_survey}
provides a comprehensive assessment of the relationship between video generation
models and world simulation capability, identifying key gaps in physical
consistency, causal understanding, and controllability that remain open
in current video foundation models.
This gap analysis is directly relevant to our survey because it quantifies how
far the current best video generation systems fall short of a full world model.

\paragraph{The NVS-to-WM transition.}
A key architectural trend is the convergence of NVS and WM techniques.
Whereas early world models like GAIA-1 operated in 2-D video space,
recent systems increasingly incorporate explicit 3-D representations---whether
Gaussian splats, triplane NeRFs, or occupancy grids---as the world state
that the generative model predicts and renders.
This convergence has both practical and theoretical significance.
First, 3-D-aware world models naturally support multi-camera and novel-viewpoint
prediction, which is essential for robotic and driving applications where the
sensor configuration may change at test time.
Second, the NVS community's work on generalisable reconstruction provides the
initialisation and geometry backbone that world models need to maintain spatial
consistency.
Third, the world model community's work on temporal prediction and action
conditioning provides the dynamics prior that pure NVS systems lack.
Table~\ref{tab:nvs_wm} captures this convergence precisely via the dual axis of
\emph{spatial generalisation}~(D1--D2) and \emph{temporal/action conditioning}~(D3--D5).

Table~\ref{tab:worldmodels} compares world modeling methods.

\begin{table}[t]
  \centering
  \caption{World modeling and neural scene simulation methods.}
  \label{tab:worldmodels}
  \scriptsize
  \setlength{\tabcolsep}{3pt}
  \begin{tabular}{@{}llccccc@{}}
    \toprule
    \textbf{Method} & \textbf{Year}
      & \textbf{3D Aware} & \textbf{Video} & \textbf{Action}
      & \textbf{Domain} & \textbf{Key Feature} \\
    \midrule
    GAIA-1~\citep{hu2023gaia}                  & 2023 & \pt  & \yes & \yes & Driving    & Seq. token WM \\
    UniSim~\citep{yang2023unisim}              & 2023 & \yes & \yes & \yes & Driving    & Sensor sim. \\
    MARS~\citep{wu2023mars}                    & 2023 & \yes & \yes & \pt  & Driving    & Compositional \\
    DriveDreamer~\citep{wang2023drivedreamer}  & 2024 & \pt  & \yes & \yes & Driving    & Diffusion WM \\
    OccWorld~\citep{zheng2024occworld}         & 2024 & \yes & \yes & \yes & Driving    & 3D occupancy WM \\
    WoVoGen~\citep{lu2023wovogen}              & 2024 & \yes & \yes & \yes & Driving    & World volume \\
    Drive-WM~\citep{wang2023drive_wm}          & 2024 & \pt  & \yes & \yes & Driving    & MV forecasting \\
    Genie~\citep{bruce2024genie}               & 2024 & \no  & \yes & \yes & Games      & Latent action \\
    Genie~2~\citep{parkerholder2024genie2}     & 2024 & \yes & \yes & \yes & Open-world & Interactive 3D \\
    WonderWorld~\citep{yu2024wonderworld}      & 2024 & \yes & \no  & \no  & Open-world & Instant 3DGS \\
    ViewCrafter~\citep{yu2024viewcrafter}      & 2024 & \yes & \yes & \no  & Open-world & Cam-cond video \\
    WorldDreamer~\citep{worlddreamer2024}      & 2024 & \no  & \yes & \pt  & General    & Masked token WM \\
    AVID~\citep{rigter2024avid}                & 2024 & \no  & \yes & \yes & General    & VDM$\to$WM \\
    Cosmos~\citep{agarwal2025cosmos}           & 2025 & \yes & \yes & \yes & Physical   & WFM training \\
    Street Gaussians~\citep{yan2024streetgaussian}& 2024& \yes & \yes & \pt & Driving   & 3DGS driving \\
    OmniRe~\citep{chen2024omnire}              & 2024 & \yes & \yes & \yes & Driving    & All-in-one \\
    DrivingGaussian~\citep{wang2024drivinggaussian}& 2024& \yes& \yes& \pt & Driving   & Large-scale GS \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Embodied AI and Robotics}

NVS is more than a rendering tool in robotics: it provides the
\emph{spatial prior} that allows agents to reason about unseen parts of
the environment before taking action.
This is precisely the world-model capability that matters for robotic
manipulation and navigation---the ability to predict ``what would I see
if I moved there?'' without physically executing the motion.
We identify three use patterns that span the spectrum from pure pose
estimation to full sim-to-real transfer.

\textbf{Pose estimation and object localisation.}
iNeRF~\citep{yen2021inerf} inverts a trained NeRF to estimate 6-DoF camera
pose by gradient descent on the rendered photometric error---the first
demonstration that NeRF's differentiable renderer can be repurposed for
\emph{inference} (estimating the state of the world) as well as reconstruction
(building a model of it).
This inversion paradigm anticipates the broader use of differentiable
neural renderers for perception tasks.
Dex-NeRF~\citep{ichnowski2021dex} applies NeRF to transparent-object
grasping, where conventional depth sensors fail; by leveraging NeRF's
ability to model view-dependent appearance, it infers geometry from
specular reflections that are invisible to depth cameras.

\textbf{Visual planning and manipulation.}
NeRF-Supervisor~\citep{ze2023nerf_supervisor} uses a pre-trained NeRF to
synthesise robot-wrist camera views at hypothetical future poses, enabling
visual planning without physical rollouts---a direct application of NVS as
a world model for robot manipulation.
GaussianGrasping~\citep{shorinwa2024gaussiangrasping} replaces NeRF with
3DGS for real-time manipulation planning, exploiting the explicit Gaussian
representation for fast collision checking and grasp pose optimisation;
the real-time rendering speed of 3DGS is essential for reactive planning
in dynamic environments.

\textbf{Simulation-to-real transfer.}
RoboGSim~\citep{li2024robogssim} reconstructs robot workspaces in 3DGS and
renders synthetic training demonstrations from novel viewpoints, directly
closing the sim-to-real gap for visuomotor policy learning.
By generating photorealistic training data at arbitrary viewpoints and
configurations---including viewpoints that were never physically observed---it
demonstrates that a 3DGS world model can serve as a data augmentation engine
for downstream policy learning.

The robotic applications of NVS reveal the concrete value proposition of
geometric world models: they enable agents to reason about unseen viewpoints
and states without physical exploration, reducing the cost and risk of
real-world data collection.
This is the most direct current application of world modeling to embodied AI,
and it points toward a future in which robots continuously update and query
their spatial world models as part of closed-loop task execution.

\subsection{Human Novel View Synthesis}

Synthesising novel views of humans is a critical subproblem with applications
in AR/VR, telepresence, and embodied AI.
The challenge is distinct from general scene NVS because human bodies have
a known structural prior (the kinematic skeleton and body shape parameterised
by SMPL) that can serve as a geometric scaffold, but also exhibit complex
non-rigid deformation (clothing, hair, facial expression) that the scaffold
alone cannot represent.

\paragraph{Neural implicit human models.}
Neural Body~\citep{peng2021neuralbody} was the first work to combine the
SMPL structural prior with NeRF, anchoring structured latent codes on
SMPL vertices to enable novel view and pose synthesis with strong
body geometry constraints.
Rotationally-Temporally Consistent~\citep{liu2021rtcnvs} enforces both
rotation and temporal consistency for human performance video synthesis,
addressing the specific failure mode of per-frame NeRF methods that produce
flickering or geometrically inconsistent results across frames.
NHR~\citep{wu2020nhr} adopts multi-view point-cloud-based rendering for
dynamic human synthesis, demonstrating that explicit point representations
can support high-quality human rendering before the 3DGS era.

\paragraph{Generalisable and diffusion-based human NVS.}
CFSynthesis~\citep{li2023cfsynth} enables controllable and free-view 3-D
human video synthesis via a disentangled representation that separates body
pose, shape, and appearance, enabling independent control of each.
iVS-Net~\citep{chen2024ivsnet} learns human view synthesis from internet
videos without any 3-D supervision, demonstrating that the abundance of
2-D human video on the internet is sufficient to learn a strong human NVS
prior without expensive multi-view capture studios.
Diffuman4D~\citep{zheng2024diffuman4d} achieves 4-D consistent human view
synthesis from sparse-view videos via video diffusion, extending the
video-diffusion-as-4D-prior paradigm to the human domain.
EVA-Gaussian~\citep{hu2024evacg} achieves real-time 3DGS-based human NVS
under diverse multi-view settings, combining the real-time rendering speed
of 3DGS with a feed-forward generalisation architecture for fast inference
on unseen subjects.
Blur2Sharp~\citep{dong2024blur2sharp} addresses human novel pose and view
synthesis from blurry input with a generative prior, targeting the practical
setting where captured reference images have motion blur.

Human NVS methods represent an important special case of the general world
modeling problem: the human body is a system with known structure, highly
constrained dynamics (governed by skeletal kinematics), and rich
appearance variation (clothing, lighting, expression).
The success of structured human models suggests a general design principle
for world modeling: incorporating domain-specific structural priors
substantially reduces the burden on the learned generative component.

\subsection{The Boundary: NVS vs.\ World Modeling}

Table~\ref{tab:nvs_wm} delineates the capability boundary between NVS and
full World Modeling, using the five desiderata (D1--D5) introduced in
\S\ref{sec:background}.
The table reveals that contemporary \nvs{} systems have substantially closed
gaps D1 and D2---geometric consistency and sparse-input generalisation---but
that D3 (temporal dynamics), D4 (action conditioning), and D5 (physical
plausibility) remain only partially addressed even by the most advanced
world foundation models.
This analysis defines the open research frontier we discuss in \S\ref{sec:open}.

\begin{table}[t]
  \centering
  \caption{NVS vs.\ World Modeling capability boundary.}
  \label{tab:nvs_wm}
  \small
  \setlength{\tabcolsep}{5pt}
  \begin{tabular}{@{}lcc@{}}
    \toprule
    \textbf{Capability} & \textbf{NVS (static)} & \textbf{World Model} \\
    \midrule
    Viewpoint-consistent rendering~(D1)   & \yes & \yes \\
    Sparse/single-image input~(D2)        & \pt  & \yes \\
    Temporal dynamics~(D3)                & \no  & \yes \\
    Action/agent conditioning~(D4)        & \no  & \yes \\
    Physical plausibility~(D5)            & \no  & \yes \\
    Cross-scene generalisation            & \pt  & \yes \\
    Causal scene understanding            & \no  & \yes \\
    \bottomrule
  \end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
