\section{Datasets and Benchmarks}
\label{sec:datasets}

Empirical progress in \nvs{} and world modeling is inseparable from the
availability of suitable datasets.
A training or evaluation dataset defines the implicit scope of the problem:
it determines which scene types, capture modalities, viewpoint distributions,
and annotation types a method must handle.
The field has developed a rich ecosystem of datasets spanning a wide range of
domains---from controlled synthetic objects to urban driving sequences---and
understanding their respective strengths and limitations is essential for
interpreting benchmark results.
We organise the dataset survey along four axes: object-centric datasets
for single-image and multi-view reconstruction, static scene datasets for
per-scene NeRF and 3DGS evaluation, dynamic scene datasets for temporal NVS,
and large-scale autonomous driving datasets for neural scene simulation.
Table~\ref{tab:datasets} at the end of this section provides a structured comparison.

\subsection{Object-Level Datasets}

Object-centric datasets are the foundation of the generative 3-D literature,
because they provide clean, well-posed reconstruction problems with known
geometry and camera parameters, enabling rigorous evaluation of single-image
and sparse-view reconstruction methods.

\textbf{ShapeNet}~\citep{chang2015shapenet} established the canonical benchmark
for single-image 3-D reconstruction with its collection of 51K CAD models
across 55 object categories.
Its clean geometry and known cameras make it ideal for controlled evaluation,
though the synthetic rendering distribution departs substantially from real-world
photograph appearance.
\textbf{Objaverse}~\citep{deitke2023objaverse} resolved the scale bottleneck
by providing over 800K 3-D objects sourced from the internet, enabling the
large-scale pretraining required by LRM-family methods that would overfit on
ShapeNet's limited diversity.
\textbf{Objaverse-XL}~\citep{deitke2024objaversexl} further scales to 10M+ assets,
providing the data regime required by the next generation of foundation reconstruction models.
\textbf{Co3D}~\citep{reizenstein2021co3d} provides 19K+ annotated per-category
real video sequences, enabling evaluation of category-level NVS at a much higher
realism than synthetic benchmarks; its structured train/test splits support
rigorous evaluation of generalisation within and across categories.
\textbf{MVImgNet}~\citep{yu2023mvimgnet} adds 238K multi-view real-image sequences
across 238 categories, bridging the synthetic-to-real gap while maintaining
the structured multi-view annotation needed for NVS evaluation.

\subsection{Static Scene Datasets}

Static scene datasets test the fidelity and generalisation of per-scene
and generalisable NVS methods on real-world complexity, without the
confounding factor of temporal dynamics.
They span the full range of scene types encountered in practice, from
controlled tabletop objects to unbounded outdoor environments.

\textbf{NeRF-Synthetic}~\citep{mildenhall2020nerf} provides 8 synthetic
bounded objects with known geometry and high-quality Blender renders, making it
the canonical baseline for evaluating per-scene NeRF methods under controlled
conditions.
\textbf{LLFF}~\citep{mildenhall2019llff} provides 8 real forward-facing scenes
captured with a handheld camera, testing methods' ability to handle real
image noise, exposure variation, and limited view range.
\textbf{DTU}~\citep{jensen2014dtu} offers 124 controlled multi-view stereo scenes
with ground-truth point clouds, enabling combined NVS and geometry evaluation.
\textbf{Tanks and Temples}~\citep{knapitsch2017tanks} provides large-scale
indoor and outdoor scenes captured with professional cameras, testing scalability
to scenes that are substantially larger and more complex than tabletop objects.
\textbf{Mip-NeRF~360}~\citep{barron2022mipnerf360} is the current gold standard
for unbounded 360$^\circ$ evaluation, with 9 scenes carefully selected to
represent diverse outdoor and indoor environments with non-trivial background depth.
\textbf{ScanNet}~\citep{dai2017scannet} provides 1513 indoor RGB-D sequences with
semantic and instance labels, serving as the primary benchmark for indoor NVS
methods that must handle the large viewpoint variation and partial occlusion
typical of room-scale captures.
\textbf{ScanNet++}~\citep{yeshwanth2023scannetpp} upgrades this to DSLR-quality
imagery alongside iPhone captures, enabling evaluation of high-fidelity
indoor NVS models.
\textbf{RealEstate10K}~\citep{zhou2018stereoMagnification} provides 10K real
estate walkthrough videos with estimated camera trajectories, serving as the
primary training and evaluation set for generalisable video NVS methods
that must handle large camera motions and diverse indoor scenes.

\subsection{Dynamic Scene Datasets}

Dynamic scene datasets are essential for evaluating methods that target
the temporal generalisation requirements of world modeling.
They expose the additional challenges of large non-rigid deformation,
occlusion handling under motion, and the depth-motion ambiguity that
plagues monocular dynamic reconstruction.

\textbf{D-NeRF}~\citep{pumarola2021dnerf} provides 6 monocular synthetic
dynamic objects with known deformation fields, enabling quantitative
evaluation of dynamic reconstruction methods under controlled conditions.
\textbf{HyperNeRF}~\citep{park2021hypernerf} provides 15 real monocular sequences
of objects undergoing topological changes (opening a noodle box, cutting a
cucumber), testing methods' ability to represent topology-changing dynamics
that smooth deformation fields cannot model.
\textbf{NHR}~\citep{wu2020nhr} and \textbf{CMU Panoptic}~\citep{joo2015panoptic}
provide multi-view captures of human performance, with Panoptic offering 65
sequences of social activities from a dense 31-camera array---the richest
multi-view human performance dataset publicly available.
\textbf{Neu3D}~\citep{li2022neural3d} provides 6 high-quality multi-view
human video sequences that have become the standard benchmark for
multi-view dynamic NeRF evaluation.
\textbf{iPhone Dataset}~\citep{guo2022neural} provides 14 casually-captured
phone video sequences of dynamic scenes, testing methods' ability to handle
the most challenging regime: monocular capture with arbitrary camera and
scene motion.

\subsection{Autonomous Driving and Large-Scale}

Autonomous driving datasets provide the large-scale, sensor-rich captures
needed to evaluate neural scene simulators and driving-domain world models.
They are distinguished from the above categories by their multi-sensor
modality (cameras plus LiDAR), large-scale sequential capture,
semantic and instance annotations, and ego-motion ground truth.

\textbf{Waymo Open}~\citep{sun2020waymo} provides 1000 multi-camera plus LiDAR
driving segments covering diverse geographic and weather conditions, serving as
the primary evaluation benchmark for neural driving simulators.
\textbf{nuScenes}~\citep{caesar2020nuscenes} adds 360$^\circ$ camera coverage
and HD maps to 1000 driving scenes, enabling evaluation of methods that
must maintain consistency across the full surround-view camera ring.
\textbf{KITTI-360}~\citep{liao2022kitti360} provides urban driving sequences
with panoptic segmentation labels, enabling combined NVS and semantic evaluation.
\textbf{Argoverse~2}~\citep{wilson2023argoverse2} provides 1000 high-resolution
360$^\circ$ driving sequences with precise sensor calibration and HD map
annotations, supporting the evaluation of large-scale neural scene simulators
at the current frontier of sensor quality.
\textbf{PandaSet}~\citep{xiao2021pandaset} provides 103 scenes with a
forward-facing LIDAR plus multiple cameras, useful for evaluating fusion of
camera and LiDAR NVS in a single consistent benchmark.

Table~\ref{tab:datasets} provides a structured comparison of key datasets.

\begin{table*}[t]
  \centering
  \caption{NVS and world modeling datasets.
    \textbf{Type}: RT=real, Syn=synthetic.
    \textbf{Dyn.}: contains dynamic content.
    \textbf{Depth}: provides ground-truth or LiDAR depth.
    \textbf{Pose}: provides camera/LiDAR extrinsics.
    \textbf{Sem.}: provides semantic / instance labels.
    \textbf{Scale}: approximate number of scenes/objects/sequences.
    \textbf{Primary use}: main application in the NVS literature.}
  \label{tab:datasets}
  \scriptsize
  \setlength{\tabcolsep}{2.5pt}
  \begin{tabular}{@{}llccccccp{3.5cm}@{}}
    \toprule
    \textbf{Dataset} & \textbf{Year} & \textbf{Type}
      & \textbf{Dyn.} & \textbf{Depth} & \textbf{Pose}
      & \textbf{Sem.} & \textbf{Scale}
      & \textbf{Primary Use} \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Object-centric}} \\
    ShapeNet~\citep{chang2015shapenet}               & 2015 & Syn & \no & \yes & \yes & \yes & 51K obj.   & Generative 3-D training \\
    NeRF-Synthetic~\citep{mildenhall2020nerf}        & 2020 & Syn & \no & \yes & \yes & \no  & 8 obj.     & Per-scene NeRF evaluation \\
    Co3Dv2~\citep{reizenstein2021co3d}               & 2021 & RT  & \pt & \no  & \yes & \yes & 19K seq.   & Category-level NVS \\
    Objaverse~\citep{deitke2023objaverse}            & 2023 & Syn & \pt & \yes & \yes & \pt  & 800K obj.  & LRM / diffusion training \\
    Objaverse-XL~\citep{deitke2024objaversexl}       & 2024 & Syn & \pt & \yes & \yes & \pt  & 10M+ obj.  & Large-scale 3-D pretraining \\
    MVImgNet~\citep{yu2023mvimgnet}                  & 2023 & RT  & \no & \no  & \yes & \yes & 238K seq.  & Category generalizable NVS \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Static indoor / outdoor scenes}} \\
    DTU~\citep{jensen2014dtu}                        & 2014 & RT  & \no & \yes & \yes & \no  & 124 scn.   & MVS / NVS evaluation \\
    LLFF~\citep{mildenhall2019llff}                  & 2019 & RT  & \no & \no  & \yes & \no  & 8 scn.     & Forward-facing NeRF \\
    Tanks\&Temples~\citep{knapitsch2017tanks}        & 2017 & RT  & \no & \yes & \yes & \no  & 14 scn.    & Large-scale reconstruction \\
    Mip-NeRF360~\citep{barron2022mipnerf360}         & 2022 & RT  & \no & \no  & \yes & \no  & 9 scn.     & Unbounded 360$^\circ$ NVS \\
    ScanNet~\citep{dai2017scannet}                   & 2017 & RT  & \no & \yes & \yes & \yes & 1513 seq.  & Indoor 3-D reconstruction \\
    ScanNet++~\citep{yeshwanth2023scannetpp}         & 2023 & RT  & \no & \yes & \yes & \yes & 460 scn.   & Generalizable NVS, hi-res \\
    RealEstate10K~\citep{zhou2018stereoMagnification}& 2018 & RT  & \no & \no  & \yes & \no  & 10K vid.   & Generalizable video NVS \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Dynamic scenes}} \\
    D-NeRF~\citep{pumarola2021dnerf}                 & 2021 & Syn & \yes& \no  & \yes & \no  & 6 obj.     & Dynamic NeRF evaluation \\
    HyperNeRF~\citep{park2021hypernerf}              & 2021 & RT  & \yes& \no  & \yes & \no  & 15 seq.    & Topological deformation \\
    CMU Panoptic~\citep{joo2015panoptic}             & 2015 & RT  & \yes& \yes & \yes & \yes & 65 seq.    & Multi-view social activities \\
    Neu3D~\citep{li2022neural3d}                     & 2022 & RT  & \yes& \no  & \yes & \no  & 6 seq.     & Multi-view human video NeRF \\
    iPhone Dataset~\citep{guo2022neural}             & 2022 & RT  & \yes& \no  & \no  & \no  & 14 seq.    & Casually-captured dynamic NVS \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Autonomous driving \& large-scale}} \\
    KITTI~\citep{geiger2013kitti}                    & 2013 & RT  & \yes& \yes & \yes & \yes & Urban      & Early driving benchmark \\
    Waymo Open~\citep{sun2020waymo}                  & 2020 & RT  & \yes& \yes & \yes & \yes & 1000 seg.  & Driving sensor simulation \\
    nuScenes~\citep{caesar2020nuscenes}              & 2020 & RT  & \yes& \pt  & \yes & \yes & 1000 scn.  & 360$^\circ$ driving world model \\
    KITTI-360~\citep{liao2022kitti360}               & 2022 & RT  & \yes& \yes & \yes & \yes & Urban      & Urban scene NVS + panoptic \\
    Argoverse~2~\citep{wilson2023argoverse2}         & 2023 & RT  & \yes& \yes & \yes & \yes & 1000 seq.  & High-res 360$^\circ$ driving \\
    PandaSet~\citep{xiao2021pandaset}                & 2020 & RT  & \yes& \yes & \yes & \yes & 103 scn.   & LiDAR+camera NVS \\
    \bottomrule
  \end{tabular}
\end{table*}

% ─────────────────────────────────────────────────────────────────────────────
