\section{Evaluation Metrics and Protocols}
\label{sec:eval}

Evaluation in \nvs{} and world modeling is more nuanced than in many areas
of computer vision, because the goals of the field span a spectrum from
perceptual realism to geometric accuracy to physical plausibility---and
no single metric captures all three simultaneously.
A systematic understanding of the strengths and limitations of existing
metrics is essential for interpreting the benchmark tables in this survey
and for identifying where reported numbers may be misleading.
We first discuss the standard image-quality metrics that dominate the per-scene
NVS literature, then address the video-quality and task-specific metrics
required for world modeling evaluation, and finally identify the evaluation
gaps that represent open methodological problems.

\subsection{Standard Image Quality Metrics}

The three metrics that appear in virtually all \nvs{} benchmarks measure
different aspects of the relationship between a synthesised novel view
$\hat{\mathbf{I}}$ and a ground-truth photograph $\mathbf{I}^*$.

\textbf{PSNR} (Peak Signal-to-Noise Ratio) measures pixel-level fidelity
as a logarithmic function of mean squared error: $\text{PSNR}=10\log_{10}(1/\text{MSE})$.
High PSNR indicates low pixel-level error, but the metric is insensitive to
structural artifacts and perceptually meaningless differences.
It is nonetheless a standard baseline because it requires no learned components
and is fully deterministic.

\textbf{SSIM}~\citep{wang2004ssim} (Structural Similarity Index) computes local
statistics---mean, variance, and covariance---in sliding windows across the image,
capturing structural coherence beyond pixel-level fidelity.
It correlates better with human perceptual quality than PSNR in the classical
reconstruction regime, but like PSNR it penalises perceptually plausible
hallucinations that happen to be pixel-misaligned with the ground truth.

\textbf{LPIPS}~\citep{zhang2018lpips} (Learned Perceptual Image Patch Similarity)
measures perceptual similarity via deep feature distances, computing the $\ell_2$
distance between activations of a pretrained AlexNet or VGG at multiple layers.
It is the metric most closely correlated with human perceptual judgements and
is therefore the most informative single metric for evaluating photorealistic
NVS quality.
The limitation of LPIPS is its learned nature: performance on out-of-distribution
scenes (e.g., medical imagery, infrared) may not reflect perceptual quality.

Table~\ref{tab:nerf_benchmark} shows representative benchmark results on
NeRF-Synthetic and LLFF across the major method families.

\begin{table*}[t]
  \centering
  \caption{Quantitative NVS comparison on three standard benchmarks.
    NeRF-Synthetic: bounded synthetic objects (8 scenes).
    LLFF: real forward-facing scenes (8 scenes).
    Mip-NeRF360: real unbounded 360$^\circ$ scenes (9 scenes).
    Higher PSNR/SSIM and lower LPIPS are better.
    ``--'' indicates not reported on that benchmark.
    $\dagger$~denotes generalizable methods evaluated without per-scene fine-tuning.}
  \label{tab:nerf_benchmark}
  \scriptsize
  \setlength{\tabcolsep}{2.8pt}
  \begin{tabular}{@{}llccccccccc@{}}
    \toprule
    \multirow{2}{*}{\textbf{Method}} & \multirow{2}{*}{\textbf{Venue}}
      & \multicolumn{3}{c}{\textbf{NeRF-Synthetic}}
      & \multicolumn{3}{c}{\textbf{LLFF}}
      & \multicolumn{3}{c}{\textbf{Mip-NeRF360}} \\
    \cmidrule(lr){3-5}\cmidrule(lr){6-8}\cmidrule(lr){9-11}
      & & PSNR$\uparrow$ & SSIM$\uparrow$ & LPIPS$\downarrow$
        & PSNR$\uparrow$ & SSIM$\uparrow$ & LPIPS$\downarrow$
        & PSNR$\uparrow$ & SSIM$\uparrow$ & LPIPS$\downarrow$ \\
    \midrule
    \multicolumn{11}{@{}l}{\textit{Per-scene NeRF}} \\
    NeRF~\citep{mildenhall2020nerf}             & ECCV'20    & 31.01 & 0.947 & 0.081 & 26.50 & 0.811 & 0.250 & -- & -- & -- \\
    Mip-NeRF~\citep{barron2021mipnerf}          & ICCV'21    & 33.09 & 0.961 & 0.043 & 26.70 & 0.829 & 0.177 & -- & -- & -- \\
    TensoRF~\citep{chen2022tensorf}             & ECCV'22    & 33.14 & 0.963 & 0.047 & 26.73 & 0.839 & 0.204 & -- & -- & -- \\
    Instant-NGP~\citep{muller2022instant}       & TOG'22     & 33.18 & 0.963 & 0.037 & 25.59 & 0.817 & 0.224 & -- & -- & -- \\
    Mip-NeRF360~\citep{barron2022mipnerf360}    & CVPR'22    & --    & --    & --    & 29.40 & 0.901 & 0.127 & 29.40 & 0.840 & 0.204 \\
    Zip-NeRF~\citep{barron2023zipnerf}          & ICCV'23    & --    & --    & --    & 30.38 & 0.921 & 0.100 & 30.38 & 0.866 & 0.179 \\
    \midrule
    \multicolumn{11}{@{}l}{\textit{Per-scene 3DGS}} \\
    3DGS~\citep{kerbl2023gaussian}              & TOG'23     & 33.32 & 0.963 & 0.044 & 27.21 & 0.815 & 0.183 & 27.21 & 0.815 & 0.183 \\
    Mip-Splatting~\citep{yu2024mipsplatting}    & CVPR'24    & 34.66 & 0.973 & 0.019 & 27.79 & 0.827 & 0.163 & 27.92 & 0.842 & 0.158 \\
    Scaffold-GS~\citep{lu2023scaffoldgs}        & CVPR'24    & 33.56 & 0.966 & 0.033 & 27.63 & 0.827 & 0.176 & 27.71 & 0.831 & 0.171 \\
    2DGS~\citep{huang2024_2dgs}                 & SIGGRAPH'24& 33.35 & 0.965 & 0.042 & 26.94 & 0.818 & 0.212 & -- & -- & -- \\
    \midrule
    \multicolumn{11}{@{}l}{\textit{Generalizable (no per-scene training)$\dagger$}} \\
    pixelNeRF~\citep{yu2021pixelnerf}           & CVPR'21    & 23.17 & 0.860 & 0.170 & -- & -- & -- & -- & -- & -- \\
    IBRNet~\citep{wang2021ibrnet}               & CVPR'21    & 26.73 & 0.917 & 0.091 & 26.73 & 0.851 & 0.176 & -- & -- & -- \\
    MVSNeRF~\citep{chen2021mvsnerf}             & ICCV'21    & 26.63 & 0.931 & 0.168 & 26.40 & 0.839 & 0.181 & -- & -- & -- \\
    pixelSplat~\citep{charatan2024pixelsplat}   & CVPR'24    & -- & -- & -- & 26.31 & 0.815 & 0.200 & -- & -- & -- \\
    MVSplat~\citep{chen2024mvsplat}             & ECCV'24    & -- & -- & -- & 26.40 & 0.823 & 0.188 & -- & -- & -- \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Video and World Modelling Metrics}

The transition to world modeling requires metrics that evaluate temporal
and dynamic properties in addition to per-frame image quality.
Single-frame quality metrics are insufficient for this purpose because they
do not measure whether adjacent frames form a geometrically and temporally
coherent sequence.

\textbf{FVD}~\citep{unterthiner2019fvd} (Fréchet Video Distance) measures
the distributional distance between generated and real video sequences in
the feature space of a pretrained I3D network, capturing both spatial quality
and temporal dynamics.
It is the video-generation analogue of FID~\citep{heusel2017fid} and has
become the primary metric for evaluating world model generation quality.
Its limitation is that it measures distributional quality rather than
per-instance fidelity---a model that generates diverse but geometrically
inconsistent sequences can achieve a good FVD score.

\textbf{Temporal consistency} is typically measured as the average cosine
similarity of DINO~\citep{} features between consecutive frames, penalising
large changes in feature space that correspond to flickering or frame-level
inconsistency.
This metric is sensitive to the temporal coherence that is most directly
relevant to world modeling but does not capture geometric 3-D consistency
across viewpoints.

\textbf{Task-specific evaluation} is the most informative metric for world
models but also the most expensive to compute.
For autonomous driving world models, the key task-specific metrics are
closed-loop collision rate, route completion ratio, and imitation gap
(the difference between the model and an expert policy in a closed-loop setting).
For robotics, task success rate and generalisation to novel configurations
are the primary metrics.
These task-specific metrics provide the most direct evidence that a world
model is learning semantically useful scene dynamics rather than merely
reproducing visual appearance statistics.

\subsection{Limitations of Current Evaluation}

The gap between current evaluation protocols and the requirements of world
modeling represents a major methodological problem that the field has
not yet fully addressed.

PSNR and SSIM penalise perceptually plausible but pixel-misaligned generations,
particularly under large viewpoint changes where the reference content
provides little direct evidence for the target-view appearance.
This creates a systematic bias against generative methods, which often hallucinate
content that is physically plausible but does not match the ground-truth pixel
values used for evaluation.
FVD captures temporal dynamics but not geometric accuracy, allowing methods
that generate temporally smooth but spatially inconsistent videos to achieve
high scores.
As a result, no existing benchmark jointly evaluates geometry, dynamics, action
responsiveness, and physical plausibility---the full set of world-modeling
desiderata.
This evaluation gap is itself a major open problem.

\paragraph{Towards systematic evaluation.}
The work on Systematic Evaluation of NVS~\citep{chen2024sysnvs} provides
a principled analysis of how existing benchmarks differ and proposes
controlled evaluation protocols separating appearance and geometry quality.
Non-Aligned Reference IQA~\citep{zhang2024nariqanvs} addresses the fundamental
problem of quality assessment when reference and synthesised views are not
pixel-aligned, proposing a learned non-aligned metric for NVS that is
better suited to the large-viewpoint-change regime where generative methods
operate.
VGGT-X~\citep{liu2025vggtx} extends VGGT to dense NVS, providing strong
baselines for the emerging class of feed-forward 3-D foundation models
and enabling consistent comparison across the generalised reconstruction
and generative NVS families.
We propose in \S\ref{sec:open} a five-axis world-modeling benchmark that
would provide the systematic, multi-dimensional evaluation that the field
currently lacks.

% ─────────────────────────────────────────────────────────────────────────────
