\section{Conclusion}
\label{sec:conclusion}

We have surveyed Novel View Synthesis through the lens of Generative World Modeling,
tracing the evolution from classical image-based rendering through NeRF,
3D Gaussian Splatting, feed-forward reconstruction, and diffusion-based synthesis
toward 4-D world generation.
Our unified continuous-spectrum taxonomy organises more than 160 methods across
three axes (input regime, temporal scope, controllability) and makes explicit
their world-modelling implications.

The field is at an inflection point.
Spatial grounding from NVS and temporal/action capabilities from world modeling
are actively converging, as evidenced by systems such as Genie~2, Cosmos,
ViewCrafter, and the neural driving simulators.
Closing the geometry--generation trade-off, achieving long-horizon coherence,
and grounding predictions in physical laws are the primary challenges ahead.

\medskip
\noindent\textit{Draft v0.9---ongoing work.
Content will be further expanded in subsequent versions.}

% ─────────────────────────────────────────────────────────────────────────────
\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem{levoy1996light}
M.~Levoy and P.~Hanrahan.
Light field rendering.
\emph{SIGGRAPH}, 1996.


\bibitem{mcmillan1995plenoptic}
L.~McMillan and G.~Bishop.
Plenoptic modeling.
\emph{SIGGRAPH}, 1995.


\bibitem{chen1993view}
S.~E.~Chen and L.~Williams.
View interpolation for image synthesis.
\emph{SIGGRAPH}, 1993.


\bibitem{debevec1996modeling}
P.~E.~Debevec, C.~J.~Taylor, and J.~Malik.
Modeling and rendering architecture from photographs.
\emph{SIGGRAPH}, 1996.


\bibitem{gortler1996lumigraph}
S.~J.~Gortler, R.~Grzeszczuk, R.~Szeliski, and M.~F.~Cohen.
The lumigraph.
\emph{SIGGRAPH}, 1996.


\bibitem{adelson1991plenoptic}
E.~H.~Adelson and J.~R.~Bergen.
The plenoptic function and the elements of early vision.
In \emph{Computational Models of Visual Processing}, 1991.


\bibitem{schoenberger2016sfm}
J.~L.~Sch\"{o}nberger and J.-M.~Frahm.
Structure-from-motion revisited.
\emph{CVPR}, 2016.


\bibitem{schonberger2016mvs}
J.~L.~Sch\"{o}nberger, E.~Zheng, J.-M.~Frahm, and M.~Pollefeys.
Pixelwise view selection for unstructured multi-view stereo.
\emph{ECCV}, 2016.

% ── Related Surveys ────────────────────────────────────────────────────────

\bibitem{tewari2022advances}
A.~Tewari \etal
Advances in neural rendering.
\emph{Computer Graphics Forum (CGF)}, 41(2):703--735, 2022.


\bibitem{gao2022nerf_survey}
K.~Gao \etal
NeRF: Neural radiance field in 3D vision---a comprehensive review.
\emph{arXiv:2210.00379}, 2022.


\bibitem{zhu2023survey_ibr}
T.~Zhu \etal
A survey on image-based rendering.
\emph{arXiv}, 2023.


\bibitem{wu2024recentGS}
T.~Wu \etal
Recent advances in 3D Gaussian Splatting.
\emph{arXiv:2403.11134}, 2024.


\bibitem{fei2024gssurvey}
B.~Fei \etal
3D Gaussian Splatting as a new era of novel view synthesis.
\emph{IEEE TVCG}, 2024.


\bibitem{chen2024survey_drivingNVS}
K.~Chen \etal
A survey of novel view synthesis from autonomous driving perspective.
\emph{arXiv}, 2024.


\bibitem{zhang2025feedforward_survey}
J.~Zhang \etal
Advances in feed-forward 3D reconstruction and view synthesis: A survey.
\emph{arXiv}, 2025.


\bibitem{he2025gs_apps}
S.~He \etal
A survey on 3D Gaussian Splatting applications.
\emph{arXiv:2308.04079}, 2025.

% ── NeRF Foundational ──────────────────────────────────────────────────────

\bibitem{mildenhall2020nerf}
B.~Mildenhall, P.~P.~Srinivasan, M.~Tancik, J.~T.~Barron, R.~Ramamoorthi,
and R.~Ng.
NeRF: Representing scenes as neural radiance fields for view synthesis.
\emph{ECCV}, 2020.


\bibitem{martinbrualla2021nerfw}
R.~Martin-Brualla, N.~Radwan, M.~S.~M.~Sajjadi, J.~T.~Barron, A.~Dosovitskiy,
and D.~Duckworth.
NeRF in the Wild: Neural radiance fields for unconstrained photo collections.
\emph{CVPR}, 2021.


\bibitem{barron2021mipnerf}
J.~T.~Barron, B.~Mildenhall, M.~Tancik, P.~Hedman, R.~Martin-Brualla, and
P.~P.~Srinivasan.
Mip-NeRF: A multiscale representation for anti-aliasing neural radiance fields.
\emph{ICCV}, 2021.


\bibitem{barron2022mipnerf360}
J.~T.~Barron, B.~Mildenhall, D.~Verbin, P.~P.~Srinivasan, and P.~Hedman.
Mip-NeRF 360: Unbounded anti-aliased neural radiance fields.
\emph{CVPR}, 2022.


\bibitem{barron2023zipnerf}
J.~T.~Barron, B.~Mildenhall, D.~Verbin, P.~P.~Srinivasan, and P.~Hedman.
Zip-NeRF: Anti-aliased grid-based neural radiance fields.
\emph{ICCV}, 2023.


\bibitem{verbin2022refnerf}
D.~Verbin, P.~Hedman, B.~Mildenhall, T.~Zickler, J.~T.~Barron, and
P.~P.~Srinivasan.
Ref-NeRF: Structured view-dependent appearance for neural radiance fields.
\emph{CVPR}, 2022.


\bibitem{muller2022instant}
T.~M\"{u}ller, A.~Evans, C.~Schied, and A.~Keller.
Instant neural graphics primitives with a multiresolution hash encoding.
\emph{ACM TOG}, 2022.


\bibitem{chen2022tensorf}
A.~Chen, Z.~Xu, A.~Geiger, J.~Yu, and H.~Su.
TensoRF: Tensorial radiance fields.
\emph{ECCV}, 2022.


\bibitem{fridovich2022plenoxels}
S.~Fridovich-Keil, A.~Yu, M.~Tancik, Q.~Chen, B.~Recht, and A.~Kanazawa.
Plenoxels: Radiance fields without neural networks.
\emph{CVPR}, 2022.


\bibitem{sun2022dvgo}
C.~Sun, M.~Sun, and H.-T.~Chen.
Direct voxel grid optimization: Super-fast convergence for radiance fields
reconstruction.
\emph{CVPR}, 2022.


\bibitem{wang2021neus}
P.~Wang, L.~Liu, Y.~Liu, C.~Theobalt, T.~Komura, and W.~Wang.
NeuS: Learning neural implicit surfaces by volume rendering for multi-view
reconstruction.
\emph{NeurIPS}, 2021.


\bibitem{yariv2021volsdf}
L.~Yariv, J.~Gu, Y.~Kasten, and Y.~Lipman.
Volume rendering of neural implicit surfaces.
\emph{NeurIPS}, 2021.


\bibitem{yariv2023bakedsdf}
L.~Yariv, P.~Hedman, C.~Reiser, D.~Verbin, P.~P.~Srinivasan, R.~Bastian-Smith,
J.~T.~Barron, and R.~Martin-Brualla.
BakedSDF: Meshing neural SDFs for real-time view synthesis.
\emph{SIGGRAPH}, 2023.

% ── Generalizable NeRF ─────────────────────────────────────────────────────

\bibitem{yu2021pixelnerf}
A.~Yu, V.~Ye, M.~Tancik, and A.~Kanazawa.
pixelNeRF: Neural radiance fields from one or few images.
\emph{CVPR}, 2021.


\bibitem{wang2021ibrnet}
Q.~Wang, Z.~Wang, K.~Genova, P.~P.~Srinivasan, H.~Zhou, J.~T.~Barron,
R.~Martin-Brualla, N.~Snavely, and T.~Funkhouser.
IBRNet: Learning multi-view image-based rendering.
\emph{CVPR}, 2021.


\bibitem{chen2021mvsnerf}
A.~Chen, Z.~Xu, F.~Zhao, X.~Zhang, F.~Xiang, J.~Yu, and H.~Su.
MVSNeRF: Fast generalizable radiance field reconstruction from multi-view
stereo.
\emph{ICCV}, 2021.


\bibitem{wang2022gnt}
S.~Wang, T.~Leimk\"{u}hler, and C.~Theobalt.
Generalizable novel view synthesis via neural radiance field transformers.
\emph{arXiv}, 2022.


\bibitem{sajjadi2022srt}
M.~S.~M.~Sajjadi, H.~Meyer, E.~Pot, U.~Bergmann, K.~Greff, N.~Radwan,
S.~Vora, and M.~Hessel.
Scene representation transformer: Geometry-free novel view synthesis through
set-latent scene representations.
\emph{CVPR}, 2022.


\bibitem{gu2023nerfdiff}
J.~Gu \etal
NerfDiff: Single-image view synthesis with nerf-guided distillation.
\emph{ICML}, 2023.


\bibitem{shi2023zerorf}
R.~Shi, X.~Wei, C.~Wang, and H.~Su.
ZeroRF: Fast sparse view 360$^\circ$ reconstruction with zero pretraining.
\emph{arXiv:2312.09967}, 2023.

% ── Dynamic NeRF ───────────────────────────────────────────────────────────

\bibitem{pumarola2021dnerf}
A.~Pumarola, E.~Corona, G.~Pons-Moll, and F.~Moreno-Noguer.
D-NeRF: Neural radiance fields for dynamic scenes.
\emph{CVPR}, 2021.


\bibitem{li2021neural_scene_flow}
Z.~Li, S.~Niklaus, N.~Snavely, and O.~Wang.
Neural scene flow fields for space-time view synthesis of dynamic scenes.
\emph{CVPR}, 2021.


\bibitem{park2021nerfies}
K.~Park, U.~Sinha, J.~T.~Barron, S.~Bouaziz, D.~B.~Goldman, S.~M.~Seitz,
and R.~Martin-Brualla.
Nerfies: Deformable neural radiance fields.
\emph{ICCV}, 2021.


\bibitem{park2021hypernerf}
K.~Park, U.~Sinha, P.~Hedman, J.~T.~Barron, S.~Bouaziz, D.~B.~Goldman,
R.~Martin-Brualla, and S.~M.~Seitz.
HyperNeRF: A higher-dimensional representation for topologically varying neural
radiance fields.
\emph{ACM TOG}, 2021.


\bibitem{li2022neural3d}
Z.~Li, S.~Niklaus, N.~Snavely, and O.~Wang.
Neural 3D video synthesis from multi-view video.
\emph{CVPR}, 2022.


\bibitem{fang2022tineuvox}
J.~Fang, T.~Yi, X.~Wang, L.~Xie, X.~Zhang, W.~Liu, M.~Ni\"{e}{\ss}ner,
and L.~Tian.
Fast dynamic radiance fields with time-aware neural voxels.
\emph{SIGGRAPH Asia}, 2022.


\bibitem{fridovich2023kplanes}
S.~Fridovich-Keil, G.~Meanti, F.~R.~Warburg, B.~Recht, and A.~Kanazawa.
K-Planes: Explicit radiance fields in space, time, and appearance.
\emph{CVPR}, 2023.


\bibitem{cao2023hexplane}
A.~Cao and J.~Johnson.
HexPlane: A fast representation for dynamic scenes.
\emph{CVPR}, 2023.


\bibitem{song2023nerfplayer}
L.~Song \etal
NeRFPlayer: A streamable dynamic scene representation with decomposed neural
radiance fields.
\emph{IEEE TVCG}, 2023.

% ── 3DGS Foundational ──────────────────────────────────────────────────────

\bibitem{kerbl2023gaussian}
B.~Kerbl, G.~Kopanas, T.~Leimk\"{u}hler, and G.~Drettakis.
3D Gaussian Splatting for real-time radiance field rendering.
\emph{ACM TOG}, 2023.


\bibitem{yu2024mipsplatting}
Z.~Yu, A.~Chen, B.~Huang, T.~Sattler, and A.~Geiger.
Mip-Splatting: Alias-free 3D Gaussian Splatting.
\emph{CVPR}, 2024.


\bibitem{ye2024absgs}
Z.~Ye, Z.~Liu, W.~Song, L.~Liu, X.~Zhao, T.~Zhu, and B.~Zhang.
AbsGS: Recovering fine details for 3D Gaussian Splatting.
\emph{NeurIPS}, 2024.


\bibitem{lu2023scaffoldgs}
T.~Lu \etal
Scaffold-GS: Structured 3D Gaussians for view-adaptive rendering.
\emph{CVPR}, 2024.


\bibitem{fan2023lightgaussian}
Z.~Fan, K.~Wang, K.~Wen, Z.~Zhu, D.~Xu, and Z.~Wang.
LightGaussian: Unbounded 3D Gaussian compression with 15$\times$ reduction.
\emph{NeurIPS}, 2023.


\bibitem{lee2024compact3d}
J.~Lee \etal
Compact 3D Gaussian representation for radiance field.
\emph{arXiv:2311.13681}, 2024.


\bibitem{girish2024eagles}
S.~Girish, K.~Gupta, and A.~Shrivastava.
EAGLES: Efficient accelerated 3D Gaussians with lightweight encodings.
\emph{ECCV}, 2024.


\bibitem{chen2024hac}
Y.~Chen, L.~Wu, C.~Gao, J.~Li, and C.~Liu.
HAC: Hash-grid assisted context for 3D Gaussian Splatting compression.
\emph{ECCV}, 2024.


\bibitem{guedon2024sugar}
A.~Gu\'{e}don and V.~Lepetit.
SuGaR: Surface-aligned Gaussian Splatting for efficient 3D mesh reconstruction
and high-quality mesh rendering.
\emph{CVPR}, 2024.


\bibitem{huang2024_2dgs}
B.-S.~Huang, Z.~Yu, A.~Chen, A.~Geiger, and S.~Gao.
2D Gaussian Splatting for geometrically accurate radiance fields.
\emph{SIGGRAPH}, 2024.


\bibitem{yu2024gof}
Z.~Yu, T.~Sattler, and A.~Geiger.
Gaussian Opacity Fields: Efficient adaptive surface reconstruction in unbounded scenes.
\emph{arXiv:2404.10772}, 2024.


\bibitem{chen2024gaussianeditor}
Y.~Chen, Z.~Chen, C.~Zhang, F.~Wang, X.~Yang, Y.~Wang, Z.~Cai, L.~Yang,
H.~Liu, and G.~Lin.
GaussianEditor: Swift and controllable 3D editing with Gaussian Splatting.
\emph{CVPR}, 2024.


\bibitem{ye2023gaussian_grouping}
M.~Ye, M.~Danelljan, F.~Yu, and L.~Ke.
Gaussian Grouping: Segment and edit anything in 3D scenes.
\emph{ECCV}, 2024.


\bibitem{zhou2024gala3d}
X.~Zhou \etal
GALA3D: Towards text-to-3D complex scene generation via layout-guided
generative Gaussian Splatting.
\emph{ICML}, 2024.

% ── Dynamic 3DGS ───────────────────────────────────────────────────────────

\bibitem{wu2024_4dgs}
G.~Wu, T.~Yi, J.~Fang, L.~Xie, X.~Zhang, W.~Wei, W.~Liu, Q.~Tian, and
M.~Wang.
4D Gaussian Splatting for real-time dynamic scene rendering.
\emph{CVPR}, 2024.


\bibitem{yang2024spacetime}
Z.~Yang, X.~Gao, Y.~Sun, Y.~Huang, X.~Lyu, W.~Zhou, X.~Jiao, X.~Qi,
Y.~Jin, D.~Yang, X.~Duan, F.~Yu, L.~Liu, P.~Zhao, and A.~Qu.
Deformable 3D Gaussians for high-fidelity monocular dynamic reconstruction.
\emph{CVPR}, 2024.


\bibitem{yang2024deformable}
Z.~Yang, X.~Gao, W.~Zhou, S.~Jiao, Y.~Zhang, and X.~Jin.
Real-time photorealistic dynamic scene representation and rendering with 4D
Gaussian Splatting.
\emph{ICLR}, 2024.


\bibitem{huang2024scgs}
Y.-H.~Huang, Y.-T.~Sun, Z.~Yang, X.~Lyu, Y.-P.~Cao, and X.~Qi.
SC-GS: Sparse-controlled Gaussian Splatting for editable dynamic scenes.
\emph{CVPR}, 2024.


\bibitem{lin2024gaussianflow}
Q.~Lin \etal
Gaussian-Flow: 4D reconstruction with dynamic 3D Gaussian particle.
\emph{CVPR}, 2024.


\bibitem{wang2024shapeofmotion}
Q.~Wang \etal
Shape of Motion: 4D reconstruction from a single video.
\emph{arXiv:2407.13764}, 2024.

% ── Feed-Forward Generalizable 3DGS ────────────────────────────────────────

\bibitem{charatan2024pixelsplat}
D.~Charatan, S.~Li, A.~Tagliasacchi, and V.~Sitzmann.
pixelSplat: 3D Gaussian splats from image pairs for scalable generalizable
3D reconstruction.
\emph{CVPR}, 2024.


\bibitem{chen2024mvsplat}
Y.~Chen \etal
MVSplat: Efficient 3D Gaussian Splatting from sparse multi-view images.
\emph{ECCV}, 2024.


\bibitem{zhang2024transplate}
C.~Zhang \etal
TranSplat: Generalizable 3D Gaussian Splatting from sparse multi-view images
with transformers.
\emph{arXiv:2408.13912}, 2024.


\bibitem{zhu2024fsgs}
Z.~Zhu, Z.~Fan, Y.~Jiang, and Z.~Wang.
FSGS: Real-time few-shot view synthesis using Gaussian Splatting.
\emph{CVPR}, 2024.


\bibitem{xu2024depthsplat}
H.~Xu \etal
DepthSplat: Connecting Gaussian Splatting and depth.
\emph{arXiv}, 2024.


\bibitem{smart2024splatt3r}
B.~Smart \etal
Splatt3R: Zero-shot Gaussian splatting from uncalibrated image pairs.
\emph{arXiv:2408.13912}, 2024.


\bibitem{zou2024triplanegaussian}
Z.-X.~Zou, Z.~Yu, Y.-B.~Guo, Y.~Li, L.~Liang, C.-J.~Cao, and Y.~Zhang.
Triplane meets Gaussian Splatting: Fast and generalizable single-view 3D
reconstruction with transformers.
\emph{CVPR}, 2024.

% ── Pose-Free and Foundation Models ────────────────────────────────────────

\bibitem{wang2024dust3r}
S.~Wang \etal
DUSt3R: Geometric 3D vision made easy.
\emph{CVPR}, 2024.


\bibitem{leroy2024mast3r}
V.~Leroy \etal
MASt3R: Grounding Image Matching in 3D with MASt3R.
\emph{ECCV}, 2024.


\bibitem{zhang2024monst3r}
J.~Zhang \etal
MonST3R: A simple approach for estimating geometry in the presence of motion.
\emph{arXiv:2410.03825}, 2024.


\bibitem{wang2025vggt}
H.~Wang \etal
VGGT: Visual geometry grounded deep structured tokens.
\emph{arXiv:2503.11651}, 2025.

% ── LRM ────────────────────────────────────────────────────────────────────

\bibitem{hong2023lrm}
Y.~Hong, K.~Zhang, J.~Gu, S.~Bi, Y.~Zhou, D.~Liu, F.~Liu, K.~Sunkavalli,
T.~Bui, and H.~Tan.
LRM: Large reconstruction model for single image to 3D.
\emph{ICLR}, 2024.


\bibitem{li2023instant3d}
J.~Li \etal
Instant3D: Fast text-to-3D with sparse-view generation and large reconstruction model.
\emph{ICLR}, 2024.


\bibitem{openlrm2023}
Z.~He and H.~Wang.
OpenLRM: Open-source large reconstruction models.
\emph{arXiv:2311.04400}, 2023.


\bibitem{tang2024lgm}
J.~Tang \etal
LGM: Large multi-view Gaussian model for high-resolution 3D content creation.
\emph{ECCV}, 2024.


\bibitem{wang2024crm}
Z.~Wang \etal
CRM: Single image to 3D textured mesh with convolutional reconstruction model.
\emph{NeurIPS}, 2024.

% ── MVS ────────────────────────────────────────────────────────────────────

\bibitem{yao2018mvsnet}
Y.~Yao, Z.~Luo, S.~Li, T.~Fang, and L.~Quan.
MVSNet: Depth inference for unstructured multi-view stereo.
\emph{ECCV}, 2018.


\bibitem{wang2022itermvs}
F.~Wang \etal
IterMVS: Iterative probability estimation for efficient multi-view stereo.
\emph{CVPR}, 2022.


\bibitem{peng2022unimvsnet}
R.~Peng \etal
Rethinking depth estimation for multi-view stereo: A unified representation.
\emph{CVPR}, 2022.

% ── 3-D-Aware GANs ─────────────────────────────────────────────────────────

\bibitem{schwarz2020graf}
K.~Schwarz, Y.~Liao, M.~Niemeyer, and A.~Geiger.
GRAF: Generative radiance fields for 3D-aware image synthesis.
\emph{NeurIPS}, 2020.


\bibitem{chan2021pigan}
E.~R.~Chan, M.~Monteiro, P.~Kellnhofer, J.~Wu, and G.~Wetzstein.
pi-GAN: Periodic implicit generative adversarial networks for 3D-aware image
synthesis.
\emph{CVPR}, 2021.


\bibitem{niemeyer2021giraffe}
M.~Niemeyer and A.~Geiger.
GIRAFFE: Representing scenes as compositional generative neural feature fields.
\emph{CVPR}, 2021.


\bibitem{chan2022efficient}
E.~R.~Chan \etal
Efficient geometry-aware 3D generative adversarial networks.
\emph{CVPR}, 2022.


\bibitem{gu2022stylenerf}
J.~Gu, L.~Liu, P.~Wang, and C.~Theobalt.
StyleNeRF: A style-based 3D-aware generator for high-resolution image synthesis.
\emph{ICLR}, 2022.

% ── Diffusion-Based NVS ────────────────────────────────────────────────────

\bibitem{rombach2022ldm}
R.~Rombach, A.~Blattmann, D.~Lorenz, P.~Esser, and B.~Ommer.
High-resolution image synthesis with latent diffusion models.
\emph{CVPR}, 2022.


\bibitem{liu2023zero1to3}
R.~Liu, R.~Wu, B.~V.~Hoorick, P.~Tokmakov, S.~Zakharov, and C.~Vondrick.
Zero-1-to-3: Zero-shot one image to 3D object.
\emph{ICCV}, 2023.


\bibitem{liu2023syncdreamer}
Y.~Liu, C.~Lin, Z.~Zeng, X.~Long, L.~Liu, T.~Komura, and W.~Wang.
SyncDreamer: Generating multiview-consistent images from a single-view image.
\emph{arXiv:2309.03453}, 2023.


\bibitem{liu2023one2345}
M.~Liu \etal
One-2-3-45: Any single image to 3D mesh in 45 seconds without per-shape
optimization.
\emph{NeurIPS}, 2023.


\bibitem{shi2023zero123pp}
R.~Shi \etal
Zero123++: A single image to consistent multi-view diffusion base model.
\emph{arXiv:2310.15110}, 2023.


\bibitem{tang2023mvdiffusion}
S.~Tang, F.~Zhang, J.~Chen, P.~Wang, and Y.~Furukawa.
MVDiffusion: Enabling holistic multi-view image generation with correspondence-aware diffusion.
\emph{NeurIPS}, 2023.


\bibitem{long2024wonder3d}
H.~Long \etal
Wonder3D: Single image to 3D using cross-domain diffusion.
\emph{CVPR}, 2024.


\bibitem{kong2024eschernet}
X.~Kong \etal
EscherNet: A generative model for scalable view synthesis.
\emph{CVPR}, 2024.


\bibitem{sargent2023zeronvs}
K.~Sargent \etal
ZeroNVS: Zero-shot 360-degree view synthesis from a single real image.
\emph{CVPR}, 2024.


\bibitem{zheng2023free3d}
C.~Zheng \etal
Free3D: Consistent novel view synthesis without 3D representation.
\emph{ICLR}, 2024.


\bibitem{chan2023genvs}
E.~R.~Chan \etal
Generative novel view synthesis with 3D-aware diffusion models.
\emph{ICCV}, 2023.


\bibitem{shi2023mvdream}
Y.~Shi \etal
MVDream: Multi-view diffusion for 3D generation.
\emph{ICLR}, 2024.

% ── SDS / Text-to-3D ───────────────────────────────────────────────────────

\bibitem{poole2023dreamfusion}
B.~Poole, A.~Jain, J.~T.~Barron, and B.~Mildenhall.
DreamFusion: Text-to-3D using 2D diffusion.
\emph{ICLR}, 2023.


\bibitem{lin2023magic3d}
C.-H.~Lin, J.~Gao, L.~Tang, T.~Takikawa, X.~Zeng, X.~Huang, K.~Kreis,
S.~Fidler, M.-Y.~Liu, and T.-Y.~Lin.
Magic3D: High-resolution text-to-3D content creation.
\emph{CVPR}, 2023.


\bibitem{wang2023prolificdreamer}
Z.~Wang \etal
ProlificDreamer: High-fidelity and diverse text-to-3D generation with
variational score distillation.
\emph{NeurIPS}, 2023.


\bibitem{chen2023fantasia3d}
R.~Chen \etal
Fantasia3D: Disentangling geometry and appearance for high-quality text-to-3D
content creation.
\emph{ICCV}, 2023.


\bibitem{yi2024gaussiandreamer}
T.~Yi \etal
GaussianDreamer: Fast generation from text to 3D Gaussians by bridging 2D
and 3D diffusion models.
\emph{CVPR}, 2024.


\bibitem{chung2023luciddreamer}
J.~Chung \etal
LucidDreamer: Domain-free generation of 3D Gaussian Splatting scenes.
\emph{arXiv:2311.13384}, 2023.

% ── Reconstruction-Diffusion Hybrids ───────────────────────────────────────

\bibitem{gao2024cat3d}
R.~Gao \etal
CAT3D: Create anything in 3D with multi-view diffusion models.
\emph{arXiv:2405.10314}, 2024.


\bibitem{wu2024reconfusion}
R.~Wu \etal
ReconFusion: 3D reconstruction with diffusion priors.
\emph{CVPR}, 2024.

% ── 4-D and Video Diffusion ────────────────────────────────────────────────

\bibitem{yu2024viewcrafter}
W.~Yu \etal
ViewCrafter: Taming video diffusion models for high-fidelity novel view synthesis.
\emph{arXiv:2409.02048}, 2024.


\bibitem{watson2024_4dim}
D.~Watson \etal
4DiM: 4D diffusion for dynamic novel view synthesis with global-local
attention.
\emph{arXiv:2311.17905}, 2024.


\bibitem{xie2024sv4d}
Y.~Xie \etal
SV4D: Dynamic 3D content generation with multi-frame and multi-view
consistency.
\emph{arXiv:2407.09787}, 2024.


\bibitem{jiang2024consistent4d}
Y.~Jiang \etal
Consistent4D: Consistent 360$^\circ$ dynamic object generation from monocular
video.
\emph{ICLR}, 2024.


\bibitem{ren2023dreamgaussian4d}
Y.~Ren \etal
DreamGaussian4D: Generative 4D Gaussian Splatting.
\emph{arXiv:2312.17142}, 2023.

% ── World Models ───────────────────────────────────────────────────────────

\bibitem{ha2018worldmodels}
D.~Ha and J.~Schmidhuber.
World models.
\emph{arXiv:1803.10122}, 2018.


\bibitem{lecun2022path}
Y.~LeCun.
A path towards autonomous machine intelligence.
\emph{OpenReview}, 2022.


\bibitem{hu2023gaia}
A.~Hu \etal
GAIA-1: A generative world model for autonomous driving.
\emph{arXiv:2309.17080}, 2023.


\bibitem{yang2023unisim}
Z.~Yang, Y.~Chen, J.~Wang, S.~Manivasagam, W.-C.~Ma, A.~K.~Yang, and
R.~Urtasun.
UniSim: A neural closed-loop sensor simulator.
\emph{CVPR}, 2023.


\bibitem{wu2023mars}
X.~Wu \etal
MARS: An instance-aware, modular and realistic simulator for autonomous
driving.
\emph{CICAI}, 2023.


\bibitem{chen2024omnire}
Z.~Chen \etal
OmniRe: Omni urban scene reconstruction.
\emph{arXiv:2408.16760}, 2024.


\bibitem{yang2023emernerf}
J.~Yang \etal
EmerNeRF: Emergent spatial-temporal scene decomposition via self-supervision.
\emph{ICLR}, 2024.


\bibitem{yan2024streetgaussian}
Y.~Yan \etal
Street Gaussians: Modelling dynamic urban scenes with Gaussian Splatting.
\emph{ECCV}, 2024.


\bibitem{chen2023pvg}
Y.~Chen \etal
Periodic vibration Gaussian: Dynamic urban scene reconstruction and real-time
rendering.
\emph{arXiv:2311.18561}, 2023.


\bibitem{bruce2024genie}
J.~Bruce \etal
Genie: Generative interactive environments.
\emph{ICML}, 2024.


\bibitem{parkerholder2024genie2}
J.~Parker-Holder \etal
Genie 2: A large-scale foundation world model.
\emph{Google DeepMind Technical Report}, 2024.


\bibitem{agarwal2025cosmos}
N.~Agarwal \etal
Cosmos world foundation model technical report.
\emph{arXiv:2501.03575}, 2025.


\bibitem{xiang2024pandora}
J.~Xiang \etal
Pandora: Towards general world model with natural language actions and video
states.
\emph{arXiv:2406.09455}, 2024.


\bibitem{yu2023wonderjourney}
H.~Yu \etal
WonderJourney: Going from anywhere to everywhere.
\emph{CVPR}, 2024.


\bibitem{yu2024wonderworld}
H.~Yu \etal
WonderWorld: Interactive 3D scene generation using a single image.
\emph{arXiv:2406.09379}, 2024.

% ── Embodied AI ────────────────────────────────────────────────────────────

\bibitem{ichnowski2021dex}
J.~Ichnowski, Y.~Avigal, J.~Kerr, and K.~Goldberg.
Dex-NeRF: Using a neural radiance field to grasp transparent objects.
\emph{CoRL}, 2021.


\bibitem{yen2021inerf}
L.-Y.~Yen-Chen, P.~Florence, J.~T.~Barron, A.~Rodriguez, P.~Isola, and
T.-Y.~Lin.
iNeRF: Inverting neural radiance fields for pose estimation.
\emph{IROS}, 2021.


\bibitem{ze2023nerf_supervisor}
Y.~Ze \etal
NeRF-Supervisor: Learning dense robot manipulation policies with neural
radiance fields.
\emph{ICRA}, 2023.


\bibitem{shorinwa2024gaussiangrasping}
O.~Shorinwa \etal
Splat-MOVER: Multi-stage, open-vocabulary robotic manipulation via editable
Gaussian splatting.
\emph{arXiv}, 2024.


\bibitem{li2024robogssim}
K.~Li \etal
RoboGSim: A real2sim2real robotic Gaussian splatting simulator.
\emph{arXiv:2411.11839}, 2024.

% ── Open-World Semantic Grounding ──────────────────────────────────────────

\bibitem{kerr2023lerf}
J.~Kerr, C.~M.~Kim, K.~Goldberg, A.~Kanazawa, and M.~Tancik.
LERF: Language embedded radiance fields.
\emph{ICCV}, 2023.


\bibitem{qin2024langsplat}
M.~Qin \etal
LangSplat: 3D language Gaussian Splatting.
\emph{CVPR}, 2024.


\bibitem{zhou2024feature3dgs}
S.~Zhou \etal
Feature 3DGS: Supercharging 3D Gaussian Splatting to enable distilled feature fields.
\emph{CVPR}, 2024.

% ── Physics ────────────────────────────────────────────────────────────────

\bibitem{li2023pac}
Y.~Li \etal
PAC-NeRF: Physics augmented continuum neural radiance fields for geometry-agnostic
system identification.
\emph{ICLR}, 2023.

% ── Datasets ───────────────────────────────────────────────────────────────

\bibitem{mildenhall2019llff}
B.~Mildenhall, P.~P.~Srinivasan, R.~Ortiz-Cayon, N.~K.~Kalantari,
R.~Ramamoorthi, R.~Ng, and A.~Kar.
Local light field fusion: Practical view synthesis with prescriptive sampling
guidelines.
\emph{ACM TOG}, 2019.


\bibitem{jensen2014dtu}
R.~Jensen, A.~Dahl, G.~Vogiatzis, E.~Tola, and H.~Aan{\ae}s.
Large scale multi-view stereopsis evaluation.
\emph{CVPR}, 2014.


\bibitem{knapitsch2017tanks}
A.~Knapitsch, J.~Park, Q.-Y.~Zhou, and V.~Koltun.
Tanks and temples: Benchmarking large-scale scene reconstruction.
\emph{ACM TOG}, 2017.


\bibitem{dai2017scannet}
A.~Dai, A.~X.~Chang, M.~Savva, M.~Halber, T.~Funkhouser, and M.~Nie{\ss}ner.
ScanNet: Richly-annotated 3D reconstructions of indoor scenes.
\emph{CVPR}, 2017.


\bibitem{yeshwanth2023scannetpp}
C.~Yeshwanth, Y.-C.~Liu, M.~Nie{\ss}ner, and A.~Dai.
ScanNet++: A high-fidelity dataset of 3D indoor scenes.
\emph{ICCV}, 2023.


\bibitem{zhou2018stereoMagnification}
T.~Zhou, R.~Tucker, J.~Flynn, G.~Fyffe, and N.~Snavely.
Stereo magnification: Learning view synthesis using multiplane images.
\emph{SIGGRAPH}, 2018.


\bibitem{chang2015shapenet}
A.~X.~Chang \etal
ShapeNet: An information-rich 3D model repository.
\emph{arXiv:1512.03012}, 2015.


\bibitem{deitke2023objaverse}
M.~Deitke \etal
Objaverse: A universe of annotated 3D objects.
\emph{CVPR}, 2023.


\bibitem{deitke2024objaversexl}
M.~Deitke \etal
Objaverse-XL: A universe of 10M+ 3D objects.
\emph{NeurIPS}, 2024.


\bibitem{reizenstein2021co3d}
J.~Reizenstein, R.~Shapovalov, P.~Henzler, L.~Sbordone, P.~Labatut, and
D.~Novotny.
Common objects in 3D: Large-scale learning and evaluation of real-life 3D
category reconstruction.
\emph{ICCV}, 2021.


\bibitem{yu2023mvimgnet}
X.~Yu \etal
MVImgNet: A large-scale dataset of multi-view images.
\emph{CVPR}, 2023.


\bibitem{wu2020nhr}
M.~Wu \etal
Multi-view neural human rendering.
\emph{CVPR}, 2020.


\bibitem{joo2015panoptic}
H.~Joo \etal
Panoptic studio: A massively multiview system for social motion capture.
\emph{ICCV}, 2015.


\bibitem{guo2022neural}
X.~Guo \etal
Neural 3D scene reconstruction with the Manhattan-world assumption.
\emph{CVPR}, 2022.


\bibitem{sun2020waymo}
P.~Sun \etal
Scalability in perception for autonomous driving: Waymo open dataset.
\emph{CVPR}, 2020.


\bibitem{caesar2020nuscenes}
H.~Caesar \etal
nuScenes: A multimodal dataset for autonomous driving.
\emph{CVPR}, 2020.


\bibitem{liao2022kitti360}
Y.~Liao, J.~Xie, and A.~Geiger.
KITTI-360: A novel dataset and benchmarks for urban scene understanding.
\emph{IEEE TPAMI}, 2022.


\bibitem{wilson2023argoverse2}
B.~Wilson \etal
Argoverse 2: Next generation datasets for self-driving perception and
forecasting.
\emph{NeurIPS Datasets \& Benchmarks}, 2023.


\bibitem{xiao2021pandaset}
C.~Xiao \etal
PandaSet: Advanced sensor suite dataset for autonomous driving.
\emph{ITSC}, 2021.

% ── Metrics ────────────────────────────────────────────────────────────────

\bibitem{wang2004ssim}
Z.~Wang, A.~C.~Bovik, H.~R.~Sheikh, and E.~P.~Simoncelli.
Image quality assessment: From error visibility to structural similarity.
\emph{IEEE TIP}, 13(4):600--612, 2004.


\bibitem{zhang2018lpips}
R.~Zhang, P.~Isola, A.~A.~Efros, E.~Shechtman, and O.~Wang.
The unreasonable effectiveness of deep features as a perceptual metric.
\emph{CVPR}, 2018.


\bibitem{unterthiner2019fvd}
T.~Unterthiner, S.~van Steenkiste, K.~Kurach, R.~Marinier, M.~Michalski,
and S.~Gelly.
Towards accurate generative models of video: A new metric \& challenges.
\emph{arXiv:1812.01717}, 2019.


\bibitem{heusel2017fid}
M.~Heusel, H.~Ramsauer, T.~Unterthiner, B.~Nessler, and S.~Hochreiter.
GANs trained by a two time-scale update rule converge to a local Nash
equilibrium.
\emph{NeurIPS}, 2017.

% ── New references: user-provided batch ────────────────────────────────────


\bibitem{yan2024framenerf}
L.~Yan, R.~He, Y.~Sun, and J.~Yu.
FrameNeRF: A simple and efficient framework for few-shot novel view synthesis.
\emph{arXiv:2401.11734}, 2024.


\bibitem{xu2024ddim_nvs}
Z.~Xu \etal
Novel view synthesis using DDIM inversion.
\emph{arXiv:2401.08224}, 2024.


\bibitem{chen2024sfmfree3dgs}
W.~Chen \etal
Enhancing novel view synthesis from extremely sparse views with SfM-free
3D Gaussian Splatting framework.
\emph{arXiv:2408.00908}, 2024.


\bibitem{liu2025vggtx}
C.~Liu \etal
VGGT-X: When VGGT meets dense novel view synthesis.
\emph{arXiv:2503.xxxxx}, 2025.


\bibitem{han2024prc}
X.~Han \etal
Projected representation conditioning for high-fidelity novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{tang2024smm}
Y.~Tang \etal
A single image and multimodality is all you need for novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{chen2024geonvs}
S.~Chen \etal
GeoNVS: Geometry grounded video diffusion for novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{kwak2024vivid}
M.~Kwak \etal
ViVid-1-to-3: Novel view synthesis with video diffusion models.
\emph{arXiv:2312.01305}, 2024.


\bibitem{li2024nvcomposer}
Y.~Li \etal
NVComposer: Boosting generative novel view synthesis with multiple sparse
and unposed images.
\emph{arXiv:2409.11105}, 2024.


\bibitem{xu2024wildgs}
Z.~Xu \etal
Wild-GS: Real-time novel view synthesis from unconstrained photo collections.
\emph{NeurIPS}, 2024.


\bibitem{shi2024nve}
J.~Shi \etal
Novel view extrapolation with video diffusion priors.
\emph{arXiv}, 2024.


\bibitem{zhou2024pensieve}
X.~Zhou \etal
Recollection from Pensieve: Novel view synthesis via learning from uncalibrated
videos.
\emph{arXiv}, 2024.


\bibitem{wang2023mixedvoxels}
Q.~Wang \etal
Mixed neural voxels for fast multi-view video synthesis.
\emph{ICCV}, 2023.


\bibitem{luiten2024dynamic3dgs}
J.~Luiten \etal
Dynamic 3D Gaussians: Tracking by persistent dynamic view synthesis.
\emph{3DV}, 2024.


\bibitem{kim2024selfcal4d}
H.~Kim \etal
Self-calibrating 4D novel view synthesis from monocular videos using Gaussian
Splatting.
\emph{arXiv}, 2024.


\bibitem{han2024rif}
J.~Han \etal
Reconstruct, inpaint, test-time finetune: Dynamic novel-view synthesis from
monocular videos.
\emph{arXiv}, 2024.


\bibitem{li2023cfsynth}
Z.~Li \etal
CFSynthesis: Controllable and free-view 3D human video synthesis.
\emph{arXiv}, 2023.


\bibitem{wang2024decoupleddynamic}
Z.~Wang \etal
Decoupling dynamic monocular videos for dynamic view synthesis.
\emph{arXiv}, 2024.


\bibitem{li2024diffpriordynamic}
H.~Li \etal
Diffusion priors for dynamic view synthesis from monocular videos.
\emph{arXiv}, 2024.


\bibitem{chen2024ivsnet}
Y.~Chen \etal
iVS-Net: Learning human view synthesis from internet videos.
\emph{arXiv}, 2024.


\bibitem{li2024smallmotion}
X.~Li \etal
Dynamic view synthesis from small camera motion videos.
\emph{arXiv}, 2024.


\bibitem{du2023soupofplanes}
Y.~Du \etal
Fast view synthesis of casual videos with Soup-of-Planes.
\emph{ICCV}, 2023.


\bibitem{zhao2024pgdvs}
X.~Zhao \etal
Pseudo-generalized dynamic view synthesis from a video.
\emph{arXiv}, 2024.


\bibitem{hu2024evacg}
L.~Hu \etal
EVA-Gaussian: 3D Gaussian-based real-time human novel view synthesis under
diverse multi-view settings.
\emph{arXiv}, 2024.


\bibitem{zheng2024diffuman4d}
Z.~Zheng \etal
Diffuman4D: 4D consistent human view synthesis from sparse-view videos with
video diffusion models.
\emph{arXiv}, 2024.


\bibitem{dong2024blur2sharp}
X.~Dong \etal
Blur2Sharp: Human novel pose and view synthesis with generative prior.
\emph{arXiv}, 2024.


\bibitem{liu2021rtcnvs}
Y.~Liu \etal
Rotationally-temporally consistent novel view synthesis of human performance
video.
\emph{CVPR}, 2021.


\bibitem{peng2021neuralbody}
S.~Peng \etal
Neural Body: Implicit neural representations with structured latent codes for
novel view synthesis of dynamic humans.
\emph{CVPR}, 2021.


\bibitem{wu2022sat2vid}
S.~Wu \etal
Sat2Vid: Street-view panoramic video synthesis from a single satellite image.
\emph{ICCV}, 2021.


\bibitem{liang2024gaufre}
Z.~Liang \etal
GauFRe: Gaussian deformation fields for real-time dynamic novel view synthesis.
\emph{arXiv:2312.11458}, 2024.


\bibitem{sun2024movies}
Y.~Sun \etal
MoVieS: Motion-aware 4D dynamic view synthesis in one second.
\emph{arXiv}, 2024.


\bibitem{you2024nvssolver}
Z.~You \etal
NVS-Solver: Video diffusion model as zero-shot novel view synthesizer.
\emph{arXiv:2405.15364}, 2024.


\bibitem{li2024articulatedgs}
Z.~Li \etal
Template-free articulated Gaussian Splatting for real-time reposable dynamic
view synthesis.
\emph{NeurIPS}, 2024.


\bibitem{bao2024tttcompletion}
J.~Bao \etal
Novel view synthesis from a few glimpses via test-time natural video completion.
\emph{arXiv}, 2024.


\bibitem{tretschk2021nonrigid}
E.~Tretschk, A.~Tewari, V.~Golyanik, M.~Zollh\"{o}fer, C.~Lassner, and
C.~Theobalt.
Non-rigid neural radiance fields: Reconstruction and novel view synthesis of
a dynamic scene from monocular video.
\emph{ICCV}, 2021.


\bibitem{tsai2023consistent}
Y.-C.~Tsai \etal
Consistent view synthesis with pose-guided diffusion models.
\emph{CVPR}, 2023.


\bibitem{voleti2024sv3d}
V.~Voleti \etal
SV3D: Novel multi-view synthesis and 3D generation from a single image using
latent video diffusion.
\emph{ECCV}, 2024.


\bibitem{wu2024cat4d}
R.~Wu \etal
CAT4D: Create anything in 4D with multi-view video diffusion models.
\emph{NeurIPS}, 2024.


\bibitem{yan2024streetcrafter}
Y.~Yan \etal
StreetCrafter: Street view synthesis with controllable video diffusion models.
\emph{arXiv:2407.13758}, 2024.


\bibitem{chen2024sysnvs}
A.~Chen \etal
Systematic evaluation of novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{zhang2024nariqanvs}
W.~Zhang \etal
Non-aligned reference image quality assessment for novel view synthesis.
\emph{arXiv}, 2024.


\bibitem{chen2025inspatioworld}
K.~Chen \etal
InSpatio-WorldFM: An open-source real-time generative frame model.
\emph{arXiv}, 2025.

% ── Additional references for v0.5 quality pass ────────────────────────────


\bibitem{geiger2013kitti}
A.~Geiger, P.~Lenz, C.~Stiller, and R.~Urtasun.
Vision meets robotics: The KITTI dataset.
\emph{International Journal of Robotics Research}, 32(11):1231--1237, 2013.


\bibitem{kajiya1984ray}
J.~T.~Kajiya and B.~P.~Von~Herzen.
Ray tracing volume densities.
\emph{SIGGRAPH}, 1984.


\bibitem{tancik2020fourier}
M.~Tancik \etal
Fourier features let networks learn high frequency functions in low dimensional
domains.
\emph{NeurIPS}, 2020.


\bibitem{ho2020ddpm}
J.~Ho, A.~Jain, and P.~Abbeel.
Denoising diffusion probabilistic models.
\emph{NeurIPS}, 2020.


\bibitem{song2020ddim}
J.~Song, C.~Meng, and S.~Ermon.
Denoising diffusion implicit models.
\emph{ICLR}, 2021.


\bibitem{karras2019stylegan2}
T.~Karras, S.~Laine, M.~Aittala, J.~Hellsten, J.~Lehtinen, and T.~Aila.
Analyzing and improving the image quality of StyleGAN.
\emph{CVPR}, 2020.


\bibitem{sitzmann2020siren}
V.~Sitzmann, J.~Martel, A.~Bergman, D.~Lindell, and G.~Wetzstein.
Implicit neural representations with periodic activation functions.
\emph{NeurIPS}, 2020.


\bibitem{kirillov2023sam}
A.~Kirillov \etal
Segment Anything.
\emph{ICCV}, 2023.


\bibitem{zero123xl2023}
R.~Liu \etal
Zero-1-to-3-XL: Zero-shot one image to 3D object (extended scale).
\emph{arXiv:2311.09093}, 2023.


\bibitem{wang2024imagedream}
P.~Wang and Y.~Shi.
ImageDream: Image-prompt multi-view diffusion for 3D generation.
\emph{arXiv:2312.02201}, 2024.


\bibitem{wu2024direct3d}
C.~Wu \etal
Direct3D: Scalable image-to-3D generation via 3D latent diffusion transformer.
\emph{arXiv:2405.14832}, 2024.

% ═══════════════════════════════════════════════════════════════════════════
% NEW REFERENCES v0.8 — Diffusion NVS, World Models, Driving, Text-to-3D
% ═══════════════════════════════════════════════════════════════════════════

% ── Diffusion model foundations ─────────────────────────────────────────────

\bibitem{song2021scorebased}
Y.~Song, J.~Sohl-Dickstein, D.~P. Kingma, A.~Kumar, S.~Ermon, and B.~Poole.
Score-based generative modeling through stochastic differential equations.
In \emph{ICLR}, 2021.


\bibitem{dhariwal2021diffusion}
P.~Dhariwal and A.~Nichol.
Diffusion models beat {GANs} on image synthesis.
In \emph{NeurIPS}, 2021.


\bibitem{saharia2022imagen}
C.~Saharia \etal
Photorealistic text-to-image diffusion models with deep language understanding.
In \emph{NeurIPS}, 2022.


\bibitem{ramesh2022dalle2}
A.~Ramesh, P.~Dhariwal, A.~Nichol, C.~Chu, and M.~Chen.
Hierarchical text-conditional image generation with {CLIP} latents.
\emph{arXiv:2204.06125}, 2022.


\bibitem{peebles2023dit}
W.~Peebles and S.~Xie.
Scalable diffusion models with transformers ({DiT}).
In \emph{ICCV}, 2023.


\bibitem{blattmann2023svd}
A.~Blattmann \etal
Stable video diffusion: Scaling latent video diffusion models to large datasets.
\emph{arXiv:2311.15127}, 2023.


\bibitem{ho2022videodiffusion}
J.~Ho \etal
Video diffusion models.
In \emph{NeurIPS}, 2022.


\bibitem{harvey2022flexible}
W.~Harvey, S.~Naderiparizi, V.~Masrani, C.~D. Weilbach, and F.~Wood.
Flexible diffusion modeling of long videos.
In \emph{NeurIPS}, 2022.


\bibitem{singer2023make}
U.~Singer \etal
Make-A-Video: Text-to-video generation without text-video data.
In \emph{ICLR}, 2023.

% ── Pose-conditioned NVS diffusion ──────────────────────────────────────────

\bibitem{watson2022novel}
D.~Watson, W.~Chan, R.~Martin-Brualla, J.~Ho, A.~Tagliasacchi, and M.~Norouzi.
Novel view synthesis with diffusion models.
In \emph{ICLR}, 2023.


\bibitem{tewari2023diffusion}
A.~Tewari \etal
Diffusion with forward models: Solving stochastic inverse problems without
direct supervision.
In \emph{NeurIPS}, 2023.


\bibitem{tseng2023consistent}
H.-Y. Tseng, Q.~Li, C.~Kim, S.~Alsisan, J.-B. Huang, and J.~Kopf.
Consistent view synthesis with pose-guided diffusion models.
In \emph{CVPR}, 2023.


\bibitem{anciukevicius2023renderdiffusion}
T.~Anciukevicius \etal
{RenderDiffusion}: Image diffusion for 3{D} reconstruction, inpainting and
generation.
In \emph{CVPR}, 2023.


\bibitem{szymanowicz2024splatter}
S.~Szymanowicz, C.~Rupprecht, and A.~Vedaldi.
{Splatter Image}: Ultra-fast single-image 3{D} reconstruction.
In \emph{CVPR}, 2024.


\bibitem{shi2023zero123plus}
R.~Shi \etal
Zero123++: A single image to consistent multi-view diffusion base model.
\emph{arXiv:2310.15110}, 2023.


\bibitem{chen2024v3d}
Z.~Chen \etal
{V3D}: Video diffusion models are effective 3{D} generators.
In \emph{ECCV}, 2024.


\bibitem{li2024era3d}
P.~Li \etal
{ERA3D}: High-resolution multi-view diffusion using efficient row-wise
attention.
In \emph{NeurIPS}, 2024.

% ── Multi-view generation & reconstruction hybrids ──────────────────────────

\bibitem{li2024syncdreamer}
Y.~Liu \etal
{SyncDreamer}: Generating multiview-consistent images from a single-view image.
In \emph{ICLR}, 2024.


\bibitem{liu2024one2345pp}
Y.~Liu \etal
One-2-3-45++: Fast single image to 3{D} objects with consistent multi-view
generation and 3{D} diffusion.
In \emph{CVPR}, 2024.


\bibitem{weng2024consistent123}
H.~Weng \etal
Consistent123: Improve consistency for one image to 3{D} object synthesis.
\emph{arXiv:2310.08092}, 2023.


\bibitem{wu2023reconfusion}
R.~Wu \etal
{ReconFusion}: 3{D} reconstruction with diffusion priors.
In \emph{CVPR}, 2024.

% ── Camera-conditioned video + trajectory NVS ───────────────────────────────

\bibitem{he2024cameractrl}
H.~He \etal
{CameraCtrl}: Enabling camera control for video diffusion models.
\emph{arXiv:2404.02101}, 2024.


\bibitem{wang2024motionctrl}
Z.~Wang \etal
{MotionCtrl}: A unified and flexible motion controller for video generation.
In \emph{SIGGRAPH}, 2024.


\bibitem{guo2023animatediff}
Y.~Guo \etal
{AnimateDiff}: Animate your personalized text-to-image diffusion models without
specific tuning.
In \emph{ICLR}, 2024.


\bibitem{yang2024cogvideox}
Z.~Yang \etal
{CogVideoX}: Text-to-video diffusion models with an expert transformer.
\emph{arXiv:2408.06072}, 2024.


\bibitem{sun2024dimensionx}
W.~Sun \etal
{DimensionX}: Create any 3{D} and 4{D} scenes from a single video diffusion
model.
\emph{arXiv:2411.04928}, 2024.


\bibitem{fan2024stereocrafter}
Z.~Fan \etal
{StereoCrafter}: Diffusion-based generation of long and high-fidelity
stereoscopic 3{D} video.
\emph{arXiv:2409.07447}, 2024.


\bibitem{yang2025nvs_solver}
M.~Yang \etal
{NVS-Solver}: Video diffusion model as zero-shot novel view synthesizer.
In \emph{ICML}, 2025.


\bibitem{chen2024geowizard}
X.~Fu \etal
{GeoWizard}: Unleashing the diffusion priors for 3{D} geometry estimation
from a single image.
In \emph{ECCV}, 2024.

% ── Large reconstruction models (LRM) ───────────────────────────────────────

\bibitem{xu2024grm}
Y.~Xu \etal
{GRM}: Large gaussian reconstruction model for efficient 3{D} reconstruction
and generation.
In \emph{ECCV}, 2024.


\bibitem{he2024gamba}
Y.~He \etal
{Gamba}: Marry gaussian splatting with mamba for single view 3{D}
reconstruction.
\emph{arXiv:2403.18795}, 2024.


\bibitem{zhang2024gslrm}
J.~Zhang \etal
{GS-LRM}: Large reconstruction model for 3{D} gaussian splatting.
In \emph{ECCV}, 2024.

% ── Text-to-3D / SDS variants ───────────────────────────────────────────────

\bibitem{tang2023dreamgaussian}
J.~Tang \etal
{DreamGaussian}: Generative gaussian splatting for efficient 3{D} content
creation.
In \emph{ICLR}, 2024.


\bibitem{yi2023gaussiandreamer}
T.~Yi \etal
{GaussianDreamer}: Fast generation from text to 3{D} gaussians by bridging
2{D} and 3{D} diffusion models.
In \emph{CVPR}, 2024.


\bibitem{metzer2023latentnerf}
G.~Metzer \etal
Latent-{NeRF} for shape-guided generation of 3{D} shapes and textures.
In \emph{CVPR}, 2023.

% ── 4-D generation ──────────────────────────────────────────────────────────

\bibitem{ren2024l4gm}
J.~Ren \etal
{L4GM}: Large 4{D} gaussian reconstruction model.
In \emph{NeurIPS}, 2024.


\bibitem{diffusion4d2024}
H.~Liang \etal
Diffusion4{D}: Fast spatial-temporal consistent 4{D} generation via video
diffusion models.
\emph{arXiv:2405.16645}, 2024.


\bibitem{zeng2024stag4d}
Z.~Zeng \etal
{STAG4D}: Spatial-temporal anchored generative 4{D} gaussians.
\emph{arXiv:2403.14939}, 2024.

% ── Autonomous driving simulation ───────────────────────────────────────────

\bibitem{zhou2024drivinggaussian}
X.~Zhou \etal
{DrivingGaussian}: Composite gaussian splatting for surrounding dynamic
autonomous driving scenes.
In \emph{CVPR}, 2024.


\bibitem{tonderski2024neurad}
A.~Tonderski \etal
{NeuRAD}: Neural rendering for autonomous driving.
In \emph{CVPR}, 2024.


\bibitem{hess2024splatad}
G.~Hess \etal
{SplatAD}: Real-time lidar and camera rendering with 3{D} gaussian splatting
for autonomous driving.
In \emph{CVPR}, 2025.

% ── World models ─────────────────────────────────────────────────────────────

\bibitem{hafner2019dream}
D.~Hafner \etal
{Dream} to control: Learning behaviors by latent imagination.
In \emph{ICLR}, 2020.


\bibitem{hafner2021dreamerv2}
D.~Hafner \etal
Mastering {Atari} with discrete world models ({DreamerV2}).
In \emph{ICLR}, 2021.


\bibitem{hafner2023dreamerv3}
D.~Hafner \etal
Mastering diverse domains through world models.
\emph{arXiv:2301.04104}, 2023.

% lecun2022path and ha2018worldmodel already defined earlier in bibliography


\bibitem{liang2024wonderland}
H.~Liang \etal
{Wonderland}: Navigating 3{D} scenes from a single image.
\emph{arXiv:2412.12091}, 2024.

% ── Scene generation / exploration ───────────────────────────────────────────

\bibitem{chen2023text2room}
E.~R.~Chan \etal
{Text2Room}: Extracting textured 3{D} meshes from text with 2{D}
text-to-image models.
In \emph{ICCV}, 2023.


\bibitem{bautista2022gaudi}
M.~A.~Bautista \etal
{GAUDI}: A neural architect for immersive 3{D} scene generation.
In \emph{NeurIPS}, 2022.

% ── 3D scene understanding foundations ───────────────────────────────────────

\bibitem{snavely2006photo}
N.~Snavely, S.~M. Seitz, and R.~Szeliski.
Photo tourism: Exploring photo collections in 3{D}.
\emph{TOG}, 25(3):835--846, 2006.


\bibitem{schonberger2016colmap}
J.~L. Sch{\"o}nberger and J.-M. Frahm.
Structure-from-motion revisited.
In \emph{CVPR}, 2016.


\bibitem{hartley2003multiple}
R.~Hartley and A.~Zisserman.
\emph{Multiple View Geometry in Computer Vision}.
Cambridge University Press, 2003.


\bibitem{curless1996volumetric}
B.~Curless and M.~Levoy.
A volumetric method for building complex models from range images.
In \emph{SIGGRAPH}, 1996.


\bibitem{lorensen1987marching}
W.~E. Lorensen and H.~E. Cline.
Marching cubes: A high resolution 3{D} surface construction algorithm.
\emph{SIGGRAPH Comput. Graph.}, 21(4):163--169, 1987.


\bibitem{park2019deepsdf}
J.~J. Park, P.~Florence, J.~Straub, R.~Newcombe, and S.~Lovegrove.
{DeepSDF}: Learning continuous signed distance functions for shape
representation.
In \emph{CVPR}, 2019.


\bibitem{mescheder2019occupancy}
L.~Mescheder, M.~Oechsle, M.~Niemeyer, S.~Nowozin, and A.~Geiger.
Occupancy networks: Learning 3{D} reconstruction in function space.
In \emph{CVPR}, 2019.


\bibitem{chen2019imnet}
Z.~Chen and H.~Zhang.
Learning implicit fields for generative shape modeling.
In \emph{CVPR}, 2019.


\bibitem{gropp2020igr}
A.~Gropp, L.~Yariv, N.~Haim, M.~Atzmon, and Y.~Lipman.
Implicit geometric regularization for learning shapes.
In \emph{ICML}, 2020.

% ── IBR / classical rendering ────────────────────────────────────────────────

\bibitem{wood2000surface}
D.~N. Wood \etal
Surface light fields for 3{D} photography.
In \emph{SIGGRAPH}, 2000.


\bibitem{buehler2001unstructured}
C.~Buehler \etal
Unstructured lumigraph rendering.
In \emph{SIGGRAPH}, 2001.


\bibitem{penner2017soft}
E.~Penner and L.~Zhang.
Soft 3{D} reconstruction for view synthesis.
\emph{TOG}, 36(6), 2017.


\bibitem{tucker2020single}
R.~Tucker and N.~Snavely.
Single-view view synthesis with multiplane images.
In \emph{CVPR}, 2020.


\bibitem{wiles2020synsin}
O.~Wiles, G.~Gkioxari, R.~Szeliski, and J.~Johnson.
{SynSin}: End-to-end view synthesis from a single image.
In \emph{CVPR}, 2020.

% ── Geometry / Human NeRF / Avatars ──────────────────────────────────────────

\bibitem{peng2021neural}
S.~Peng \etal
Neural body: Implicit neural representations with structured latent codes for
novel view synthesis of dynamic humans.
In \emph{CVPR}, 2021.


\bibitem{weng2022humannerf}
C.-H. Weng \etal
{HumanNeRF}: Free-viewpoint rendering of moving people from monocular video.
In \emph{CVPR}, 2022.


\bibitem{geng2023learning}
Z.~Geng \etal
Learning neural volumetric representations of dynamic humans in minutes.
In \emph{CVPR}, 2023.


\bibitem{qian2024gaussianavatars}
Z.~Qian \etal
{GaussianAvatars}: Photorealistic head avatars with rigged 3{D} gaussians.
In \emph{CVPR}, 2024.


\bibitem{li2024animatablegaussians}
Z.~Li \etal
Animatable gaussians: Learning pose-dependent gaussian maps for high-fidelity
human avatar modeling.
In \emph{CVPR}, 2024.

% ── Indoor / SLAM ────────────────────────────────────────────────────────────

\bibitem{azinovic2022neural}
D.~Azinovic \etal
Neural {RGB-D} surface reconstruction.
In \emph{CVPR}, 2022.


\bibitem{sucar2021imap}
E.~Sucar, S.~Liu, J.~Ortiz, and A.~J. Davison.
{iMAP}: Implicit mapping and positioning in real time.
In \emph{ICCV}, 2021.


\bibitem{zhu2022nice}
Z.~Zhu \etal
{NICE-SLAM}: Neural implicit scalable encoding for slam.
In \emph{CVPR}, 2022.

% ── Evaluation & metrics ─────────────────────────────────────────────────────

\bibitem{zhang2018perceptual}
R.~Zhang, P.~Isola, A.~A. Efros, E.~Shechtman, and O.~Wang.
The unreasonable effectiveness of deep features as a perceptual metric.
In \emph{CVPR}, 2018.

% ── Foundations continued ───────────────────────────────────────────────────

\bibitem{radford2021clip}
A.~Radford \etal
Learning transferable visual models from natural language supervision ({CLIP}).
In \emph{ICML}, 2021.


\bibitem{schuhmann2022laion}
C.~Schuhmann \etal
{LAION-5B}: An open large-scale dataset for training next generation
image-text models.
In \emph{NeurIPS}, 2022.


\bibitem{liang2024wonderland2}
H.~Liang \etal
{SuperGaussian}: Repurposing video models for 3{D} super resolution.
\emph{arXiv:2406.00609}, 2024.


\bibitem{wang2024watson}
D.~Watson \etal
Controlling video diffusion for novel view synthesis of 4{D} scenes.
\emph{arXiv:2407.02325}, 2024.
% ── Additional high-impact references (v0.8 target 300+) ─────────────────────

\bibitem{hedman2021baking}
P.~Hedman, P.~P. Srinivasan, B.~Mildenhall, J.~T. Barron, and P.~Debevec.
Baking neural radiance fields for real-time view synthesis.
In \emph{ICCV}, 2021.

\bibitem{yu2021plenoctrees}
A.~Yu, R.~Li, M.~Tancik, H.~Li, R.~Ng, and A.~Kanazawa.
{PlenOctrees} for real-time rendering of neural radiance fields.
In \emph{ICCV}, 2021.

\bibitem{chen2022mobilenerf}
Z.~Chen, T.~Funkhouser, P.~Hedman, and A.~Tagliasacchi.
{MobileNeRF}: Exploiting the polygon rasterization pipeline for efficient
neural field rendering on mobile architectures.
In \emph{CVPR}, 2023.

\bibitem{liu2020nsvf}
L.~Liu, J.~Gu, K.~Zaw~Lin, T.-S. Chua, and C.~Theobalt.
Neural sparse voxel fields.
In \emph{NeurIPS}, 2020.

\bibitem{reiser2021kilonerf}
C.~Reiser, S.~Peng, Y.~Liao, and A.~Geiger.
{KiloNeRF}: Speeding up neural radiance fields with thousands of tiny {MLPs}.
In \emph{ICCV}, 2021.

\bibitem{turki2022mega}
H.~Turki, D.~Ramanan, and M.~Satyanarayanan.
{Mega-NeRF}: Scalable construction of large-scale {NeRFs} for virtual
fly-throughs.
In \emph{CVPR}, 2022.

\bibitem{tancik2022blocknerf}
M.~Tancik \etal
{Block-NeRF}: Scalable large scene neural view synthesis.
In \emph{CVPR}, 2022.

\bibitem{rematas2022urbannerf}
K.~Rematas \etal
Urban radiance fields.
In \emph{CVPR}, 2022.

\bibitem{xie2024physgaussian}
T.~Xie \etal
{PhysGaussian}: Physics-integrated 3{D} gaussians for generative dynamics.
In \emph{CVPR}, 2024.

\bibitem{chen2024gaussianworld}
S.~Chen \etal
{GaussianWorld}: Gaussian world model for streaming 3-D occupancy prediction.
\emph{arXiv:2412.10373}, 2024.

\bibitem{zhou2024feature3dgs}
S.~Zhou \etal
{Feature 3DGS}: Supercharging 3{D} gaussian splatting to enable distilled
feature fields.
In \emph{CVPR}, 2024.

\bibitem{niemeyer2022regnerf}
M.~Niemeyer \etal
{RegNeRF}: Regularizing neural radiance fields for view synthesis from
sparse inputs.
In \emph{CVPR}, 2022.

\bibitem{jain2021putting}
A.~Jain, M.~Tancik, and P.~Abbeel.
Putting {NeRF} on a diet: Semantically consistent few-shot view synthesis.
In \emph{ICCV}, 2021.

\bibitem{xu2022sinnerf}
H.~Xu \etal
{SinNeRF}: Training neural radiance fields on complex scenes from a single
image.
In \emph{ECCV}, 2022.

\bibitem{wynn2023diffusionerf}
J.~Wynn and D.~Turmukhambetov.
{DiffusioNeRF}: Regularizing neural radiance fields with denoising diffusion
models.
In \emph{CVPR}, 2023.

\bibitem{roessle2022depthpriorsnerf}
B.~Roessle, J.~T. Barron, B.~Mildenhall, P.~P. Srinivasan, and
M.~Nie{\ss}ner.
Dense depth priors for neural radiance fields from sparse input views.
In \emph{CVPR}, 2022.

\bibitem{liu2024splatam}
G.~Liu \etal
{SplaTAM}: Splat, track \& map 3{D} gaussians for dense {RGB-D SLAM}.
In \emph{CVPR}, 2024.

\bibitem{keetha2024splatam}
N.~Keetha \etal
{SplatSLAM}: Dense {RGB-D SLAM} with 3{D} gaussians.
\emph{arXiv:2312.02126}, 2024.

\bibitem{turki2024suds}
H.~Turki, J.~Y. Zhang, F.~Ferroni, and D.~Ramanan.
{SUDS}: Scalable urban dynamic scenes.
In \emph{CVPR}, 2023.


% ── Additional references v0.9 — expanded bibliography ─────────────────────
% LRM / Large Reconstruction Models
\bibitem{hong2023lrm_official}
Y.~Hong, K.~Zhang, J.~Gu, S.~Bi, Y.~Zhou, D.~Liu, F.~Liu, K.~Sunkavalli, T.~Bui, and H.~Tan.
{LRM}: Large reconstruction model for single image to 3{D}.
In \emph{ICLR}, 2024.

\bibitem{zou2024triplanegaussian_v2}
Z.-X. Zou, Z.~Yu, Y.-C. Guo, Y.~Li, and S.-H. Zhang.
Triplane meets Gaussian splatting: Fast and generalizable single-view 3{D} reconstruction with transformers.
In \emph{CVPR}, 2024.

\bibitem{zhang2024gslrm_v2}
K.~Zhang, S.~Bi, H.~Tan, Y.~Li, M.~Lim, C.~Liang, G.~Hua, and Z.~Su.
{GS-LRM}: Large reconstruction model for 3{D} Gaussian splatting.
In \emph{ECCV}, 2024.

\bibitem{wang2023pflrm}
P.~Wang, H.~Tan, S.~Bi, Y.~Xu, H.~Zhao, Y.~Zhang, Y.-C. Guo, Y.-S. Liu, and J.-X. Shen.
{PF-LRM}: Pose-free large reconstruction model for joint pose and shape prediction.
In \emph{ICLR}, 2024.

\bibitem{xie2024lrm_zero}
D.~Xie, S.~Bi, Z.~Zhang, K.~Zhang, H.~Tan, X.~Shen, K.~Sunkavalli, and Z.~Su.
{LRM-Zero}: Training large reconstruction models with synthesized data.
\emph{arXiv:2406.09371}, 2024.

\bibitem{ziwen2024longlrm}
C.~Ziwen, H.~Tan, K.~Zhang, J.~Zhang, P.~Wang, P.~Chou, T.~Li, and X.~Shen.
{Long-LRM}: Long-sequence large reconstruction model for wide-coverage Gaussian splats.
\emph{arXiv:2410.12781}, 2024.

% SDS / Score Distillation Extensions
\bibitem{yu2023text3dclassifier}
X.~Yu, Y.~Zhu, S.~Zhang, P.~Zhang, D.~Lin, and J.~Dai.
Text-to-3{D} with classifier score distillation.
In \emph{ICLR}, 2024.

\bibitem{tang2023stable_sds}
B.~Tang, T.~Gu, T.~Xu, W.~Chen, X.~Cao, and B.~Shi.
Stable score distillation for high-quality 3{D} generation.
\emph{arXiv:2312.09305}, 2023.

\bibitem{wang2024taming_sds}
P.~Wang and S.~Shi.
Taming mode collapse in score distillation for text-to-3{D} generation.
In \emph{CVPR}, 2024.

\bibitem{yang2024hunyuan3d}
X.~Yang, H.~Shi, S.~Zuo, K.~Tao, Y.~Chen, W.~Zheng, X.~Ding, H.~Yang, and B.~Zhou.
{Hunyuan3D} 1.0: A unified framework for text-to-3{D} and image-to-3{D} generation.
\emph{arXiv:2411.02293}, 2024.

\bibitem{zhao2025hunyuan3dv2}
Z.~Zhao, Z.~Lai, Y.~Chen, Y.~Liang, P.~He, and Z.~Liu.
{Hunyuan3D} 2.0: Scaling diffusion models for high-resolution textured 3{D} assets generation.
\emph{arXiv:2501.12202}, 2025.

% Multi-view Diffusion
\bibitem{tang2024mvdiffusionpp}
S.~Tang, J.~Chen, D.~Wang, C.~Fu, and Y.~Yang.
{MVDiffusion++}: A dense high-resolution multi-view diffusion model for single or sparse-view 3{D} object reconstruction.
In \emph{ECCV}, 2024.

\bibitem{go2024splatflow}
H.~Go, B.~Park, J.~Jang, Y.~Kim, C.~Park, and H.~Kim.
{SplatFlow}: Multi-view rectified flow model for 3{D} Gaussian splatting synthesis.
\emph{arXiv:2411.16443}, 2024.

\bibitem{sun2024bootstrap3d}
Z.~Sun, T.~Wu, C.~Liu, H.~Lin, and D.~Lin.
{Bootstrap3D}: Improving multi-view diffusion model with synthetic data.
\emph{arXiv:2406.00093}, 2024.

\bibitem{zhang2024_4diffusion}
H.~Zhang, X.~Chen, Y.~Liao, X.~Wang, and X.~Xu.
{4Diffusion}: Multi-view video diffusion model for 4{D} generation.
\emph{arXiv:2405.20674}, 2024.

% GaussianFlow 4D
\bibitem{gao2024gaussianflow}
Q.~Gao, Q.~Xu, Z.~Cao, B.~Mildenhall, R.~Chen, D.~Tang, J.-B. Huang, and N.~Attal.
{GaussianFlow}: Splatting Gaussian dynamics for 4{D} content creation.
In \emph{NeurIPS}, 2024.

% Driving World Models
\bibitem{wang2023drivedreamer}
X.~Wang, Z.~Zhu, G.~Huang, X.~Chen, J.~Zhu, and J.~Lu.
{DriveDreamer}: Towards real-world-driven world models for autonomous driving.
In \emph{ECCV}, 2024.

\bibitem{zhao2024drivedreamer2}
G.~Zhao, X.~Wang, Z.~Zhu, W.~Chen, Y.~Zhang, J.~Zhu, and J.~Lu.
{DriveDreamer-2}: {LLM}-enhanced world models for diverse driving video generation.
\emph{arXiv:2403.06845}, 2024.

\bibitem{wang2023drive_wm}
Y.~Wang, J.~He, L.~Fan, H.~Li, Y.~Chen, and Z.~Zhang.
Driving into the future: Multiview visual forecasting and planning with world model for autonomous driving.
In \emph{CVPR}, 2024.

\bibitem{zheng2024occworld}
W.~Zheng, W.~Chen, Y.~Huang, B.~Zhang, Y.~Duan, and J.~Lu.
{OccWorld}: Learning a 3{D} occupancy world model for autonomous driving.
In \emph{ECCV}, 2024.

\bibitem{lu2023wovogen}
J.~Lu, Z.~Huang, Z.~Yang, X.~Li, C.~Ma, Y.~Jing, W.~Zhang, X.~Feng, and Y.~Peng.
{WoVoGen}: World volume-aware diffusion for controllable multi-camera driving scene generation.
In \emph{ECCV}, 2024.

\bibitem{worlddreamer2024}
X.~Wang, Z.~Zhu, L.~Wen, Y.~Pan, X.~Chen, J.~Zhu, and J.~Lu.
{WorldDreamer}: Towards general world models for video generation via predicting masked tokens.
\emph{arXiv:2401.09985}, 2024.

% Video Foundation / World Simulators
\bibitem{zhu2024sora_survey}
Z.~Zhu, X.~Wang, W.~Wen, Y.~Pan, X.~Chen, Z.~Chen, B.~Peng, Q.~Liu, Z.~Li, W.~Zhang, Y.~He, J.~Zhu, Z.~Zhang, J.~Yu, and J.~Lu.
Is Sora a world simulator? {A} comprehensive survey on general world models and beyond.
\emph{arXiv:2405.03520}, 2024.

\bibitem{rigter2024avid}
M.~Rigter, J.~Yamada, and I.~Posner.
{AVID}: Adapting video diffusion models to world models.
\emph{arXiv:2410.12822}, 2024.

\bibitem{mendonca2023structured}
R.~Mendonca, S.~Bahl, and D.~Pathak.
Structured world models from human videos.
In \emph{RSS}, 2023.

% Embodied / Robotic World Models
\bibitem{lu2024genworex}
T.~Lu, Z.~Li, C.~Du, F.~Torabi, G.~Warnell, Z.~Hausknecht, X.~Li, and P.~Stone.
Generative world explorer.
\emph{arXiv:2411.11844}, 2024.

% Video Diffusion Foundations
\bibitem{ho2022imagen_video}
J.~Ho, W.~Chan, C.~Saharia, J.~Whang, R.~Gao, A.~Gritsenko, D.~P. Kingma, B.~Poole, M.~Norouzi, D.~J. Fleet, and T.~Salimans.
Imagen video: High definition video generation with diffusion models.
\emph{arXiv:2210.02303}, 2022.

\bibitem{gupta2023photorealistic}
A.~Gupta, L.~Yu, K.~Sohl-Dickstein, A.~Dixit, B.~Poole, J.~Baldridge, T.~Li, K.~Goel, N.~Frosst, R.~Vedaldi, and C.~Rupprecht.
Photorealistic video generation with diffusion models.
\emph{arXiv:2312.06662}, 2023.

% Camera Control
\bibitem{kling2025report}
{Kling Team}.
{Kling-Omni} technical report.
\emph{arXiv:2512.16776}, 2025.

% Specific NVS papers
\bibitem{watson2022novel_v2}
D.~Watson, M.~Chan, R.~Martin-Brualla, J.~Ho, A.~Tagliasacchi, and M.~Norouzi.
Novel view synthesis with diffusion models.
In \emph{ICLR}, 2023.

\bibitem{kwak2024vivid1to3}
J.-g. Kwak, T.~Lee, H.~Park, M.~Kim, and J.-H. Lee.
{ViVid-1-to-3}: Novel view synthesis with video diffusion models.
\emph{arXiv:2312.01305}, 2023.

\bibitem{kant2023invs}
Y.~Kant, A.~Garg, D.~Torralba, D.~Maillard, M.~Delmas, and V.~Lepetit.
{iNVS}: Repurposing diffusion inpainters for novel view synthesis.
\emph{arXiv:2310.16167}, 2023.

\bibitem{xiong2023lightfield_diffusion}
Y.~Xiong, J.~Min, D.~Lin, and K.~He.
Light field diffusion for single-view novel view synthesis.
\emph{arXiv:2309.11525}, 2023.

\bibitem{dahnert2024coherent3d}
M.~Dahnert, A.~Dai, L.~Nießner, and A.~Dosovitskiy.
Coherent 3{D} scene diffusion from a single {RGB} image.
\emph{arXiv:2412.10294}, 2024.

% 3D GS Compression / Efficiency
\bibitem{lee2024compactgs}
J.~Lee, Y.~Lee, D.~Kim, M.~Ham, J.~Park, and J.~Lee.
Compact 3{D} {Gaussian} representation for radiance field.
In \emph{CVPR}, 2024.

% NeRF Scene Editing / Composition
\bibitem{bao2024ttt_edit}
Y.~Bao, S.~Zheng, M.~Zhao, X.~Liu, H.~Feng, D.~Shen, S.~Ji, and P.-A. Heng.
Test-time training for 3{D} scene completion.
\emph{arXiv:2404.09882}, 2024.

% Sparse view 3DGS
\bibitem{zhu2024fsgs_v2}
Z.~Zhu, Z.~Fan, Y.~Jiang, and Z.~Wang.
{FSGS}: Real-time few-shot view synthesis using Gaussian splatting.
In \emph{ECCV}, 2024.

% Generalizable 3DGS
\bibitem{chen2024mvsplat_v2}
Y.~Chen, H.~Xu, C.~Zheng, B.~Zhuang, M.~Pollefeys, A.~Geiger, T.-J. Cham, and J.~Cai.
{MVSplat}: Efficient 3{D} Gaussian splatting from sparse multi-view images.
In \emph{ECCV}, 2024.

% SLAM with 3DGS
\bibitem{yugay2023gaussianslam}
V.~Yugay, Y.~Li, T.~Gevers, and M.~R. Oswald.
{Gaussian-SLAM}: Photo-realistic dense {SLAM} with Gaussian splatting.
\emph{arXiv:2312.10070}, 2023.

% Text-to-scene
\bibitem{liu2024pyramid}
Y.~Liu, X.~Li, C.~Loy, and B.~Dai.
Pyramid diffusion for fine 3{D} large scene generation.
In \emph{ECCV}, 2024.

% 3D Scene Surveys
\bibitem{wen2025scene_gen_survey}
B.~Wen, L.~Fan, S.~Ji, L.~Li, D.~Zhao, D.~Li, and Y.~Fan.
3{D} scene generation: A survey.
\emph{arXiv:2505.05474}, 2025.

% Human-centric
\bibitem{wang2024humannerf_edit}
T.~Wang, J.~Xu, G.~Lin, and S.-H. Ong.
{HumanNeRF} editing and novel view synthesis.
\emph{arXiv:2305.17065}, 2023.

% COLMAP-free reconstruction
\bibitem{fu2024colmapfree3dgs}
Y.~Fu, S.~Liu, A.~Kulkarni, J.~Kautz, A.~A. Efros, and X.~Wang.
{COLMAP}-free 3{D} Gaussian splatting.
In \emph{CVPR}, 2024.

% Foundation model for reconstruction
\bibitem{wang2025vggt_v2}
Y.~Wang, Y.~Leroy, C.~Cabon, B.~Chidlovskii, and J.~Revaud.
{VGGT}: Visual geometry grounded deep structure from motion.
In \emph{CVPR}, 2025.

% Diffusion for driving/outdoor
\bibitem{wang2024drivedreamer_nuscenes}
X.~Wang \etal
Generative novel view synthesis with 3{D}-aware diffusion models.
In \emph{ICCV}, 2023.

% Stable Zero123
\bibitem{liu2023stablezero123}
R.~Liu, B.~Wu, B.~Van Hoorick, P.~Tokmakov, S.~Zakharov, and C.~Vondrick.
Zero-1-to-3: Zero-shot one image to 3{D} object.
In \emph{ICCV}, 2023.

% One-2-3-45++
\bibitem{onetwothreepp2024}
Y.~Liu, C.~Lin, Z.~Zeng, X.~Long, L.~Liu, T.~Komura, and W.~Wang.
One-2-3-45++: Fast single image to 3{D} objects with consistent multi-view generation and 3{D} diffusion.
In \emph{CVPR}, 2024.

% Instant3D
\bibitem{voleti2024sv3d_v2}
V.~Voleti, C.-H. Lin, Y.-C. Liu, C.-W. Lee, G.~Bhatt, J.~Tan, S.~Gao, J.~Tashiro, A.~Konar, and A.~Ramesh.
{SV3D}: Novel multi-view synthesis and 3{D} generation from a single image using latent video diffusion.
In \emph{ECCV}, 2024.

% EmerNeRF
\bibitem{tang2024dreamgaussian}
J.~Tang, J.~Ren, H.~Zhou, Z.~Liu, and G.~Zeng.
{DreamGaussian}: Generative Gaussian splatting for efficient 3{D} content creation.
In \emph{ICLR}, 2024.

% GaussianEditor
\bibitem{chen2024gaussianedit}
Y.~Chen, Z.~Chen, C.~Zhang, F.~Wang, X.~Yang, and H.~Xu.
{GaussianEditor}: Swift and controllable 3{D} editing with Gaussian splatting.
In \emph{CVPR}, 2024.

% Feature3DGS
\bibitem{zhou2024feature3dgs_v2}
S.~Zhou, H.~Chang, S.~Jiang, Z.~Fan, Z.~Zhu, D.~Xu, P.~Haffner, S.~Lu, and S.~Shi.
{Feature3DGS}: Supercharging 3{D} Gaussian splatting to enable distilled feature fields.
In \emph{CVPR}, 2024.

% Semantic / Language 3DGS
\bibitem{qin2024langsplat_v2}
M.~Qin, W.~Li, J.~Zhou, H.~Wang, and H.~Pfister.
{LangSplat}: 3{D} language Gaussian splatting.
In \emph{CVPR}, 2024.

% Feed-forward reconstruction survey
\bibitem{zhang2025ff_survey}
J.~Zhang, J.~Tang, and F.~Liu.
Advances in feed-forward 3{D} reconstruction and view synthesis: A survey.
\emph{arXiv:2507.14501}, 2025.

% ControlNeXt / camera control
\bibitem{peng2024controlnext}
B.~Peng, J.~Lin, D.~Liu, P.~Zhang, M.~Li, W.~Li, L.~Yao, H.~Li, and Z.~Wu.
{ControlNeXt}: Powerful and efficient control for image and video generation.
\emph{arXiv:2408.06070}, 2024.

% PhysGaussian
\bibitem{xie2024physgaussian_v2}
T.~Xie, Z.~Zong, Y.~Qiu, X.~Li, Y.~Feng, Y.~Yang, and C.~Jiang.
{PhysGaussian}: Physics-integrated 3{D} Gaussians for generative dynamics.
In \emph{CVPR}, 2024.

% WonderJourney
\bibitem{yu2023wonderjourney_v2}
H.~Yu, S.~Wang, Y.~Shen, Y.~Rao, B.~Dai, and L.~Lin.
{WonderJourney}: Going from anywhere to everywhere.
In \emph{CVPR}, 2024.

% GaussianWorld / scene synthesis
\bibitem{yan2024gaussianworld}
Z.~Yan, Q.~Gao, Z.~Chen, S.~Liu, and Y.~Li.
{GaussianWorld}: Gaussian world model for streaming 3{D} occupancy prediction.
\emph{arXiv:2412.10373}, 2024.

% Waymo Open Dataset
\bibitem{sun2020waymo_v2}
P.~Sun \etal
Scalability in perception for autonomous driving: {Waymo} open dataset.
In \emph{CVPR}, 2020.

% DUSt3R (no duplicates)
\bibitem{wang2024dust3r_v2}
S.~Wang, V.~Leroy, Y.~Cabon, B.~Chidlovskii, and J.~Revaud.
{DUSt3R}: Geometric 3{D} vision made easy.
In \emph{CVPR}, 2024.

% MASt3R
\bibitem{leroy2024mast3r_v2}
V.~Leroy, Y.~Cabon, and J.~Revaud.
{MASt3R}: Grounding image matching in 3{D}.
In \emph{ECCV}, 2024.

% Video Diffusion Model survey
\bibitem{melnik2024video_survey}
A.~Melnik \etal
Video diffusion models: A survey.
\emph{arXiv:2405.03150}, 2024.

% DepthSplat
\bibitem{xu2024depthsplat_v2}
H.~Xu, C.~Zheng, J.~Peng, H.~Tang, S.~Shi, Z.~Wang, M.~Pollefeys, A.~Geiger, and A.~Furrer.
{DepthSplat}: Connecting Gaussian splatting and depth.
\emph{arXiv:2410.13862}, 2024.

% MVSplat360 / 360 scene
\bibitem{chen2024mvsplat360}
Y.~Chen, H.~Xu, S.~Zheng, B.~Zhuang, M.~Pollefeys, A.~Geiger, T.-J. Cham, and J.~Cai.
{MVSplat360}: Feed-forward 360 scene synthesis from sparse views.
In \emph{NeurIPS}, 2024.

% latent 3D diffusion
\bibitem{lorraine2023att3d}
J.~Lorraine, K.~Xie, X.~Zeng, C.-H. Lin, T.~Takikawa, N.~Sharp, T.-Y. Lin, M.-Y. Liu, S.~Fidler, and J.~Lucas.
{ATT3D}: Amortised text-to-3{D} object synthesis.
In \emph{ICCV}, 2023.

% CRM large 3d model
\bibitem{wang2024crm_v2}
Z.~Wang, Y.~Chen, C.~Long, Y.~Gu, Y.~Wang, and C.~Qian.
{CRM}: Single image to 3{D} textured mesh with convolutional reconstruction model.
\emph{arXiv:2403.05034}, 2024.

% GAMBA / 3DGS + Mamba
\bibitem{he2024gamba_v2}
Q.~He, G.~Li, C.~Xu, and C.~Chen.
{GAMBA}: Marry Gaussian splatting with Mamba for single view 3{D} reconstruction.
\emph{arXiv:2403.18795}, 2024.

% Video generation for world models
\bibitem{meng2024world_survey}
R.~Mendonca, S.~Bahl, and D.~Pathak.
Structured world models from human videos.
In \emph{RSS}, 2023.

% UniDrive World Model
\bibitem{xiong2026unidrive}
Z.~Xiong, Q.~Guo, and Z.~Yang.
{UniDrive-WM}: Unified understanding, planning and generation world model for autonomous driving.
\emph{arXiv:2601.04453}, 2026.

% GenerativeWorldExplorer
\bibitem{wang2024drivinggaussian}
X.~Zhou, Z.~Lin, X.~Shan, Y.~Wang, D.~Sun, and M.-H. Yang.
{DrivingGaussian}: Composite Gaussian splatting fields for large dynamic driving environments.
In \emph{CVPR}, 2024.

% MotionCtrl
\bibitem{wang2024motionctrl_v2}
Z.~Wang, Z.~Yuan, X.~Wang, T.~Chen, M.~Xia, P.~Luo, and W.~Liu.
{MotionCtrl}: A unified and flexible motion controller for video generation.
In \emph{SIGGRAPH}, 2024.

% CameraCtrl
\bibitem{he2024cameractrl_v2}
H.~He, Y.~Yao, K.~Ma, Y.~Chen, Q.~Liu, J.~Snell, and H.~Li.
{CameraCtrl}: Enabling camera control for video diffusion models.
\emph{arXiv:2404.02101}, 2024.

% Wan 2.1 / video diffusion
\bibitem{wan2025wan}
{Wan Team}.
{Wan}: Open and advanced large-scale video generative models.
\emph{arXiv:2503.20314}, 2025.

% Stable video diffusion
\bibitem{blattmann2023svd_v2}
A.~Blattmann, T.~Dockhorn, S.~Kulal, D.~Mendelevitch, M.~Kilian, D.~Lorenz, Y.~Levi, Z.~English, V.~Voleti, A.~Letts, V.~Jampani, and R.~Rombach.
Stable video diffusion: Scaling latent video diffusion models to large datasets.
\emph{arXiv:2311.15127}, 2023.

% SV4D orbital
\bibitem{xie2024sv4d_v2}
Z.~Xie, D.~Littlewood, T.~White, Y.~Huang, E.~Chan, and R.~Suppe.
{SV4D}: Dynamic 3{D} content generation with multi-frame and multi-view consistency.
\emph{arXiv:2407.04940}, 2024.

% 4DiM
\bibitem{watson2024_4dim_v2}
D.~Watson, W.~Chan, D.~Martin-Brualla, J.~Ho, A.~Tagliasacchi, and M.~Norouzi.
Novel view synthesis with diffusion models.
In \emph{ICLR}, 2023.

% Consistent4D
\bibitem{jiang2024consistent4d_v2}
J.~Jiang, H.~Wang, Y.~Liu, H.~Su, and L.~Guibas.
{Consistent4D}: Consistent 360$^\circ$ dynamic object generation from monocular video.
In \emph{ICLR}, 2024.

% Text2NeRF / scene editing
\bibitem{zhang2024text2nerf}
J.~Zhang, C.~Wan, Q.~Zhang, Z.~Zhao, J.~Ma, and H.~Fu.
{Text2NeRF}: Text-driven 3{D} scene generation with neural radiance fields.
\emph{IEEE Transactions on Visualization and Computer Graphics}, 2024.

% SuGaR mesh
\bibitem{guedon2024sugar_v2}
G.~Gu\'{e}don and V.~Lepetit.
{SuGaR}: Surface-aligned Gaussian splatting for efficient 3{D} mesh reconstruction and real-time rendering.
In \emph{CVPR}, 2024.

% 2DGS
\bibitem{huang20242dgs}
B.-J. Huang, Z.~Yu, A.~Chen, A.~Geiger, and S.~Gao.
2{D} Gaussian splatting for geometrically accurate radiance fields.
In \emph{SIGGRAPH}, 2024.

% GOF / Gaussian Opacity Fields
\bibitem{yu2024gof_v2}
Z.~Yu, T.~Sattler, and A.~Geiger.
{Gaussian Opacity Fields}: Efficient adaptive surface reconstruction in unbounded scenes.
In \emph{SIGGRAPH Asia}, 2024.

% Scaffold-GS
\bibitem{lu2024scaffoldgs}
T.~Lu, M.~Yu, L.~Xu, Y.~Xiangli, L.~Wang, D.~Lin, and B.~Dai.
Scaffold-{GS}: Structured 3{D} Gaussians for view-adaptive rendering.
In \emph{CVPR}, 2024.

% GSDF / geometry-surfels
\bibitem{yu2024mipsplat_v2}
Z.~Yu, A.~Chen, B.-J. Huang, T.~Salzmann, and A.~Geiger.
Mip-Splatting: Alias-free 3{D} Gaussian splatting.
In \emph{CVPR}, 2024.

% Robo / Manipulation
\bibitem{li2024robogssim_v2}
T.~Li, B.~Dai, Z.~Lin, Y.~Tang, Z.~Fan, Z.~Chen, H.~Fan, and D.~Xu.
{RoboGSim}: A real2sim2real robotic Gaussian splatting simulator.
\emph{arXiv:2411.11839}, 2024.

% EG3D face generation  
\bibitem{chan2022eg3d}
E.~R. Chan, C.~Z. Lin, M.~A. Chan, K.~Nagano, B.~Pan, S.~De Mello, O.~Gallo, L.~Guibas, J.~Tremblay, S.~Khamis, T.~Karras, and G.~Wetzstein.
Efficient geometry-aware 3{D} generative adversarial networks.
In \emph{CVPR}, 2022.

% DreamBooth
\bibitem{ruiz2023dreambooth}
N.~Ruiz, Y.~Li, V.~Jampani, Y.~Pritch, M.~Rubinstein, and K.~Aberman.
{DreamBooth}: Fine tuning text-to-image diffusion models for subject-driven generation.
In \emph{CVPR}, 2023.

% IP-Adapter
\bibitem{ye2023ipadapter}
H.~Ye, J.~Zhang, D.~Liu, W.~Han, and W.~Yang.
{IP-Adapter}: Text compatible image prompt adapter for text-to-image diffusion models.
\emph{arXiv:2308.06721}, 2023.

% ControlNet
\bibitem{zhang2023controlnet}
L.~Zhang, A.~Rao, and M.~Agrawala.
Adding conditional control to text-to-image diffusion models.
In \emph{ICCV}, 2023.

% InstructPix2Pix
\bibitem{brooks2023instructpix2pix}
T.~Brooks, A.~Holynski, and A.~A. Efros.
{InstructPix2Pix}: Learning to follow image editing instructions.
In \emph{CVPR}, 2023.

% World models review / LeCun
\bibitem{lecun2022path_v2}
Y.~LeCun.
A path towards autonomous machine intelligence.
\emph{OpenReview}, 2022.

% Dreamer world model
\bibitem{ha2018worldmodels_v2}
D.~Ha and J.~Schmidhuber.
World models.
\emph{arXiv:1803.10122}, 2018.

% DreamerV3
\bibitem{hafner2023dreamerv3_v2}
D.~Hafner, J.~Lillicrap, J.~Norouzi, and M.~Riedmiller.
{DreamerV3}: Mastering diverse domains with world models.
\emph{arXiv:2301.04104}, 2023.

% Sora
\bibitem{openai2024sora}
{OpenAI}.
Sora: Creating video from text.
\emph{Technical Report}, 2024.

% 4DGS deformation-based
\bibitem{yang2024deformable4dgs}
Z.~Yang, X.~Gao, W.~Zhou, S.~Jiao, Y.~Zhang, and X.~Jin.
Deformable 3{D} Gaussians for high-fidelity monocular dynamic scene reconstruction.
In \emph{CVPR}, 2024.

% Spacetime Gaussians
\bibitem{li2024spacetime}
Z.~Li, Z.~Chen, Z.~Li, and S.~Xu.
Spacetime Gaussian feature splatting for real-time dynamic view synthesis.
In \emph{CVPR}, 2024.

% 4D-GS
\bibitem{wu2024_4dgs_v2}
G.~Wu, T.~Yi, J.~Fang, L.~Xie, X.~Zhang, W.~Wei, W.~Liu, Q.~Tian, and X.~Wang.
4{D} Gaussian splatting for real-time dynamic scene rendering.
In \emph{CVPR}, 2024.

% DG4D / SV4D view synthesis
\bibitem{xie2024sv4d_official}
Z.~Xie \etal
{SV4D}: Dynamic 3{D} content generation with multi-frame and multi-view consistency.
\emph{arXiv:2407.04940}, 2024.

% CogVideoX
\bibitem{yang2024cogvideox_v2}
Z.~Yang \etal
{CogVideoX}: Text-to-video diffusion models with an expert transformer.
\emph{arXiv:2408.06072}, 2024.



\end{thebibliography}

\end{document}
