\section{Scene Representation Taxonomy}
\label{sec:representation}

The choice of scene representation determines the fundamental trade-offs a
method faces: rendering speed, reconstruction accuracy, memory footprint,
generalisability, and editability.
We identify four families, each reflecting a different position on the
speed--compactness--quality Pareto frontier.

\textbf{Implicit representations} (MLPs, hash grids) encode the scene as a
continuous function $f_\theta(\mathbf{x})$.
Their key advantages are compactness and continuous differentiability, which
support high-quality reconstruction from dense observations.
Their central limitation is rendering speed: each pixel requires integrating
along a ray through many network evaluations.
NeRF~\citep{mildenhall2020nerf} and its successors (§\ref{sec:nerf}) are the
canonical family.

\textbf{Explicit representations} (voxels, point clouds, Gaussians) store scene
attributes at discrete spatial elements, enabling tile-based or splat-based
rasterisation that achieves real-time frame rates.
The trade-off is memory: a faithful 3DGS scene may occupy hundreds of megabytes.
3D Gaussian Splatting~\citep{kerbl2023gaussian} (§\ref{sec:3dgs}) is the
dominant current representative.

\textbf{Hybrid representations} combine an explicit spatial structure---typically
a triplane~\citep{chan2022efficient} or multi-resolution
grid~\citep{muller2022instant}---with a shallow MLP decoder.
They strike a practical balance between the quality of implicit models and
the speed of explicit ones, and are the backbone of most large reconstruction
models (§\ref{sec:generalisable}).

\textbf{Compositional representations} decompose the scene into semantic
entities (objects, backgrounds, agents) with independent sub-representations.
This structural inductive bias enables object-level editing and compositional
generation~\citep{niemeyer2021giraffe,zhou2024gala3d}, but typically requires
supervision or segmentation to separate entities.
It is also the natural representation for world models that must reason about
independently acting objects.

% ─────────────────────────────────────────────────────────────────────────────
