\section{NeRF: Implicit Neural Radiance Fields}
\label{sec:nerf}

\subsection{Foundational NeRF Methods}

The original NeRF~\citep{mildenhall2020nerf} demonstrated that an MLP with
positional encoding could faithfully represent complex real scenes with
view-dependent appearance from a few dozen posed photographs.
Subsequent work systematically addressed its two principal limitations---slow
optimisation and poor generalisation to novel scenes.

\paragraph{Accelerating per-scene optimisation.}
The bottleneck in vanilla NeRF is the MLP query: every rendered pixel
requires hundreds of network evaluations along the ray.
Instant-NGP~\citep{muller2022instant} replaces the large MLP backbone with a
compact multi-resolution \emph{hash grid}, enabling training convergence in
under a minute on a single GPU while matching or exceeding prior quality.
TensoRF~\citep{chen2022tensorf} decomposes the radiance field into low-rank
CP and VM tensor components, reducing both storage and rendering overhead.
Plenoxels~\citep{fridovich2022plenoxels} eliminates the MLP entirely,
using a sparse 3-D voxel grid with trilinearly interpolated spherical harmonics;
the resulting model trains in minutes without gradient-intensive network
backward passes.
DirectVoxGO~\citep{sun2022dvgo} accelerates convergence via a two-stage
coarse-to-fine voxel grid that aggressively skips empty space.

\paragraph{Improving visual quality and handling challenging scenes.}
Mip-NeRF~\citep{barron2021mipnerf} replaces the point-sampled ray with an
\emph{integrated positional encoding} over a conical frustum, substantially
reducing aliasing at multiple viewing distances.
Mip-NeRF~360~\citep{barron2022mipnerf360} extends this to unbounded outdoor
scenes via a non-linear scene parameterisation that contracts the infinite
background into a bounded representation.
Zip-NeRF~\citep{barron2023zipnerf} unifies the hash grid of Instant-NGP with
the anti-aliasing of Mip-NeRF~360, establishing the current quality frontier
for per-scene NeRF.
NeRF-W~\citep{martinbrualla2021nerfw} introduces per-image latent appearance
codes and a transient object model, enabling reconstruction of tourist-photographed
landmarks from uncontrolled internet images.
Ref-NeRF~\citep{verbin2022refnerf} restructures the directional encoding into
a physically-based reflection model that explicitly predicts diffuse colour,
specular tint, and surface roughness, dramatically improving the rendering of
specular and metallic materials.

\paragraph{Surface reconstruction.}
While volume rendering naturally produces accurate alpha composites, it does not
enforce a well-defined surface, leading to noisy geometry.
NeuS~\citep{wang2021neus} reformulates volume rendering using a logistic
density derived from a Signed Distance Function~(SDF), yielding watertight,
high-quality surface reconstruction.
VolSDF~\citep{yariv2021volsdf} derives a rigorous analytical mapping from
SDF values to volume density, providing a theoretically grounded alternative.
BakedSDF~\citep{yariv2023bakedsdf} combines the surface quality of SDF-based
NeRF with real-time rendering by extracting and baking the learned SDF into a
high-quality mesh.

Table~\ref{tab:nerf_methods} provides a comprehensive comparison of NeRF variants
across all three sub-families: per-scene optimisation, generalizable,
and dynamic methods.

\begin{table*}[t]
  \centering
  \caption{Comprehensive NeRF method comparison.
    \textbf{Category}: PS=per-scene, GEN=generalizable, DYN=dynamic.
    \textbf{Train}/\textbf{Render}: speed relative to vanilla NeRF (Fast/Med/Slow).
    \textbf{Gen.}: generalizes across scenes without per-scene re-training.
    \textbf{Dyn.}: handles dynamic or deformable scenes.
    \textbf{Surf.}: produces explicit surface (mesh/SDF).}
  \label{tab:nerf_methods}
  \scriptsize
  \setlength{\tabcolsep}{3pt}
  \begin{tabular}{@{}llccccccp{4.2cm}@{}}
    \toprule
    \textbf{Method} & \textbf{Venue} & \textbf{Cat.}
      & \textbf{Train} & \textbf{Render}
      & \textbf{Gen.} & \textbf{Dyn.} & \textbf{Surf.}
      & \textbf{Key Contribution} \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Per-scene optimisation}} \\
    NeRF~\citep{mildenhall2020nerf}            & ECCV'20    & PS & Slow & Slow & \no & \no & \no & MLP + positional encoding \\
    NeRF-W~\citep{martinbrualla2021nerfw}       & CVPR'21    & PS & Slow & Slow & \no & \no & \no & In-the-wild appearance codes \\
    Mip-NeRF~\citep{barron2021mipnerf}         & ICCV'21    & PS & Slow & Slow & \no & \no & \no & Integrated positional encoding \\
    NeuS~\citep{wang2021neus}                  & NeurIPS'21 & PS & Slow & Slow & \no & \no & \yes & SDF-based volume rendering \\
    VolSDF~\citep{yariv2021volsdf}             & NeurIPS'21 & PS & Slow & Slow & \no & \no & \yes & Analytical SDF-to-density \\
    Plenoxels~\citep{fridovich2022plenoxels}   & CVPR'22    & PS & Fast & Fast & \no & \no & \no & No-MLP voxel grid + SH \\
    TensoRF~\citep{chen2022tensorf}            & ECCV'22    & PS & Med  & Fast & \no & \no & \no & CP/VM tensor decomposition \\
    DirectVoxGO~\citep{sun2022dvgo}            & CVPR'22    & PS & Fast & Fast & \no & \no & \no & Two-stage coarse-to-fine voxel \\
    Instant-NGP~\citep{muller2022instant}      & TOG'22     & PS & Fast & Fast & \no & \no & \no & Multi-resolution hash grid \\
    Mip-NeRF360~\citep{barron2022mipnerf360}   & CVPR'22    & PS & Slow & Slow & \no & \no & \no & Unbounded scene contraction \\
    Ref-NeRF~\citep{verbin2022refnerf}         & CVPR'22    & PS & Slow & Slow & \no & \no & \no & Physics-based reflection model \\
    Zip-NeRF~\citep{barron2023zipnerf}         & ICCV'23    & PS & Fast & Fast & \no & \no & \no & Hash grid + anti-aliasing \\
    BakedSDF~\citep{yariv2023bakedsdf}         & SIGGRAPH'23& PS & Med  & Fast & \no & \no & \yes & SDF mesh extraction + baking \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Generalizable methods}} \\
    pixelNeRF~\citep{yu2021pixelnerf}          & CVPR'21    & GEN& Fast & Slow & \yes& \no & \no & Pixel-aligned CNN features \\
    IBRNet~\citep{wang2021ibrnet}              & CVPR'21    & GEN& Fast & Slow & \yes& \no & \no & Cross-view transformer aggregation \\
    MVSNeRF~\citep{chen2021mvsnerf}            & ICCV'21    & GEN& Fast & Slow & \yes& \no & \no & Cost-volume geometry prior \\
    SRT~\citep{sajjadi2022srt}                 & CVPR'22    & GEN& Fast & Fast & \yes& \no & \no & Set-latent scene representation \\
    GNT~\citep{wang2022gnt}                    & arXiv'22   & GEN& Fast & Med  & \yes& \no & \no & Fully attention-based, no NeRF MLP \\
    NerfDiff~\citep{gu2023nerfdiff}            & ICML'23    & GEN& Fast & Slow & \yes& \no & \no & Diffusion prior for single-view \\
    ZeroRF~\citep{shi2023zerorf}               & arXiv'23   & GEN& Med  & Slow & \pt & \no & \no & Zero-pretrain sparse-view 360$^\circ$ \\
    FrameNeRF~\citep{yan2024framenerf}         & arXiv'24   & GEN& Fast & Slow & \yes& \no & \no & Video frames as pseudo views \\
    \midrule
    \multicolumn{9}{@{}l}{\textit{Dynamic \& deformable methods}} \\
    D-NeRF~\citep{pumarola2021dnerf}           & CVPR'21    & DYN& Slow & Slow & \no & \yes& \no & Canonical deformation network \\
    NSFF~\citep{li2021neural_scene_flow}       & CVPR'21    & DYN& Slow & Slow & \no & \yes& \no & Explicit scene flow field \\
    Nerfies~\citep{park2021nerfies}            & ICCV'21    & DYN& Slow & Slow & \no & \yes& \no & Per-frame latent deformation \\
    HyperNeRF~\citep{park2021hypernerf}        & TOG'21     & DYN& Slow & Slow & \no & \yes& \no & Higher-dimensional ambient space \\
    Non-Rigid NeRF~\citep{tretschk2021nonrigid}& ICCV'21    & DYN& Slow & Slow & \no & \yes& \no & Monocular ray bending \\
    DyNeRF~\citep{li2022neural3d}              & CVPR'22    & DYN& Slow & Slow & \no & \yes& \no & Time-conditioned multi-view NeRF \\
    TiNeuVox~\citep{fang2022tineuvox}          & SIGGRAPH'22& DYN& Med  & Med  & \no & \yes& \no & Voxel-accelerated dynamic NeRF \\
    K-Planes~\citep{fridovich2023kplanes}      & CVPR'23    & DYN& Fast & Fast & \no & \yes& \no & 4-D plane decomposition \\
    HexPlane~\citep{cao2023hexplane}           & CVPR'23    & DYN& Fast & Fast & \no & \yes& \no & Concurrent 4-D plane factorisation \\
    NeRFPlayer~\citep{song2023nerfplayer}      & TOG'23     & DYN& Med  & Med  & \no & \yes& \no & Static/deforming/new region decomp. \\
    Soup-of-Planes~\citep{du2023soupofplanes}  & ICCV'23    & DYN& Fast & Fast & \no & \yes& \no & Planar fitting for casual video \\
    Pseudo-GDV~\citep{zhao2024pgdvs}           & CVPR'24    & DYN& Med  & Slow & \no & \yes& \no & Static + dynamic inpainting \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Generalizable NeRF}

Per-scene NeRF requires lengthy per-scene optimisation, limiting scalability.
\emph{Generalisable} methods learn a shared prior across scenes, enabling
fast inference on new scenes.

pixelNeRF~\citep{yu2021pixelnerf} conditions the NeRF MLP on pixel-aligned
CNN features extracted from reference views, enabling single-image reconstruction.
IBRNet~\citep{wang2021ibrnet} uses a transformer to aggregate features from
multiple reference views at query points along the ray.
MVSNeRF~\citep{chen2021mvsnerf} builds a cost volume from stereo for
geometry-aware feature extraction.
GNT~\citep{wang2022gnt} dispenses with the NeRF MLP entirely, replacing it
with a two-stage transformer that first encodes reference views into a
view-independent representation and then renders the target view via
cross-attention, learning geometric reasoning as an emergent property of
attention rather than an explicit inductive bias.
SRT~\citep{sajjadi2022srt} pursues a similar philosophy but uses a
set-latent architecture, encoding an arbitrary number of views into a
fixed-size scene representation without pose assumptions.
NerfDiff~\citep{gu2023nerfdiff} augments generalizable NeRF with diffusion
priors for single-image synthesis.
ZeroRF~\citep{shi2023zerorf} achieves sparse-view 360$^\circ$ reconstruction
via per-scene optimisation with zero pretraining, leveraging factorised feature grids.

\paragraph{Few-shot and test-time methods.}
FrameNeRF~\citep{yan2024framenerf} presents a simple and efficient framework
for few-shot NVS by exploiting temporal video frames as dense pseudo views.
Recollection from Pensieve~\citep{zhou2024pensieve} learns NVS from uncalibrated
internet videos without posed supervision.
Test-Time Completion~\citep{bao2024tttcompletion} repurposes natural video
completion as a zero-shot NVS prior at test time.
Mixed Neural Voxels~\citep{wang2023mixedvoxels} enables fast multi-view video
synthesis by decomposing scenes into static and dynamic voxel grids.

\subsection{Dynamic NeRF}

Modelling non-rigid dynamic scenes is a central challenge for world modeling.

D-NeRF~\citep{pumarola2021dnerf} introduces a deformation network mapping each
spacetime point to a canonical static space.
Nerfies~\citep{park2021nerfies} and
HyperNeRF~\citep{park2021hypernerf} model complex non-rigid deformations via
latent codes and higher-dimensional ambient spaces.
Neural Scene Flow Fields~(NSFF)~\citep{li2021neural_scene_flow} models dynamic
scenes with scene flow representation.
Non-Rigid NeRF~\citep{tretschk2021nonrigid} reconstructs and synthesises
novel views of dynamic scenes from monocular video with explicit ray bending.
DyNeRF~\citep{li2022neural3d} extends NeRF to monocular video via a
time-conditioned model.
TiNeuVox~\citep{fang2022tineuvox} uses voxel features to accelerate dynamic NeRF.
K-Planes~\citep{fridovich2023kplanes} decomposes 4-D space-time into six
feature planes---three spatial (XY, XZ, YZ) and three spatial-temporal
(XT, YT, ZT)---each encoded as a learned 2-D grid, making the architecture
highly memory-efficient and interpretable.
HexPlane~\citep{cao2023hexplane} independently proposes the same factorisation,
demonstrating its generality; together these works establish plane-based
4-D decomposition as the dominant acceleration paradigm for dynamic NeRF.
NeRFPlayer~\citep{song2023nerfplayer} separately encodes static, deforming and
newly appearing regions.
Soup-of-Planes~\citep{du2023soupofplanes} enables fast view synthesis of
casual monocular videos by fitting a set of planar scenes.
Pseudo-Generalised Dynamic View~\citep{zhao2024pgdvs} synthesises dynamic views
from monocular video by combining static and dynamic video inpainting.
Decoupled Dynamic Video~\citep{wang2024decoupleddynamic} disentangles foreground
dynamics and background statics for monocular dynamic view synthesis.

The dynamic NeRF methods are included in the comprehensive
Table~\ref{tab:nerf_methods} above.

% ─────────────────────────────────────────────────────────────────────────────
