\section{3D Gaussian Splatting}
\label{sec:3dgs}

\subsection{Foundational 3DGS}

3DGS~\citep{kerbl2023gaussian} achieved real-time ($>$100\,fps at 1080p)
NVS with training times of 30--45 minutes on commodity hardware,
surpassing NeRF-quality baselines on standard benchmarks.
Its three key innovations are:
(i)~anisotropic Gaussian primitives that align naturally with planar scene
structures;
(ii)~tile-based rasterisation that enables massively parallel GPU rendering;
(iii)~adaptive densification that automatically allocates capacity to
under-reconstructed regions.
The resulting representation is fully explicit and editable, opening new
directions in scene editing, simulation, and avatar rendering.

\paragraph{Quality and anti-aliasing.}
Mip-Splatting~\citep{yu2024mipsplatting} identifies aliasing in vanilla 3DGS
as arising from the approximation of 3-D Gaussian convolution during 2-D
projection.
It introduces a 3-D smoothing filter derived from the sampling footprint,
enabling accurate multi-scale rendering and substantially reducing floaters
at large distances.
AbsGS~\citep{ye2024absgs} revisits the densification criterion: replacing
the mean gradient with an absolute-value measure improves coverage in
thin structures and reduces needle-like artifacts near object boundaries.

\paragraph{Compression and memory efficiency.}
A single 3DGS scene may require 300MB--1GB of storage, limiting deployment.
Scaffold-GS~\citep{lu2023scaffoldgs} introduces a sparse set of anchor
primitives from which neural networks predict neighbouring Gaussian attributes;
this structured representation reduces storage by $3$--$5\times$ with
comparable quality.
LightGaussian~\citep{fan2023lightgaussian} achieves $15\times$ compression via
global Gaussian pruning based on visibility contribution and subsequent
vector quantisation of attributes.
EAGLES~\citep{girish2024eagles} applies residual vector quantisation
for fine-grained attribute compression.
HAC~\citep{chen2024hac} proposes a hash-grid-based context model that predicts
per-Gaussian attribute entropy, enabling learned arithmetic coding for
extreme compression ratios.

\paragraph{Surface reconstruction.}
SuGaR~\citep{guedon2024sugar} aligns Gaussians to the scene surface by adding
a regularisation term that encourages Gaussians to become flat discs at
the surface, followed by mesh extraction and joint Gaussian--mesh refinement.
2DGS~\citep{huang2024_2dgs} replaces 3-D Gaussians with 2-D disc primitives,
which by construction lie on surfaces and produce accurate normal maps.
Gaussian Opacity Fields~(GOF)~\citep{yu2024gof} fuses Gaussian Splatting with
a continuous opacity field for topologically correct surface extraction in
unbounded scenes.

\paragraph{Scene editing and compositionality.}
The explicit Gaussian representation is particularly amenable to editing:
individual primitives can be identified, moved, duplicated, or removed.
GaussianEditor~\citep{chen2024gaussianeditor} harnesses a large vision-language
model to select semantically relevant Gaussians from a text description and
optimises their appearance with a diffusion prior.
Gaussian Grouping~\citep{ye2023gaussian_grouping} augments each Gaussian with
a learned identity field, trained with SAM~\citep{kirillov2023sam} masks from
rendered views, enabling instance-level part decomposition and editing.
GALA3D~\citep{zhou2024gala3d} uses an LLM to generate a 3-D layout from text
and initialises Gaussians per-object, enabling compositional scene generation.

Table~\ref{tab:3dgs_methods} provides a comprehensive view of the 3DGS family.

\begin{table*}[t]
  \centering
  \caption{Comprehensive 3D Gaussian Splatting method comparison.
    \textbf{Category}: STATIC=per-scene static, DYN=dynamic/deformable, FF=feed-forward generalizable.
    \textbf{RT}: real-time rendering ($>$30\,fps).
    \textbf{Comp.}: dedicated compression or storage reduction.
    \textbf{Surf.}: explicit surface extraction.
    \textbf{Unp.}: works on unposed images.
    \textbf{Input}: minimum input views for FF methods.}
  \label{tab:3dgs_methods}
  \scriptsize
  \setlength{\tabcolsep}{2.8pt}
  \begin{tabular}{@{}llcccccccp{3.8cm}@{}}
    \toprule
    \textbf{Method} & \textbf{Venue} & \textbf{Cat.}
      & \textbf{RT} & \textbf{Comp.} & \textbf{Surf.}
      & \textbf{Edit.} & \textbf{Unp.} & \textbf{Input}
      & \textbf{Key Contribution} \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Static per-scene}} \\
    3DGS~\citep{kerbl2023gaussian}               & TOG'23      & S & \yes& \no & \no & \no & \no & MV  & Tile-based Gaussian splatting \\
    Mip-Splatting~\citep{yu2024mipsplatting}     & CVPR'24     & S & \yes& \no & \no & \no & \no & MV  & 3-D smoothing filter, anti-alias \\
    AbsGS~\citep{ye2024absgs}                    & NeurIPS'24  & S & \yes& \no & \no & \no & \no & MV  & Absolute-gradient densification \\
    Scaffold-GS~\citep{lu2023scaffoldgs}         & CVPR'24     & S & \yes& \yes& \no & \no & \no & MV  & Anchor-driven neural attributes \\
    LightGaussian~\citep{fan2023lightgaussian}   & NeurIPS'23  & S & \yes& \yes& \no & \no & \no & MV  & Visibility pruning + VQ, 15$\times$ \\
    Compact3D~\citep{lee2024compact3d}           & arXiv'24    & S & \yes& \yes& \no & \no & \no & MV  & k-means codebook compression \\
    EAGLES~\citep{girish2024eagles}              & ECCV'24     & S & \yes& \yes& \no & \no & \no & MV  & Residual vector quantisation \\
    HAC~\citep{chen2024hac}                      & ECCV'24     & S & \yes& \yes& \no & \no & \no & MV  & Hash context entropy coding \\
    SuGaR~\citep{guedon2024sugar}                & CVPR'24     & S & \yes& \no & \yes& \pt & \no & MV  & Flat-disc + mesh joint opt. \\
    2DGS~\citep{huang2024_2dgs}                  & SIGGRAPH'24 & S & \yes& \no & \yes& \no & \no & MV  & 2-D disc primitives \\
    GOF~\citep{yu2024gof}                        & arXiv'24    & S & \yes& \no & \yes& \no & \no & MV  & Opacity field for surface \\
    GaussianEditor~\citep{chen2024gaussianeditor} & CVPR'24    & S & \yes& \no & \no & \yes& \no & MV  & LLM-guided semantic editing \\
    Gaussian Grouping~\citep{ye2023gaussian_grouping}& ECCV'24  & S & \yes& \no & \no & \yes& \no & MV  & SAM-based identity field \\
    GALA3D~\citep{zhou2024gala3d}                & ICML'24     & S & \yes& \no & \no & \yes& \no & MV  & LLM-layout compositional gen. \\
    LERF~\citep{kerr2023lerf}                    & ICCV'23     & S & \no & \no & \no & \pt & \no & MV  & CLIP language grounding in NeRF \\
    LangSplat~\citep{qin2024langsplat}           & CVPR'24     & S & \yes& \no & \no & \pt & \no & MV  & Language field in 3DGS \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Dynamic \& deformable}} \\
    4DGS~\citep{wu2024_4dgs}                     & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & 4-D Gaussian + temporal field \\
    Deformable3DGS~\citep{yang2024deformable}    & ICLR'24     & D & \yes& \no & \no & \no & \no & Mono& Canonical deformation network \\
    SpacetimeGaussians~\citep{yang2024spacetime} & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & Polynomial time trajectories \\
    SC-GS~\citep{huang2024scgs}                  & CVPR'24     & D & \yes& \no & \no & \no & \no & Mono& Sparse control point deform. \\
    Gaussian-Flow~\citep{lin2024gaussianflow}    & CVPR'24     & D & \yes& \no & \no & \no & \no & MV  & Fourier polynomial motion \\
    Shape-of-Motion~\citep{wang2024shapeofmotion}& arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Per-Gaussian 3-D motion \\
    Dynamic 3DGS~\citep{luiten2024dynamic3dgs}   & 3DV'24      & D & \yes& \no & \no & \no & \no & MV  & Persistent Gaussian tracking \\
    GauFRe~\citep{liang2024gaufre}               & arXiv'24    & D & \yes& \no & \no & \no & \no & Mono& Flow-regularised deformation \\
    MoVieS~\citep{sun2024movies}                 & arXiv'24    & D & \yes& \no & \no & \no & \no & Mono& 4-D synthesis in $\sim$1 second \\
    SC-4D~\citep{kim2024selfcal4d}               & arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Self-calibrating 4-D from video \\
    Reconstruct-Inpaint~\citep{han2024rif}       & arXiv'24    & D & \no & \no & \no & \no & \no & Mono& Recon + inpaint + TTF pipeline \\
    Articulated-GS~\citep{li2024articulatedgs}   & NeurIPS'24  & D & \yes& \no & \no & \no & \no & MV  & Template-free articulated 3DGS \\
    EVA-Gaussian~\citep{hu2024evacg}             & arXiv'24    & D & \yes& \no & \no & \no & \no & MV  & Real-time human NVS \\
    \midrule
    \multicolumn{10}{@{}l}{\textit{Feed-forward generalizable}} \\
    pixelSplat~\citep{charatan2024pixelsplat}    & CVPR'24     & FF& \yes& \no & \no & \no & \no & 2   & Stereo epipolar features \\
    MVSplat~\citep{chen2024mvsplat}              & ECCV'24     & FF& \yes& \no & \no & \no & \no & 2--6& Cost-volume geometry prior \\
    TranSplat~\citep{zhang2024transplate}        & arXiv'24    & FF& \yes& \no & \no & \no & \no & 2+  & Transformer MV prediction \\
    FSGS~\citep{zhu2024fsgs}                     & CVPR'24     & FF& \yes& \no & \no & \no & \no & 3   & Gaussian unpooling, few-shot \\
    DepthSplat~\citep{xu2024depthsplat}          & arXiv'24    & FF& \yes& \no & \no & \no & \no & 2+  & Joint depth + Gaussian pred. \\
    Splatt3R~\citep{smart2024splatt3r}           & arXiv'24    & FF& \yes& \no & \no & \no & \yes& 2   & DUSt3R backbone, unposed \\
    TriplaneGaussian~\citep{zou2024triplanegaussian}& CVPR'24   & FF& \yes& \no & \no & \no & \no & 1   & Triplane + Gaussian, single-view \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Dynamic 3DGS}

4DGS~\citep{wu2024_4dgs} extends Gaussians into 4-D space-time with a
temporal radiance field.
SpacetimeGaussians~\citep{yang2024spacetime} assigns polynomial time
trajectories to each Gaussian.
Deformable3DGS~\citep{yang2024deformable} applies a deformation network to
canonical Gaussians.
SC-GS~\citep{huang2024scgs} uses sparse control points for monocular dynamic
scenes.
Gaussian-Flow~\citep{lin2024gaussianflow} models Gaussian dynamics via
polynomial motion fields in Fourier space.
Shape-of-Motion~\citep{wang2024shapeofmotion} reconstructs per-Gaussian 3-D
motion from monocular videos.
Dynamic 3D Gaussians~\citep{luiten2024dynamic3dgs} tracks scene dynamics
through persistent Gaussian views, jointly optimising NVS and tracking.
GauFRe~\citep{liang2024gaufre} uses Gaussian deformation fields with
flow regularisation for real-time dynamic NVS.
MoVieS~\citep{sun2024movies} generates motion-aware 4-D dynamic views in
approximately one second via efficient Gaussian motion estimation.
Self-Calibrating 4D~\citep{kim2024selfcal4d} performs 4-D NVS from monocular
videos by jointly estimating camera parameters and Gaussian deformations.
Reconstruct-Inpaint-Finetune~\citep{han2024rif} addresses dynamic monocular
NVS via a three-stage pipeline of reconstruction, inpainting, and test-time
refinement.
EVA-Gaussian~\citep{hu2024evacg} achieves real-time human NVS under diverse
multi-view settings.
Articulated Gaussian~\citep{li2024articulatedgs} enables template-free
articulated Gaussian Splatting for real-time reposable dynamic view synthesis.

\subsection{Feed-Forward Generalizable 3DGS}

pixelSplat~\citep{charatan2024pixelsplat} predicts Gaussian primitives from
stereo image pairs in a single forward pass.
MVSplat~\citep{chen2024mvsplat} adds cost-volume priors for better geometry.
TranSplat~\citep{zhang2024transplate} introduces a transformer architecture for
multi-view Gaussian prediction.
Splatt3R~\citep{smart2024splatt3r} builds on DUSt3R~\citep{wang2024dust3r}
for unposed image pairs.
FSGS~\citep{zhu2024fsgs} enables few-shot Gaussian Splatting via Gaussian
unpooling and monocular depth supervision.
DepthSplat~\citep{xu2024depthsplat} jointly predicts depth and Gaussians
for generalizable NVS.
TriplaneGaussian~\citep{zou2024triplanegaussian} combines triplane and
Gaussian representations in a transformer-based single-view reconstruction model.

The feed-forward generalizable methods are included in the comprehensive
Table~\ref{tab:3dgs_methods} above (FF category).

% ─────────────────────────────────────────────────────────────────────────────
