\section{Open Problems and Research Roadmap}
\label{sec:open}

\subsection{The Geometry--Generation Trade-Off}

Reconstruction methods deliver geometric precision but lack generative priors;
diffusion methods provide rich priors but can violate multi-view consistency.
Methods like ReconFusion, Cat3D, and ViewCrafter begin to bridge this gap,
but the fundamental tension remains.
Principled solutions (\eg, geometry-grounded diffusion, diffusion-seeded 3DGS)
are an active frontier.

\subsection{Long-Horizon Temporal Coherence}

Existing 4-D methods maintain coherence over a few seconds.
Two promising directions:
(1) \emph{Persistent scene memory}: Gaussian maps that accumulate observations
across time and preserve global consistency;
(2) \emph{World model pre-training}: leveraging internet-scale video prediction
as a temporal prior that can be fine-tuned to specific domains.

\subsection{Physical Grounding}

Current methods learn appearance correlations, not physical laws.
Encouraging directions include:
PAC-NeRF~\citep{li2023pac} for continuum mechanics integration;
physics-informed deformable Gaussians for material simulation.
The central challenge is integrating differentiable physics into the rendering pipeline
in a way that scales to complex, open-world scenes.

\subsection{Open-Vocabulary Semantic Integration}

Methods like LERF~\citep{kerr2023lerf} ground CLIP features in 3-D via NeRF,
enabling open-vocabulary 3-D queries.
LangSplat~\citep{qin2024langsplat} brings language grounding to 3DGS.
Feature3DGS~\citep{zhou2024feature3dgs} distils various 2-D foundation model features.
Extending these to dynamic, actionable scenes is a key step toward language-grounded
world models.

\subsection{Evaluation and Benchmarks}

We propose a five-axis world-modelling benchmark:
(1)~3-D consistency under adversarial viewpoints;
(2)~temporal coherence over long horizons ($>$30 s);
(3)~action responsiveness in interactive settings;
(4)~physical plausibility via physics-engine verification;
(5)~open-world generalisation from a single image.

\subsection{Research Roadmap}

\begin{enumerate}[leftmargin=*,label=\textbf{R\arabic*.}]
  \item Unified architectures combining explicit 3-D with diffusion priors.
  \item Internet-scale feed-forward generalisation for open-world NVS.
  \item Differentiable physics priors in 4-D world models.
  \item Language and semantic grounding for action-conditioned generation.
  \item Comprehensive world-modelling evaluation suite.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
